---
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Using the **loo** package

*The following is excerpted (with minor edits) from our paper Vehtari, A.,
Gelman, A., and Gabry, J. (2015). Efficient implementation of leave-one-out 
cross-validation and WAIC for evaluating fitted Bayesian models.*

<br>

This example comes from a survey of residents from a small area in Bangladesh
that was affected by arsenic in drinking water. Respondents with elevated
arsenic levels in their wells were asked if they were interested in getting
water from a neighbor's well, and a series of logistic regressions were fit to
predict this binary response given various information about the households
(Gelman and Hill, 2007). Here we fit a model for the well-switching response
given two predictors: the arsenic level of the water in the resident's home, and
the distance of the house from the nearest safe well.

The sample size in this example is $N=3020$, which is not huge but is large
enough that it is important to have a computational method for LOO that is fast
for each data point. On the plus side, with such a large dataset, the influence
of any given observation is small, and so the computations should be stable.

### Stan and R code 

Here is the Stan code for fitting the logistic regression model:

```
'logistic.stan'

data { 
  int N; // number of data points
  int K; // number of predictors (including intercept)
  int<lower=0,upper=1> y[N]; // outcome
  matrix[N,K] X; // predictors (including intercept)
}
parameters {
  vector[K] beta;
}
model {
  y ~ bernoulli_logit(X * beta);
}
generated quantities {
  vector[N] log_lik;
  for (n in 1:N)
    log_lik[n] <- bernoulli_logit_log(y[n], X[n] * beta);
}
```

We have defined the log likelihood as a vector named `log_lik` in the generated 
quantities block so that the individual terms will be saved by Stan. After 
running Stan, `log_lik` can be extracted (using the `extract_log_lik` function
provided in the **loo** package) as an $S \times N$ matrix, where $S$ is the 
number of simulations (posterior draws) and $N$ is the number of data points.

<br>

Next we fit the Stan model using the **rstan** package

```{r, eval=FALSE}
library("rstan")

# Prepare data 
url <- "http://stat.columbia.edu/~gelman/arm/examples/arsenic/wells.dat"
wells <- read.table(url)
X <- model.matrix(~I(dist/100) + arsenic, wells) 
data <- list(N = nrow(wells), K = ncol(X), X = X, y = wells$switch)

# Fit model
fit <- stan("logistic.stan", data = data) # defaults to iter = 2000, chains = 4
```

and then use the **loo** package to compute LOO and WAIC:

```{r, eval=FALSE}
library("loo")

# Extract log-likelihood and compute LOO, WAIC
log_lik <- extract_log_lik(fit)
loo <- loo_and_waic(log_lik)
print(loo)
# loo and waic are the same out to several decimal places for this model
print(loo, digits = 4) 
```

<br>

To compare this model to alternative model for the same data we can 
use the `loo_and_waic_diff` function. 

```{r, eval=FALSE}
# First run a second model using log(arsenic) instead of arsenic
data$X[,"arsenic"] <- log(data$X[,"arsenic"])
fit2 <- stan(fit = fit, data = data)
log_lik2 <- extract_log_lik(fit2)
loo2 <- loo_and_waic(log_lik2)

# Compare using loo_and_waic_diff
diff <- loo_and_waic_diff(loo, loo2)
```

### References

Gelman, A., and Hill, J. (2007).  *Data Analysis Using Regression and Multilevel Hierarchical Models.*  Cambridge University Press.

Stan Development Team (2015a). Stan: A C++ library for probability and sampling, version 2.6. http://mc-stan.org

Stan Development Team (2015b). RStan, version 2.6. http://mc-stan.org/rstan.html.
