---
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Using the **loo** package

*The following is excerpted (with minor edits) from our paper*   
Vehtari, A., Gelman, A., and Gabry, J. (2016).   
[*Practical Bayesian model evaluation using leave-one-out cross-validation and WAIC*](http://arxiv.org/abs/1507.04544).

<br>

This example comes from a survey of residents from a small area in Bangladesh
that was affected by arsenic in drinking water. Respondents with elevated
arsenic levels in their wells were asked if they were interested in getting
water from a neighbor's well, and a series of logistic regressions were fit to
predict this binary response given various information about the households
(Gelman and Hill, 2007). Here we fit a model for the well-switching response
given two predictors: the arsenic level of the water in the resident's home, and
the distance of the house from the nearest safe well.

The sample size in this example is $N=3020$, which is not huge but is large
enough that it is important to have a computational method for LOO that is fast
for each data point. On the plus side, with such a large dataset, the influence
of any given observation is small, and so the computations should be stable.

### Stan and R code 

Here is the Stan code for fitting the logistic regression model:

```{r, eval=FALSE}
'logistic.stan'

data {
  int<lower=0> N; # number of data points
  int<lower=0> P; # number of predictors (including intercept)
  int<lower=0,upper=1> y[N]; # binary outcome
  matrix[N,P] x; # predictors (including intercept)
  real a;
}
parameters {
  real beta0;
  vector[P] beta;
}
model {
  beta0 ~ student_t(7, a, 0.1);
  beta ~ student_t(7, 0, 1);
  y ~ bernoulli_logit(beta0 + x * beta);
}
generated quantities {
  vector[N] log_lik;
  for (n in 1:N)
    log_lik[n] <- bernoulli_logit_log(y[n], beta0 + x[n] * beta);
}
```

We have defined the log likelihood as a vector named `log_lik` in the generated 
quantities block so that the individual terms will be saved by Stan. After 
running Stan, `log_lik` can be extracted (using the `extract_log_lik` function
provided in the **loo** package) as an $S \times N$ matrix, where $S$ is the 
number of simulations (posterior draws) and $N$ is the number of data points.

<br>

Next we fit the Stan model using the **rstan** package

```{r, eval=FALSE}
library("rstan")

# Prepare data 
url <- "http://stat.columbia.edu/~gelman/arm/examples/arsenic/wells.dat"
wells <- read.table(url)
wells$dist100 <- wells$dist / 100 # rescale 
y <- wells$switch
a <- qlogis(mean(y)) # i.e., a = logit(Pr(y = 1))
x <- scale(model.matrix(~ 0 + dist + arsenic, wells))
data <- list(N = nrow(x), P = ncol(x), a = a, x = x, y = y)

# Fit model
fit1 <- stan("logistic.stan", data = data)
```

and then use the **loo** package to compute LOO.

```{r, eval=FALSE}
library("loo")

# Extract log-likelihood and compute LOO
log_lik1 <- extract_log_lik(fit1)
loo1 <- loo(log_lik1) # or waic(log_lik1) to compute WAIC
```

<br>

To compare this model to alternative model for the same data we can 
use the `compare` function. 

```{r, eval=FALSE}
# First run a second model using log(arsenic) instead of arsenic
data$x <- scale(model.matrix(~ 0 + dist100 + log(arsenic), wells))
fit2 <- stan(fit = fit1, data = data)
log_lik2 <- extract_log_lik(fit2)
loo2 <- loo(log_lik2)

# Compare
diff <- compare(loo1, loo2) 
```

### References

Gelman, A., and Hill, J. (2007).  *Data Analysis Using Regression and Multilevel Hierarchical Models.*  Cambridge University Press.

Stan Development Team (2015a). Stan: A C++ library for probability and sampling, version 2.6. http://mc-stan.org

Stan Development Team (2015b). RStan, version 2.6. http://mc-stan.org/interfaces/rstan.html
