---
title: "Parallel and distributed loo"
author: "The loo developers"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Parallel and distributed loo}
-->
```{r, child="children/SETTINGS-knitr.txt"}
```

```{r, child="children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

# Introduction

Computing PSIS-LOO, WAIC, and related quantities is *embarrassingly parallel*:
the expensive work is done once per observation (or per fold), and these
computations are independent of one another. The __loo__ package can therefore
spread this work across multiple cores or even multiple machines.

As of __loo__ 2.10 the parallel backend is built on the
[__mirai__](https://mirai.r-lib.org/) package, with
[__mori__](https://mirai.r-lib.org/) providing zero-copy sharing of large
objects (such as the posterior draws) between local workers. You do not need to
learn either package to benefit from parallelism, but understanding the three
ways to control the backend lets you pick the right one for your workflow:

1. **Per-call parallelism** with the `cores` argument --- the simplest option,
   good for one-off calls.
2. **A persistent session pool** via `options(loo.daemons = k)` or the
   `LOO_DAEMONS` environment variable --- best when you call __loo__ functions
   *many* times in one session (simulations, benchmarks, batch/HPC scripts).
3. **A pool you manage yourself** with `mirai::daemons()` --- needed for
   remote/SSH/HPC workers, and always takes precedence over the options above.

Every __loo__ function that does per-observation work accepts a `cores`
argument and respects these settings, including `loo()`, `psis()`, `waic()`,
`relative_eff()`, `loo_subsample()`, `loo_moment_match()`, and
`loo_model_weights()`.

One rule ties all three together: **when a daemon pool is already connected
(options 2 or 3), __loo__ always uses it and the `cores` argument is ignored.**
`cores` only takes effect when no pool is connected (option 1). The section
[How `cores` interacts with a pool](#how-cores-interacts-with-a-pool) spells
this out.

# Setup

The examples below only need the __loo__ package and its built-in example
log-likelihood objects, so they are fast and self-contained.

```{r setup, message=FALSE}
library("loo")

# A 500 (draws) x 2 (chains) x 32 (observations) log-likelihood array.
LLarr <- example_loglik_array()
r_eff <- relative_eff(exp(LLarr))
```

# 1. Per-call parallelism with `cores`

The `cores` argument is the simplest way to parallelize. Passing `cores > 1`
tells __loo__ to create a local pool of worker processes *for the duration of
that single call* and to shut it down automatically when the call returns:

```{r cores-basic}
# Serial (the default)
loo_serial <- loo(LLarr, r_eff = r_eff, cores = 1)

# Parallel across 2 local workers, then automatically cleaned up
loo_parallel <- loo(LLarr, r_eff = r_eff, cores = 2)
```

The results are identical --- parallelism only changes *how* the work is
scheduled, never the answer:

```{r cores-equal}
all.equal(loo_serial$estimates, loo_parallel$estimates)
```

You can set the default number of cores for a whole session with the standard
`mc.cores` option, so you don't have to pass `cores` to every call:

```{r mc-cores, eval=FALSE}
options(mc.cores = 4)
loo(LLarr, r_eff = r_eff) # uses 4 cores
```

Because the pool is created and destroyed on every call, this mode pays a small
start-up/teardown cost each time. For a single analysis that cost is
negligible, but if you call __loo__ functions repeatedly it can add up --- which
is exactly what the next section addresses.

# 2. A persistent session pool: `loo.daemons` / `LOO_DAEMONS`

When you call __loo__ functions many times in one session --- for example in a
simulation study, a benchmark, or a script that loops over many models --- you
can avoid repeatedly spinning workers up and down by keeping a pool **warm for
the whole session**.

Opt in by setting either the `loo.daemons` R option or the `LOO_DAEMONS`
environment variable to the desired number of workers (an integer `>= 2`):

```{r persistent-option, eval=FALSE}
# Either set the R option ...
options(loo.daemons = 4)

# ... or, equivalently, set the environment variable (useful for batch/HPC
# scripts and `Rscript`):
Sys.setenv(LOO_DAEMONS = 4)
```

With this enabled, __loo__:

* creates a local pool **lazily**, on the first call made with `cores > 1`
  (so nothing is started if you only do serial work),
* **reuses** that same warm pool for every later call --- which, because a
  connected pool always wins, then run in parallel *regardless* of `cores`, and
* **cleans it up automatically** when the R session ends (so no worker
  processes are left behind, even in non-interactive scripts).

This is ideal for loops like the following, where the pool is created once on
the first iteration and reused thereafter:

```{r persistent-loop, eval=FALSE}
options(loo.daemons = 4)

# Imagine `model_loglik` returns a log-likelihood matrix for each model.
# `cores = 4` on the first iteration triggers creation of the warm pool; on
# later iterations the pool already exists, so `cores` is ignored and the work
# simply runs on it.
results <- lapply(seq_len(100), function(i) {
  ll <- model_loglik(i)
  loo(ll, r_eff = relative_eff(exp(ll)), cores = 4)
})
```

A couple of details worth knowing:

* `cores > 1` is what *triggers* the lazy creation of the pool on the first
  call; the size of the persistent pool comes from `loo.daemons`/`LOO_DAEMONS`,
  not from `cores`. Once the pool is warm, every later call uses it regardless
  of `cores` (see "How `cores` interacts with a pool" below).
* Invalid values disable the feature: anything below `2`, a non-integer, or
  unset all mean "no persistent pool" (and __loo__ falls back to the per-call
  behavior from Section 1).

From a script, the most convenient pattern is to set the environment variable
outside R so you never touch __loo__-specific code:

```bash
LOO_DAEMONS=8 Rscript my_simulation.R
```

# 3. User-managed pools, including remote/HPC workers

If you configure a pool yourself with `mirai::daemons()`, __loo__ will **reuse
it and never tear it down**. A user-managed pool always takes precedence over
both the `cores` per-call behavior and the `loo.daemons` option, which makes it
the right tool for advanced setups --- in particular **remote, SSH, or HPC
cluster** workers.

A local pool you manage by hand behaves just like the persistent pool from
Section 2. Because a connected pool always wins, you do not even need to pass
`cores` --- __loo__ uses the pool whatever `cores` is:

```{r user-local, eval=FALSE}
library(mirai)

daemons(4)                       # start 4 local workers
loo1 <- loo(LLarr, r_eff = r_eff) # reuses your pool (cores not needed)
loo2 <- loo(LLarr, r_eff = r_eff) # reuses it again
daemons(0)                       # you decide when to stop them
```

For distributed computing, point __mirai__ at remote machines or a cluster
scheduler. __loo__ does not need to know any of the details --- it simply uses
whatever pool is connected:

```{r user-remote, eval=FALSE}
library(mirai)

# Launch workers on remote hosts over SSH (see ?mirai::daemons and
# ?mirai::ssh_config for the full set of options, including HPC launchers).
daemons(
  n = 8,
  url = host_url(),
  remote = ssh_config(c("ssh://node1", "ssh://node2"))
)

loo_big <- loo(large_loglik, r_eff = r_eff_big, cores = 8)

daemons(0)
```

# How `cores` interacts with a pool

The rule is simple: **whenever a daemon pool is connected, __loo__ uses it and
ignores `cores`.** This applies both to a pool you set up yourself (Section 3)
and to a `loo.daemons` pool once it has been created (Section 2). In that state
the *value* of `cores` does not matter --- `cores = 2` and `cores = 6` behave
identically, and even `cores = 1` (the default when `mc.cores` is unset) still
runs in parallel on the pool. The first time a call would otherwise have looked
like a request for serial execution (`cores <= 1`) while a pool is connected,
__loo__ prints a one-time message noting that `cores` is being ignored.

`cores` only controls things when **no** pool is connected: there `cores > 1`
both enables parallelism and sets the size of the temporary per-call pool
(Section 1), and it is what triggers lazy creation of a `loo.daemons` pool
(Section 2). If you genuinely want a single call to run serially while a pool
is up, stop the pool first with `mirai::daemons(0)`.

## When is the number of worker daemons equal to `cores`?

It follows directly from the rule above. The number of workers actually doing
the computation equals `cores` **only in the per-call case**; in every other
case the worker count comes from the pool and `cores` is irrelevant:

| How parallelism is set up | Number of worker daemons | Equal to `cores`? |
|---|---|---|
| Per-call, no pool connected (`cores = k`, Section 1) | `k` | **Yes** --- loo spawns exactly `cores` daemons for the call |
| Persistent pool (`options(loo.daemons = k)`, Section 2) | `k` (from `loo.daemons`) | Only if you happen to pass the same `cores` |
| User-managed pool (`mirai::daemons(n)`, Section 3) | `n` (whatever you started) | Only if you happen to pass the same `cores` |

In other words, `cores` *is* the daemon count in the per-call path, but for a
persistent or user-managed pool the daemon count is fixed by `loo.daemons` or
by your `mirai::daemons()` call, and passing a different `cores` (including the
default `1`) does **not** change how many workers run. Internally __loo__ always
splits the work according to the number of daemons that are actually connected,
not the requested `cores`.

# Large objects and shared memory

When __loo__ uses the function method (e.g. `loo(llfun, data = ..., draws =
...)`), the posterior `draws` are reused identically by every observation. On a
**local** pool these large objects are written once into shared memory with
__mori__ so each worker maps the same physical pages (zero-copy), which keeps
memory use low even with many workers.

On a **remote** pool shared memory is not available, so __loo__ falls back to
ordinary serialization and chunks the work to send each large object roughly
once per worker rather than once per observation. You don't need to do anything
to get this behavior --- __loo__ detects whether the active pool is local and
chooses the appropriate transport automatically.

# Choosing a strategy

| Situation | Recommended approach |
|---|---|
| A single `loo()`/`psis()` call | `cores = k` (Section 1) |
| Default cores for a session | `options(mc.cores = k)` |
| Many calls in one session (simulations, benchmarks, batch jobs) | `options(loo.daemons = k)` or `LOO_DAEMONS=k` (Section 2) |
| Remote / SSH / HPC workers | `mirai::daemons(url = ..., remote = ...)` (Section 3) |

A few general notes:

* All approaches give identical results; they differ only in scheduling and
  set-up cost.
* These mechanisms compose safely. A pool you set up yourself is always reused
  and never removed by __loo__, and nested __loo__ calls reuse an outer pool
  rather than creating a new one. For example, `loo_model_weights()` sets up a
  single pool and reuses it across all of the models it evaluates.
* A connected pool always wins: when one is active, `cores` is ignored and the
  work runs on the pool (the first time a `cores <= 1` call is overridden this
  way, __loo__ prints a one-time message). Stop the pool with
  `mirai::daemons(0)` if you need a call to run serially.
* On Windows, avoid setting `mc.cores` in your `.Rprofile`
  (see [this issue](https://github.com/stan-dev/loo/issues/94)); pass `cores`
  explicitly or set the option in your script instead.

# See also

* `help("loo", package = "loo")` for the `cores` argument documentation.
* The [__mirai__](https://mirai.r-lib.org/) documentation, especially
  `?mirai::daemons`, for configuring local, remote, and HPC pools.
