---
title: "Parallel and distributed loo"
author: "The loo developers"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Parallel and distributed loo}
-->
```{r, child="children/SETTINGS-knitr.txt"}
```

```{r, child="children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

# Introduction

Computing PSIS-LOO, WAIC, and related quantities is *embarrassingly parallel*:
the expensive work is done once per observation (or per fold), and these
computations are independent of one another. The __loo__ package can therefore
spread this work across multiple cores or even multiple machines.

As of __loo__ 2.10 the parallel backend is built on the
[__mirai__](https://mirai.r-lib.org/) package, with
[__mori__](https://mirai.r-lib.org/) providing zero-copy sharing of large
objects (such as the posterior draws) between local workers. You do not need to
learn mirai to benefit from parallelism, but understanding the recommended
workflows (and the deprecated bridges) helps you pick the right approach:

1. **A pool you manage yourself** with `mirai::daemons()` --- recommended for
   repeated calls in one session and for remote/SSH/HPC workers. A connected
   pool always takes precedence over deprecated parallel controls.
2. **[loo_mirai()]** --- a convenience wrapper for map-style workflows over
   many models (optional `n_daemons` argument).

**Deprecated (still works, with a once-per-session warning):**

* The `cores` argument on __loo__ functions (`cores > 1` starts a per-call
  local pool when no mirai pool is connected).
* `options(mc.cores)` as a session default for loo parallelism.

Every __loo__ function that does per-observation work accepts a deprecated
`cores` argument, including `loo()`, `psis()`, `waic()`, `relative_eff()`,
`loo_subsample()`, `loo_moment_match()`, and `loo_model_weights()`.

See `vignette("migration-guide")` for migration examples. The section
[How deprecated `cores` / `mc.cores` interact with a pool](#how-deprecated-cores--mc.cores-interact-with-a-pool) spells
out how deprecated controls interact with user-managed pools.

# Setup

The examples below only need the __loo__ package and its built-in example
log-likelihood objects, so they are fast and self-contained.

```{r setup, message=FALSE}
library("loo")

# A 500 (draws) x 2 (chains) x 32 (observations) log-likelihood array.
LLarr <- example_loglik_array()
r_eff <- relative_eff(exp(LLarr))
```

# 1. Deprecated per-call parallelism (`cores` and `mc.cores`)

The `cores` argument and `options(mc.cores)` are **deprecated** and will be
removed in a future release. They **still work** as a bridge:
when the resolved core count is `> 1` and no mirai pool is connected, __loo__
starts a local per-call daemon pool. A once-per-session deprecation warning is
emitted if you pass `cores` explicitly (even `cores = 1`), if `options(mc.cores)`
is set, or if the resolved value is `> 1`. Prefer `mirai::daemons()` or
[loo_mirai()] for new code (Sections 2--3).

```{r cores-basic}
# Deprecated: explicit cores (warns once per session)
loo_serial <- loo(LLarr, r_eff = r_eff, cores = 1)
loo_parallel <- loo(LLarr, r_eff = r_eff, cores = 2)

# Deprecated: session default via mc.cores (also warns once per session)
# options(mc.cores = 2)
# loo_via_mc <- loo(LLarr, r_eff = r_eff)
```

The results are identical --- parallelism only changes *how* the work is
scheduled, never the answer:

```{r cores-equal}
all.equal(loo_serial$estimates, loo_parallel$estimates)
```

You can still set a session default with `options(mc.cores = k)` --- it remains currently
functional but is deprecated for loo parallelism (use
`mirai::daemons(k)` instead; see Section 2).

Because the pool is created and destroyed on every call, this mode pays a small
start-up/teardown cost each time. For a single analysis that cost is
negligible, but if you call __loo__ functions repeatedly it can add up ---
which is exactly what the next sections address.

# 2. User-managed pools, including remote/HPC workers (recommended)

If you configure a pool yourself with `mirai::daemons()`, __loo__ will **reuse
it and never tear it down**. A user-managed pool always takes precedence over
the `cores` per-call behavior, which makes it the right tool for repeated
calls in one session and for **remote, SSH, or HPC cluster** workers.

A local pool you manage by hand behaves like a persistent session pool. Because
a connected pool always wins, you do not even need to pass `cores` --- __loo__
uses the pool whatever `cores` is:

```{r user-local, eval=FALSE}
library(mirai)

daemons(4)                       # start 4 local workers
loo1 <- loo(LLarr, r_eff = r_eff) # reuses your pool (cores not needed)
loo2 <- loo(LLarr, r_eff = r_eff) # reuses it again
daemons(0)                       # you decide when to stop them
```

For distributed computing, point __mirai__ at remote machines or a cluster
scheduler. __loo__ does not need to know any of the details --- it simply uses
whatever pool is connected:

```{r user-remote, eval=FALSE}
library(mirai)

# Launch workers on remote hosts over SSH (see ?mirai::daemons and
# ?mirai::ssh_config for the full set of options, including HPC launchers).
daemons(
  n = 8,
  url = host_url(),
  remote = ssh_config(c("ssh://node1", "ssh://node2"))
)

loo_big <- loo(large_loglik, r_eff = r_eff_big)

daemons(0)
```

# 3. Map-style workflows with `loo_mirai()`

When you have many independent jobs --- for example, one `psis()` or `loo()`
call per model in a simulation study --- [loo_mirai()] provides a thin
convenience wrapper around `mirai::mirai_map()`. Each element of `args_list`
is a named list of arguments passed to a \pkg{loo} function via `do.call()`:

```{r loo-mirai, eval=FALSE}
r_eff <- relative_eff(exp(example_loglik_array()))
ll_list <- list(model_a_loglik, model_b_loglik)

args_list <- lapply(ll_list, function(ll) {
  list(log_ratios = -ll, r_eff = r_eff)
})

# Start workers, map, tear down automatically.
results <- loo_mirai(psis, args_list, n_daemons = 4)
```

Alternatively, manage the pool yourself and call `loo_mirai()` without
`n_daemons` (or call `psis()`/`loo()` directly --- both reuse a connected
pool):

```{r loo-mirai-pool, eval=FALSE}
library(mirai)

daemons(4)
results <- loo_mirai(psis, args_list)
daemons(0)
```

When no pool is connected and `n_daemons` is `NULL`, `loo_mirai()` runs
serially. Do not pass `cores` in `args_list` --- parallelism is controlled by
`n_daemons` or an existing `mirai::daemons()` pool.

# How deprecated `cores` / `mc.cores` interact with a pool

The rule for user-managed pools is unchanged: **whenever a daemon pool is
connected, __loo__ always uses it and ignores deprecated parallel controls.**
In that state the *value* of `cores` does not matter --- `cores = 2` and
`cores = 6` behave identically, and even the default (`cores = 1` when
`mc.cores` is unset) still runs in parallel on the pool. The first time a call
would otherwise have looked like a request for serial execution (`cores <= 1`)
while a pool is connected, __loo__ prints a one-time message noting that
`cores` is being ignored.

Deprecated controls only start per-call pools when **no** mirai pool is
connected: `cores > 1` or `options(mc.cores = k)` with `k > 1` both resolve to
a temporary pool for that call (Section 1). If you genuinely want a single call
to run serially while a pool is up, stop the pool first with `mirai::daemons(0)`.

## When is the number of worker daemons equal to `cores`?

It follows directly from the rule above. The number of workers actually doing
the computation equals `cores` **only in the per-call case**; when a
user-managed pool is connected the worker count comes from your
`mirai::daemons()` call and `cores` is irrelevant:

| How parallelism is set up | Number of worker daemons | Equal to `cores`? |
|---|---|---|
| Per-call, no pool connected (`cores = k`, Section 1) | `k` | **Yes** --- loo spawns exactly `cores` daemons for the call |
| User-managed pool (`mirai::daemons(n)`, Section 2) | `n` (whatever you started) | Only if you happen to pass the same `cores` |

In other words, `cores` *is* the daemon count in the per-call path, but for a
user-managed pool the daemon count is fixed by your `mirai::daemons()` call,
and passing a different `cores` (including the default `1`) does **not** change
how many workers run. Internally __loo__ always splits the work according to
the number of daemons that are actually connected, not the requested `cores`.

# Large objects and shared memory

When __loo__ uses the function method (e.g. `loo(llfun, data = ..., draws =
...)`), the posterior `draws` are reused identically by every observation. On a
**local** pool these large objects are written once into shared memory with
__mori__ so each worker maps the same physical pages (zero-copy), which keeps
memory use low even with many workers.

On a **remote** pool shared memory is not available, so __loo__ falls back to
ordinary serialization and chunks the work to send each large object roughly
once per worker rather than once per observation. You don't need to do anything
to get this behavior --- __loo__ detects whether the active pool is local and
chooses the appropriate transport automatically.

# Choosing a strategy

| Situation | Recommended approach |
|---|---|
| A single `loo()`/`psis()` call | `mirai::daemons(k)` … `loo()` … `mirai::daemons(0)` (Section 2) |
| Default parallelism for a session | `mirai::daemons(k)` (Section 2) |
| Many calls in one session (simulations, benchmarks, batch jobs) | `mirai::daemons(k)` (Section 2) or `loo_mirai(..., n_daemons = k)` (Section 3) |
| Remote / SSH / HPC workers | `mirai::daemons(url = ..., remote = ...)` (Section 2) |
| Legacy code (deprecated bridge) | `cores = k` or `options(mc.cores = k)` (Section 1; warns once per session) |

A few general notes:

* All approaches give identical results; they differ only in scheduling and
  set-up cost.
* These mechanisms compose safely. A pool you set up yourself is always reused
  and never removed by __loo__, and nested __loo__ calls reuse an outer pool
  rather than creating a new one. For example, `loo_model_weights()` sets up a
  single pool and reuses it across all of the models it evaluates.
* Deprecated `cores` / `mc.cores` controls still work on this branch but emit
  a once-per-session warning; they start per-call pools only when no mirai
  pool is connected and the resolved core count is `> 1`.
* A connected pool always wins: when one is active, deprecated parallel
  controls are ignored and the work runs on the pool (the first time a
  `cores <= 1` call is overridden this way, __loo__ prints a one-time message).
  Stop the pool with `mirai::daemons(0)` if you need a call to run serially.
* On Windows, avoid setting deprecated parallel options in your `.Rprofile`
  (see [this issue](https://github.com/stan-dev/loo/issues/94)); use
  `mirai::daemons()` in your script instead.

# See also

* `help("loo", package = "loo")` for the `cores` argument documentation.
* `?loo_mirai` for the map-style convenience wrapper.
* The [__mirai__](https://mirai.r-lib.org/) documentation, especially
  `?mirai::daemons`, for configuring local, remote, and HPC pools.
