---
title: "Migration guide"
author: "loo package developers"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Migration guide}
-->
```{r, child="children/SETTINGS-knitr.txt"}
```

```{r, child="children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

# Introduction

This vignette documents API changes for the **predictive performance measures**
feature introduced on this branch. It lists deprecated functions, their
replacements, and the new `*_pred_measure()` workflow.

| Status | Meaning |
|--------|---------|
| **Deprecated** | Still works but emits a warning (via `.Deprecated()`). Use the replacement for new code. |
| **Current** | Supported API; use for new code. |

Release notes appear in `NEWS.md`. Deprecated functions are listed under
**Deprecated functions** on the package website.

## Scope of this guide (predictive performance branch)

This guide covers migrations for `elpd()`, `crps()`, `scrps()`, `loo_crps()`,
`loo_scrps()`, and `loo_predictive_metric()` to the unified `measure_*()` and
`*_pred_measure()` API.

Migrations for `compare()`, `psislw()`, loo-object extractors, and other
changes planned for **loo 3.0.0** are documented on the `loo-v3.0.0` branch
(see `dev/migration-guide-loo-v3.Rmd`). Those sections will be merged into a
single guide when both branches are integrated.

---

# Deprecated source files

| File | Status | Replacement / notes |
|------|--------|---------------------|
| `R/elpd.R` | Deprecated | `measure_elpd()` in `R/pred_measure-builtin.R` |
| `R/crps.R` | Deprecated | `measure_rps()`, `measure_srps()`, `loo_pred_measure()` |
| `R/loo_predictive_metric.R` | Deprecated | `loo_pred_measure()`, `measure_*()` |
| `R/pred_measure.R` | **Current** | `*_pred_measure()` entry points |
| `R/pred_measure-compute.R` | **Current** | Internal orchestration |
| `R/pred_measure-helpers.R` | **Current** | Validation and helpers |
| `R/pred_measure-builtin.R` | **Current** | `measure_*()` implementations |

---

# Function migration tables

## Expected log predictive density

| Deprecated | Replacement | Since | Planned removal |
|------------|-------------|-------|-----------------|
| `elpd()` | `measure_elpd()` | 3.0.0 | |

**Return type change:** `elpd()` returns class `"elpd_generic"` with `elpd` and
`ic` columns in `pointwise`. `measure_elpd()` returns class `"measure"`.

```r
# Deprecated
elpd(log_lik_matrix)

# Current
measure_elpd(log_lik_matrix)
```

## Scoring rules (CRPS / SCRPS)

| Deprecated | Replacement | Since | Planned removal |
|------------|-------------|-------|-----------------|
| `crps(x, x2, y)` | `measure_rps(y, ypred = x)` | 3.0.0 | |
| `scrps(x, x2, y)` | `measure_srps(y, ypred = x)` | 3.0.0 | |
| `loo_crps(...)` | `loo_pred_measure(..., measure = "rps")` | 3.0.0 | |
| `loo_scrps(...)` | `loo_pred_measure(..., measure = "srps")` | 3.0.0 | |

**API change:** deprecated functions require **two** draw matrices (`x`, `x2`)
and use a permutation-based estimator. The replacement uses **one** draw
matrix (`ypred`) and a PWM estimator. Results are not identical.

```r
# Deprecated
crps(ypred1, ypred2, y = y)

# Current
measure_rps(y = y, ypred = ypred1)
```

## Leave-one-out predictive metrics

| Deprecated | Replacement | Since | Planned removal |
|------------|-------------|-------|-----------------|
| `loo_predictive_metric()` | `loo_pred_measure()` | 3.0.0 | |

**Argument mapping:**

| Old (`loo_predictive_metric`) | New (`loo_pred_measure`) |
|-------------------------------|--------------------------|
| `x` (predictions) | `mupred` |
| `log_lik` | via `loo(..., save_psis = TRUE)` or `ylp` + `psis_object` |
| `metric = "mae"` | `measure = "mae"` |
| `metric = "balanced_acc"` | `measure = "bacc"` |

```r
# Deprecated
loo_predictive_metric(x = mupred, y = y, log_lik = ll, metric = "mae")

# Current
loo_fit <- loo(ll, save_psis = TRUE)
loo_pred_measure(loo = loo_fit, y = y, mupred = mupred, measure = "mae")
```

---

# Current predictive performance API (3.0.0)

## Workflow entry points

| Function | Evaluation mode | Source file |
|----------|-----------------|-------------|
| `insample_pred_measure()` | In-sample | `pred_measure.R` |
| `loo_pred_measure()` | PSIS-LOO | `pred_measure.R` |
| `kfold_pred_measure()` | K-fold CV | `pred_measure.R` |
| `test_pred_measure()` | Holdout test set | `pred_measure.R` |
| `pred_measure()` | Add measures incrementally | `pred_measure.R` |

## Built-in measures

Pass short names to the `measure` argument, or call standalone functions:

| `measure =` | Function | Source file |
|-------------|----------|-------------|
| `"elpd"` | `measure_elpd()` | `pred_measure-builtin.R` |
| `"mlpd"` | `measure_mlpd()` | `pred_measure-builtin.R` |
| `"ic"` | `measure_ic()` | `pred_measure-builtin.R` |
| `"mae"` | `measure_mae()` | `pred_measure-builtin.R` |
| `"mse"` | `measure_mse()` | `pred_measure-builtin.R` |
| `"rmse"` | `measure_rmse()` | `pred_measure-builtin.R` |
| `"r2"` | `measure_r2()` | `pred_measure-builtin.R` |
| `"acc"` | `measure_acc()` | `pred_measure-builtin.R` |
| `"bacc"` | `measure_bacc()` | `pred_measure-builtin.R` |
| `"rps"` | `measure_rps()` | `pred_measure-builtin.R` |
| `"srps"` | `measure_srps()` | `pred_measure-builtin.R` |

List all registered built-in names:

```r
supported_measures_list()
```

`measure_brier()` is exported but not yet registered in `supported_measures_list()`.

---

# Maintainer checklist

When deprecating or removing a function on this branch:

1. Add `.Deprecated("replacement")` in each exported method (only if the
   replacement exists in the same branch).
2. Document deprecation in roxygen (`**deprecated**`, links to replacement).
3. Add a bullet to `NEWS.md`.
4. Update this migration guide.
5. Add or update tests in `tests/testthat/test_deprecated_measures.R`.
6. List the topic under **Deprecated functions** in `_pkgdown.yml`.
7. Keep the old implementation until the planned removal release.
8. On removal: delete the source file, remove from `NAMESPACE`, update this guide.

Do **not** add `compare()` / `psislw()` migration content here; that belongs on
the `loo-v3.0.0` branch migration guide.

---

# `_pkgdown.yml` reference (this branch)

Suggested `_pkgdown.yml` entries for this branch:

```yaml
articles:
  - title: Migration
    desc: |
      Guides for updating code when APIs change.
    contents:
      - migration-guide

reference:
  - title: Predictive performance measures
    desc: |
      Unified API for predictive performance measures (density scores, scoring
      rules, and point-prediction metrics) with in-sample, LOO, K-fold, and
      holdout evaluation modes.
    contents:
      - insample_pred_measure
      - loo_pred_measure
      - kfold_pred_measure
      - test_pred_measure
      - pred_measure
      - supported_measures_list
      - starts_with("measure_")
      - ptw_log_pred_density
  - title: Deprecated functions
    contents:
      - elpd
      - crps
      - loo_predictive_metric
```

After merging into `loo-v3.0.0`, combine this with the `compare` and `psislw`
entries from that branch's migration guide.
