---
title: "Migration guide"
author: "loo package developers"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Migration guide}
-->
```{r, child="children/SETTINGS-knitr.txt"}
```

```{r, child="children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

# Introduction

This vignette documents API changes for **loo 3.0.0**. It lists deprecated
functions, their replacements, and what is planned for removal in the major
release.

| Status | Meaning |
|--------|---------|
| **Deprecated** | Still works but emits a warning (via `.Deprecated()`). Use the replacement for new code. |
| **Current** | Supported API; use for new code. |
| **Planned removal** | Scheduled to be deleted in loo 3.0.0. |

Release notes appear in `NEWS.md`. Deprecated functions are listed under
**Deprecated functions** on the package website.

## Scope of this guide (loo 3.0.0 branch)

This version of the guide covers deprecations whose **replacements already exist**
in the package. Work on a unified predictive performance API (`pred_measure`,
`measure_*()`, and related deprecations) is developed on a separate feature
branch and will extend this guide when merged into `loo-v3.0.0`.

---

# Deprecated source files

| File | Status | Replacement / notes |
|------|--------|---------------------|
| `R/elpd.R` | Deprecated | `measure_elpd()` in `R/pred_measure-builtin.R` |
| `R/crps.R` | Deprecated | `measure_rps()`, `measure_srps()`, `loo_pred_measure()` |
| `R/loo_predictive_metric.R` | Deprecated | `loo_pred_measure()`, `measure_*()` |
| `R/pred_measure.R` | **Current** | `*_pred_measure()` entry points |
| `R/pred_measure-compute.R` | **Current** | Internal orchestration |
| `R/pred_measure-helpers.R` | **Current** | Validation and helpers |
| `R/pred_measure-builtin.R` | **Current** | `measure_*()` implementations |
| `R/compare.R` | Deprecated (removal in 3.0.0) | `loo_compare()` in `R/loo_compare.R` |
| `R/psislw.R` | Deprecated (removal in 3.0.0) | `psis()` in `R/psis.R` (since 2.0.0) |
| `R/loo_compare.R` | **Current** | Model comparison |
| `R/loo.R`, `R/psis.R`, `R/waic.R` | **Current** | Core LOO / PSIS / WAIC |

---

# Function migration tables

## Expected log predictive density

| Deprecated | Replacement | Since | Planned removal |
|------------|-------------|-------|-----------------|
| `elpd()` | `measure_elpd()` | 3.0.0 | |

**Return type change:** `elpd()` returns class `"elpd_generic"` with `elpd` and
`ic` columns in `pointwise`. `measure_elpd()` returns class `"measure"`.

```r
# Deprecated
elpd(log_lik_matrix)

# Current
measure_elpd(log_lik_matrix)
```

## Scoring rules (CRPS / SCRPS)

| Deprecated | Replacement | Since | Planned removal |
|------------|-------------|-------|-----------------|
| `crps(x, x2, y)` | `measure_rps(y, ypred = x)` | 3.0.0 | |
| `scrps(x, x2, y)` | `measure_srps(y, ypred = x)` | 3.0.0 | |
| `loo_crps(...)` | `loo_pred_measure(..., measure = "rps")` | 3.0.0 | |
| `loo_scrps(...)` | `loo_pred_measure(..., measure = "srps")` | 3.0.0 | |

**API change:** deprecated functions require **two** draw matrices (`x`, `x2`)
and use a permutation-based estimator. The replacement uses **one** draw
matrix (`ypred`) and a PWM estimator. Results are not identical.

```r
# Deprecated
crps(ypred1, ypred2, y = y)

# Current
measure_rps(y = y, ypred = ypred1)
```

## Leave-one-out predictive metrics

| Deprecated | Replacement | Since | Planned removal |
|------------|-------------|-------|-----------------|
| `loo_predictive_metric()` | `loo_pred_measure()` | 3.0.0 | |

**Argument mapping:**

| Old (`loo_predictive_metric`) | New (`loo_pred_measure`) |
|-------------------------------|--------------------------|
| `x` (predictions) | `mupred` |
| `log_lik` | via `loo(..., save_psis = TRUE)` or `ylp` + `psis_object` |
| `metric = "mae"` | `measure = "mae"` |
| `metric = "balanced_acc"` | `measure = "bacc"` |

```r
# Deprecated
loo_predictive_metric(x = mupred, y = y, log_lik = ll, metric = "mae")

# Current
loo_fit <- loo(ll, save_psis = TRUE)
loo_pred_measure(loo = loo_fit, y = y, mupred = mupred, measure = "mae")
```

---

# Current predictive performance API (3.0.0)

## Workflow entry points

| Function | Evaluation mode | Source file |
|----------|-----------------|-------------|
| `insample_pred_measure()` | In-sample | `pred_measure.R` |
| `loo_pred_measure()` | PSIS-LOO | `pred_measure.R` |
| `kfold_pred_measure()` | K-fold CV | `pred_measure.R` |
| `test_pred_measure()` | Holdout test set | `pred_measure.R` |
| `pred_measure()` | Add measures incrementally | `pred_measure.R` |

## Built-in measures

Pass short names to the `measure` argument, or call standalone functions:

| `measure =` | Function | Source file |
|-------------|----------|-------------|
| `"elpd"` | `measure_elpd()` | `pred_measure-builtin.R` |
| `"mlpd"` | `measure_mlpd()` | `pred_measure-builtin.R` |
| `"ic"` | `measure_ic()` | `pred_measure-builtin.R` |
| `"mae"` | `measure_mae()` | `pred_measure-builtin.R` |
| `"mse"` | `measure_mse()` | `pred_measure-builtin.R` |
| `"rmse"` | `measure_rmse()` | `pred_measure-builtin.R` |
| `"r2"` | `measure_r2()` | `pred_measure-builtin.R` |
| `"acc"` | `measure_acc()` | `pred_measure-builtin.R` |
| `"bacc"` | `measure_bacc()` | `pred_measure-builtin.R` |
| `"rps"` | `measure_rps()` | `pred_measure-builtin.R` |
| `"srps"` | `measure_srps()` | `pred_measure-builtin.R` |

List all registered built-in names:

```r
supported_measures_list()
```

`measure_brier()` is exported but not yet registered in `supported_measures_list()`.
## Model comparison

| Deprecated | Replacement | Since | Planned removal |
|------------|-------------|-------|-----------------|
| `compare()` | `loo_compare()` | 2.0.0 | 3.0.0 |

`compare()` returns a vector or matrix with class `"compare.loo"`.
`loo_compare()` returns a `data.frame` with additional diagnostic columns.
See `?loo_compare` for the current output format.

```r
# Deprecated
compare(loo1, loo2)

# Current
loo_compare(loo1, loo2)
```

## PSIS importance sampling

| Deprecated | Replacement | Since | Planned removal |
|------------|-------------|-------|-----------------|
| `psislw()` | `psis()` | 2.0.0 | 3.0.0 |
| `options(loo.cores = ...)` | `options(mc.cores = ...)` or `cores` argument | 2.0.0 | |

```r
# Deprecated
psislw(-log_lik_matrix)

# Current
psis(-log_lik_matrix)
```

## Accessing estimates from loo and waic objects

Direct extraction of summary estimates via `$`, `[`, or `[[` on loo objects
has been deprecated since loo 2.x. Use the `estimates` component instead.

| Deprecated | Replacement | Since | Planned removal |
|------------|-------------|-------|-----------------|
| `x$elpd_loo`, `x$looic`, etc. | `x$estimates["elpd_loo", "Estimate"]` | 2.x | |
| `x["elpd_loo"]`, `x[["elpd_loo"]]` | `x$estimates[...]` | 2.x | |

The same pattern applies to `waic` objects (`elpd_waic`, `waic`, etc.).

## Argument renames

| Function | Deprecated | Replacement | Since |
|----------|------------|-------------|-------|
| `psis_approximate_posterior()` | `log_q` | `log_g` | 2.x |

---

# Planned removals in 3.0.0

The following are targeted for **removal** in loo 3.0.0:

- `compare()` and `R/compare.R`
- `psislw()` and `R/psislw.R`

The deprecated option `options(loo.cores = ...)` remains for backward
compatibility until a separate removal decision is made.

---

# Upcoming (not yet in this branch)

The following deprecations and migrations are **not** covered by this version
of the guide. They will be added when the predictive performance feature branch
is merged into `loo-v3.0.0`:

- `elpd()` → `measure_elpd()`
- `crps()`, `scrps()`, `loo_crps()`, `loo_scrps()` → `measure_rps()`, `measure_srps()`, `loo_pred_measure()`
- `loo_predictive_metric()` → `loo_pred_measure()` and `measure_*()`
- New workflow entry points: `insample_pred_measure()`, `loo_pred_measure()`,
  `kfold_pred_measure()`, `test_pred_measure()`, `pred_measure()`

Until that merge, do **not** add `.Deprecated("measure_*")` or
`.Deprecated("loo_pred_measure")` warnings on this branch, because those
replacements are not yet exported here.

---

# Maintainer checklist

When deprecating or removing a function on this branch:

1. Add `.Deprecated("replacement")` in each exported method (only if the
   replacement exists in the same branch).
2. Document deprecation in roxygen (`**deprecated**`, links to replacement).
3. Add a bullet to `NEWS.md`.
4. Update this migration guide.
5. Add or update tests in `tests/testthat/test_deprecated_*.R`.
6. List the topic under **Deprecated functions** in `_pkgdown.yml`.
7. Keep the old implementation until the planned removal release.
8. On removal: delete the source file, remove from `NAMESPACE`, update this guide.

When merging feature branches that introduce new replacements, extend the
**Function migration tables** and remove items from **Upcoming**.

---

# `_pkgdown.yml` reference (this branch)

Suggested `_pkgdown.yml` entries for this branch:

```yaml
articles:
  - title: Migration
    desc: |
      Guides for updating code when APIs change.
    contents:
      - migration-guide

reference:
  - title: Predictive performance measures
    desc: |
      Unified API for predictive performance measures (density scores, scoring
      rules, and point-prediction metrics) with in-sample, LOO, K-fold, and
      holdout evaluation modes.
    contents:
      - insample_pred_measure
      - loo_pred_measure
      - kfold_pred_measure
      - test_pred_measure
      - pred_measure
      - supported_measures_list
      - starts_with("measure_")
      - ptw_log_pred_density
  - title: Deprecated functions
    contents:
      - elpd
      - crps
      - loo_predictive_metric
      - compare
      - psislw
```

After the predictive performance feature merges, add `elpd`, `crps`, and
`loo_predictive_metric` to **Deprecated functions**, and add a **Predictive
performance measures** reference section for the new API.
