---
title: "Differences and their standard errors in model comparison"
author: "Florence Bockting"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    css:
      - !expr system.file("rmarkdown/templates/html_vignette/resources/vignette.css", package = "rmarkdown")
      - styles.css
    toc: true
    toc_depth: 3
params:
  EVAL: TRUE #!r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{comparison-standard-errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="../children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = TRUE,
  warning = TRUE,
  eval = params$EVAL,
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

::: {.callout .callout-warning}
**Questions**

+ How detailed should the math for the computation be and how to attribute properly the work from the overleaf document? Are you considering to publish it? -> Answered: we target for a preprint that we can cite.
+ Currently, `elpd` is computed by default in `pred_measure`. This is helpful, when we want to have it by default in `model_compare`. Do we want this default behavior?
+ Related question: The default print method of `model_compare` shows only the `elpd`, only when `print(..., measures = "all")` is passed, all measures are shown. Do we want this?
+ For `elpd`, we provide in the `model_compare` output the additional columns `p_worse`, `diag_diff`, `diag_elpd`.
  + Do we want to have `p_worse` and `diag_diff` for all measures?
  + I moved the column `diag_elpd` out from the elpd-specific measure block and treat it as an "overall" information (see printed output below).
  + Furthermore, I renamed `diag_elpd` to `diag_psis` as it seems not be specific to elpd or do I misunderstand this?
+ General: Is the current design of `model_compare()` okay?
+ Which warnings/messages do we want to provide and which should only be part of the function documentation?
:::

```{r load-libraries}
library(loo)
```

::: {.callout .callout-note}
**Acknowledgements**

We thank Seth Axen, Frank Weber, and Aki Vehtari for deriving the formulas of
the standard errors for LOO-CV estimators and differences.
:::

## Introduction
In the following, our goal is to compare performance between two or more
models. To do this, we will use the function `loo::model_compare()`.

`model_compare()` compares model performance between multiple models based
on one or more performance criteria. We provide a list of built-in criteria
including `elpd, ic, mlpd, mae, mse, acc, brier, rps, srps, rmse, r2, bacc`
(see the [Overview of scores and metrics](overview-measures.html)
article).
Additionally, it is possible to define a custom performance criterion.

The `model_compare()` function accepts any object that inherits from the `loo`
class. This includes outputs from:

+ `loo()`, `kfold()`, and `waic()`
+ all `<type>_pred_measure` functions (with `<type>` being a placeholder for `loo`, `kfold`, `test`, or `insample`)

In this vignette we focus on the comparison of **predictive performance
measures** produced by `<type>_pred_measure()` and explain:

1. The approaches for computing the point estimate and standard error of the
   difference for each `<type>` and built-in measure.
2. How to compute the point estimate and standard error of the difference for
   a **custom** measure.

::: {.callout .callout-note}
**Relation between `model_compare` and `loo_compare`**

The `loo_compare` function has traditionally been used for model comparison in
loo, accepting any object that inherits from the `loo` class (such as those
produced by `loo()`, `kfold()`, and `waic()`).

With the introduction of the *`pred_measure` family*, we have deprecated
`loo_compare` and introduced `model_compare`. While `model_compare` retains full
backward compatibility, it also expands functionality to handle outputs from
the `pred_measure` family.

`loo_compare` still works and is still a generic, so methods registered by other
packages keep dispatching, but it now warns and is frozen at its previous
behavior: `loo` objects compared on ELPD only. The `pred_measure` sources,
`rank_by`, and `custom_se_fn` described in this article require `model_compare`.
`loo_compare` will be removed in a future release. For more details, see
`vignette("migration-guide", package = "loo")`.
:::

## Running example (count data)

To illustrate the functionality of `model_compare()` we first introduce a
*running example*. In this example, we compare three Poisson regression
models fitted to the **`roaches` data** from `rstanarm`
(Gelman and Hill, 2007). This data set records a pest-management trial in
262 apartment buildings. The outcome `y`
is the number of roaches caught in traps after treatment and ranges from 0 to
357. The predictors are:

+ `roach1`: pre-treatment roach count (used as `sqrt_roach1 = sqrt(roach1)`)
+ `treatment`: whether the building received pest management
+ `senior`: whether the building is restricted to elderly residents

The **three models** used in the following comparison tasks are defined as
follows:

```
  m1:      y ~ treatment + senior
  m2:      y ~ sqrt_roach1 + senior
  m3:      y ~ sqrt_roach1 + treatment
```

As **performance criteria** we select six predictive performance measures:
`elpd`, `mae`, `mse`, `rmse`, `r2`, and `rps` computed by `loo_pred_measure`.
Note that `rps` is the only one of these that needs the posterior *predictive*
draws `ypred`; the error measures are computed from the posterior expectation
`mupred` (see the input requirements table in `?loo_pred_measure`).

<!--
Provide a section where we use kfold_pred_measure, test_pred_measure, etc.
-->

```{r setup-example}
measures <- c("mae", "mse", "rmse", "r2", "rps") # elpd is computed by default
```

## The classical `loo_compare` output

As the actual model building workflow is not the focus of this tutorial, we skip
this part and load the results from a pre-fitted object in the variable `res`.
Among other things, it stores a `loo` object (as returned by `loo::loo()`) for
each of the three models.

```{r example}
path <- "../../tests/testthat/data-for-tests/test_data_roaches_compare.Rds"
res <- readRDS(path)

loos <- list(m1 = res$loo_p_m1, m2 = res$loo_p_m2, m3 = res$loo_p_m3)
```

Passing these `loo` objects to the classical `loo_compare()` reproduces the
familiar ELPD-only comparison, and warns once per session that `model_compare()`
is now the preferred function:

```{r loo_compare, warning=TRUE}
loo_compare(loos)
```

When we use `model_compare` instead, we get the same output for `loo` objects:

```{r model_compare_wloo, warning=TRUE}
model_compare(loos)
```

## Preview: Glimpse into the model comparison results

We now use `loo_pred_measure()` to compute the predictive performance for
each model (`m1, m2, m3`) and measure (see list above).
Finally, we compare the performance of the three models using `model_compare()`.
By default, only the `elpd` measure is shown.

<a id="example"></a>
```{r pred-measure}
fit_measure <- function(i, measure) {
  loo_pred_measure(
    y = res$y,
    ypred = res[[paste0("ypred_m", i)]],
    mupred = res[[paste0("mupred_m", i)]],
    ylp = res[[paste0("ylp_m", i)]],
    measure = measure
  )
}

m1 <- fit_measure(1, measures)
m2 <- fit_measure(2, measures)
m3 <- fit_measure(3, measures)

(comp <- model_compare(list(m1 = m1, m2 = m2, m3 = m3)))
```

When we want to see a specific measure, we can change this by passing the 
information to the `print` method:

```{r pred-measure-r2}
print(comp, measures = "r2")
```

Or in order to see all measures we can use `measures = "all"`:

```{r pred-measure-all}
print(comp, measures = "all")
```

### Output of `model_compare()`

Using `print(., measures = "all")` returns a subsection for each measure.
The *rows* represent the different models, with the best model always in the
first row, followed by the second best, and so on.
The *columns*:

  + `model`: Name of each model
  + `<measure>_diff`: The point estimate representing the difference in the measure between the current model and the reference model$^{(1)}$ (calculated as model - reference). For the reference model itself, this value is always zero.
  + `<measure>_se_diff`$^{(2)}$: The standard error of the difference
  + `p_worse`: (only for `elpd`) Estimated probability that a model has worse predictive performance than the reference model
  + `diag_diff`: (only for `elpd`) Diagnostic column flagging miscalibration of the normal approximation behind `p_worse` and `se_diff`

Above the measure subsections, `print()` names the reference model each set of
differences was computed against, and, for `loo_pred_measure()` inputs, reports
the PSIS-LOO diagnostic once per model:

  + `bad_k`: Number of observations whose Pareto $\hat{k}$ exceeds the PSIS
    reliability threshold. Pareto $\hat{k}$ describes a model's PSIS-LOO
    approximation rather than any single measure or any pair of models --- all
    LOO measures are computed from the same importance weights, and the value
    does not depend on the reference model. It is therefore reported once per
    model instead of as a column in the difference tables. On the returned
    object the same information is available in the `diag_elpd` column.

This block is omitted when no model is flagged, and for `kfold`, `test`, and
`insample` sources, which have no Pareto $\hat{k}$ diagnostic.

$^{(1)}$ By default, the reference model is the best model (per measure). For additional information, see Section [Selecting the reference model](#selecting-the-reference-model)

$^{(2)}$ The only measure that deviates from this naming convention is `elpd`, where the column is called `se_diff` instead of `elpd_se_diff` for backwards compatibility.

### The `measure_info` attribute

Each model has a `measure_info` attribute, which provides meta
information for each performance measure.

```{r meta-attribute, class.output="scroll-output"}
meta <- attr(m1, "measure_info")

# measure info for each single measure (e.g., R2)
names(meta$r2)
```

The information is stored in different *slots*:

| slot | description | functionality |
| :--- | :---- | :----- |
| `loss` | Whether the measure is a *loss* (`TRUE`; lower values of the measure are better) or a *utility* (`FALSE`, higher values are better) | During model comparison, loss measures have their sign flipped so that all differences are reported on a utility scale (higher is better) |
| `diff_method` | How paired pointwise differences are aggregated within model comparison: `"sum"`, `"mean"`, `"measure_specific"`, or `"custom"` | Selects the formula for `<measure>_se_diff` (see the table in Section [Computation of the standard error](#computation-of-the-standard-error-measure_se_diff)) |
| `se_diff_fun` | Name of the measure's own standard-error function for computing `<measure>_se_diff`, only set when `diff_method = "measure_specific"` (`"rmse"`, `"r2"`, `"bacc"`) | Points to the built-in function that computes `<measure>_se_diff` for measures whose estimate is not a plain sum or mean of pointwise values |
| `extra` | Optional auxiliary data the measure stored for its `se_diff_fun`, on the measure's natural scale (e.g., `mse_y_i` for `r2`, `class_id` for `bacc`); `NULL` for measures that need none | Carries the quantities the standard error needs but the pointwise values do not contain; passed to `se_diff_fun()` (and to `custom_se_fn()`) as part of each model's input |

## Background: Computation underlying `model_compare` output

In this section, we explain (1) how the point estimate `<measure>_diff` and the
(2) standard error of the difference `<measure>_se_diff` are computed for
each measure. While the point estimate is computed in the same way for all
measures, the computation of the standard error depends on the
measure.

### Computation of the point estimate `<measure>_diff`
The estimates `elpd_diff`, `mae_diff`, `rmse_diff`, and `r2_diff` in the
output of the running example represent the point estimates of the pairwise
model comparisons.
They are computed as the measure-specific difference between two model
estimates (comparison model - reference model).

For example, the difference in `elpd` between model `m1` (comparison) and model
`m3` (reference) is computed as follows:

```{r, eval=FALSE}
m1_elpd <- m1$estimates["elpd_loo", "Estimate"]
m3_elpd <- m3$estimates["elpd_loo", "Estimate"]

m1_elpd - m3_elpd
```

This computation is done for each measure and each pairwise model comparison.

By default, however, the reference model is chosen *per measure* as the best
model for that measure (see Section
[Selecting the reference model](#selecting-the-reference-model)), so `m3` is not
the reference for every measure. The reference used for each measure is recorded
in the `compare_reference` attribute:

```{r reference-models}
models <- list(m1 = m1, m2 = m2, m3 = m3)
ref <- attr(comp, "compare_reference")
ref
```

The following code block shows, for model `m1`, the hand-computed differences
against each measure's own reference model and the differences reported in the
`comp` object:

```{r point-estimate}
est <- function(x, measure) x$estimates[measure, "Estimate"]
measures_all <- c("elpd", "mse", "mae", "rmse", "r2")

data.frame(
  reference = ref[measures_all],
  by_hand   = sapply(measures_all, function(m) {
    est(m1, paste0(m, "_loo")) - est(models[[ref[[m]]]], paste0(m, "_loo"))
  }),
  reported  = sapply(measures_all, function(m) {
    comp[[paste0(m, "_diff")]][comp$model == "m1"]
  })
)
```

Note that in the above output, the sign for some measures is flipped between
the `by_hand` and `reported` column, specifically for `mse`, `mae`, and `rmse`.
These measures are *losses* (see callout box below).

However, when we compare all measures together, we want them all to be
on the same scale, which in our case is the **utility scale**. Therefore, losses
are converted to the utility scale. As such, higher values are always better.
Consequently, the best model has the highest value, and in the default case in
which the best model is also the reference model, the estimated difference
`<measure>_diff` (computed as comp - reference) is always non-positive.

When a measure is converted to the utility scale, `model_compare()` informs the
user about this change:

```{r sign-message, warning=TRUE, message=TRUE}
comp <- model_compare(list(m1 = m1, m2 = m2, m3 = m3))
```

::: {.callout .callout-note}
**Utility vs. loss scale**

We call a measure a **loss** when it has, by default, the orientation
*lower is better*.

By contrast, we call a measure a **utility** if it has, by default, the
orientation *higher is better*.

Which orientation a measure has can be inspected from the `loss` slot of the
`measure_info` attribute of the model object:
```{r measure-meta}
sapply(meta, function(x) x$loss)
```

:::

### Computation of the standard error `<measure>_se_diff`

<!-- Point towards material that explains the elpd diagnostics -->
<!-- Provide reference to glossary -->

The standard error summarizes the uncertainty in the estimated difference.
Depending on the measure, different approaches for computing the standard error
are used. We summarize them in four different categories:

| `diff_method` | SE of the difference | Measures |
| :--- | :--- | :--- |
| `"sum"` | `sqrt(N) * sd(d_i)` | `elpd`, `ic` |
| `"mean"` | `sd(d_i) / sqrt(N)` | `mlpd`, `mae`, `mse`, `acc`, `brier`, `rps`, `srps` |
| `"measure_specific"` | the measure's own `se_diff_fun` | `rmse`, `r2`, `bacc` |
| `"custom"` | supplied via `model_compare(custom_se_fn = )` | custom measures |

where `N` is the number of observations and `d_i` the paired difference of
the pointwise estimates of the two models (`comparison - reference`).

The user can check which approach is used for each measure via the
`diff_method` slot in the `measure_info` attribute of the model object:

```{r diff-method}
sapply(meta, function(x) x$diff_method)
```

#### Approach 1: "sum" {#approach-sum}

For `elpd` and `ic` the model estimate is a **sum** of pointwise contributions,
e.g., for `elpd`

$$
\widehat{\text{elpd}}_{\text{loo}} = \sum_{i=1}^{N} \widehat{\text{elpd}}_{\text{loo},i}.
$$

Let $d_i$ be the pointwise difference for observation $i$ between the
comparison model $M_a$ and the reference model $M_b$,

$$
d_i = \widehat{\text{elpd}}_{\text{loo},i}(M_a) - \widehat{\text{elpd}}_{\text{loo},i}(M_b),
$$

and let $D = \sum_{i=1}^{N} d_i$ be the estimated total difference. Treating the
$d_i$ as independent draws from a distribution with variance $\sigma_d^2$,

$$
\text{Var}(D) = \sum_{i=1}^{N} \text{Var}(d_i) = N \sigma_d^2
\quad\Longrightarrow\quad
\text{SE}(D) = \sqrt{N}\, \sigma_d .
$$

In practice $\sigma_d$ is replaced by the sample standard deviation of the
pointwise differences, which gives the estimator used by `model_compare()`:

$$
\widehat{\text{SE}}(D) = \sqrt{N} \cdot \text{sd}(d)
 = \Big( N \cdot \widehat{\text{Var}}(d) \Big)^{1/2}.
$$

```{r sum-method}
n <- attr(m1, "dims")[2]
d_elpd <- m1$pointwise[, "elpd_loo"] - m3$pointwise[, "elpd_loo"]

c(
  by_hand    = sqrt(n) * sd(d_elpd),
  reported   = comp$se_diff[comp$model == "m1"]
)
```

#### Approach 2: "mean" {#approach-mean}

For `mlpd`, `mae`, `mse`, and most scoring rules, the model estimate is a
**mean** of pointwise contributions, e.g., for `mse`

$$
\widehat{\text{mse}}_{\text{loo}} = \frac{1}{N} \sum_{i=1}^{N} \widehat{\text{mse}}_{\text{loo},i}.
$$

With $d_i$ defined as in [Approach 1](#approach-sum), the reported difference is
the *mean* pointwise difference $\bar{d} = \frac{1}{N}\sum_{i=1}^{N} d_i = D/N$.
As $\bar{d}$ is just $D$ rescaled by the constant $1/N$,

$$
\text{SE}(\bar{d}) = \frac{1}{N}\, \text{SE}(D) = \frac{1}{N} \sqrt{N}\, \sigma_d
= \frac{\sigma_d}{\sqrt{N}},
$$

that is, the familiar standard error of a mean. Replacing $\sigma_d$ by the
sample standard deviation of the pointwise differences gives the estimator used
by `model_compare()`:

$$
\widehat{\text{SE}}(\bar{d}) = \frac{\text{sd}(d)}{\sqrt{N}}.
$$

By default every measure uses its *own* best model as the reference (see
Section [Selecting the reference model](#selecting-the-reference-model)). For
`mse` that is `m2`, not `m3` as for `elpd` above:

```{r mean-method}
d_mse <- m1$pointwise[, "mse_loo"] - m2$pointwise[, "mse_loo"]

c(
  by_hand  = sd(d_mse) / sqrt(n),
  reported = comp$mse_se_diff[comp$model == "m1"]
)
```

Note that `mse` is a loss, so `model_compare()` flips its sign before taking the
difference (see the callout on *utility vs. loss scale* above). However,
flipping the sign leaves `sd(d)` unaffected, which is why the raw pointwise
differences above reproduce the reported value.

#### Approach 3: "measure_specific" {#approach-measure-specific}

The estimate of these measures is neither a sum nor a mean of pointwise
contributions, so their standard error cannot be obtained from the paired
pointwise differences alone. Instead, each such measure ships its own formula,
recorded in the `se_diff_fun` slot of the `measure_info` attribute. We
give the formulas for the built-in measures `rmse`, `r2`, and `bacc` below.

**RMSE**

For RMSE, the delta method (i.e., a first-order Taylor series approximation;
see Ver Hoef, 2012, for its history) on the MSE scale is used
and propagated through the square root:

$$
SE_{RMSE}(M_a, M_b) = \frac{1}{2}\sqrt{
  \frac{SE_{MSE}(M_a)^2}{{MSE}(M_a)} + \frac{SE_{MSE}(M_b)^2}{{MSE}(M_b)}
  - \frac{2 \mathrm{Cov}_{MSE}[M_a, M_b]}{{RMSE}(M_a) \cdot {RMSE}(M_b)}
}
$$

All terms on the right-hand side are computable from the stored pointwise
squared errors:

```{r pairwise-rmse}
# `rmse` is compared against its own reference model, `ref[["rmse"]]`
sqe_a <- models[[ref[["rmse"]]]]$pointwise[, "rmse_loo"]
sqe_b <- m1$pointwise[, "rmse_loo"]

mse_a <- mean(sqe_a)
mse_b <- mean(sqe_b)
se_a <- sqrt(var(sqe_a) / n)
se_b <- sqrt(var(sqe_b) / n)
cov_rc <- sum((sqe_a - mse_a) * (sqe_b - mse_b)) / (n * (n - 1))

se_rmse <- 0.5 * sqrt(
  se_a^2 / mse_a + se_b^2 / mse_b -
    2 * cov_rc / (sqrt(mse_a) * sqrt(mse_b))
)

c(
  by_hand = se_rmse,
  reported = comp$rmse_se_diff[comp$model == "m1"]
)
```

**R2**

The R2 used here is the leave-one-out version of the classical formula,
$1 - \text{MSE}(M_k) / \text{MSE}(y)$, rather than Bayesian R2
(Gelman et al., 2019).
For R2, let $d_i = \text{sqe}_i(M_a) - \text{sqe}_i(M_b)$ be the
pointwise difference in squared errors and
$\text{MSE}(M_a, M_b) = \frac{1}{n}\sum_{i=1}^n d_i$ its mean. The
difference in R2 is a ratio of two quantities estimated from the same data:

$$
R^2(M_a) - R^2(M_b) = -\frac{\text{MSE}(M_a, M_b)}{\text{MSE}(y)}
$$

A trivariate first-order Taylor expansion of that ratio gives

$$
SE_{R^2}(M_a, M_b) = \frac{1}{\text{MSE}(y)} \sqrt{
  SE_{MSE}(M_a, M_b)^2\\
  - 2\,\frac{\text{MSE}(M_a, M_b)}{\text{MSE}(y)}\,
    \mathrm{Cov}\!\left[\text{MSE}(M_a, M_b),\, \text{MSE}(y)\right]\\
  + \frac{\text{MSE}(M_a, M_b)^2}{\text{MSE}(y)^2}\,
    \mathrm{Var}\!\left[\text{MSE}(y)\right]
}
$$

With $c = \text{MSE}(M_a, M_b) / \text{MSE}(y)$ and the pointwise baseline
$\text{sqe}(y)_i = (y_i - \bar{y})^2$, the expression above is equal to

$$
SE_{R^2}(M_a, M_b) = \frac{1}{\text{MSE}(y)} \cdot
  \frac{\mathrm{sd}\!\left(d_i - c\,\text{sqe}(y)_i\right)}{\sqrt{n}}
$$

This is the form used in **loo**: it cannot go negative under the square root,
and it is exactly zero when a model is compared against itself. The computation
needs the pointwise baseline $\text{sqe}(y)_i$, which is stored in the `extra`
slot of the `measure_info` attribute:

```{r r2-extra}
str(attr(m1, "measure_info")$r2$extra)
```

The following code block shows the hand-computed standard error and the
reported standard error in the `comp` object for model `m1` against the
reference model used for `r2`:

```{r pairwise-r2}
mse_y_i <- attr(m1, "measure_info")$r2$extra$mse_y_i
mse_y   <- mean(mse_y_i)

d_sqe  <- m1$pointwise[, "r2_loo"] - models[[ref[["r2"]]]]$pointwise[, "r2_loo"]
scaled <- d_sqe - (mean(d_sqe) / mse_y) * mse_y_i

c(by_hand  = sd(scaled) / sqrt(n) / mse_y,
  reported = comp$r2_se_diff[comp$model == "m1"])
```

**bacc**

For the balanced accuracy (bacc), let
$d_i = \text{acc}_i(M_a) - \text{acc}_i(M_b)$ be the difference
in pointwise accuracies. Because every class gets the same weight regardless of
its size, the difference in bacc is not a mean of the $d_i$, but a mean of the
$K$ class-wise mean differences:

$$
\text{bacc}(M_a, M_b) = \frac{1}{K} \sum_{c=1}^{K} \bar{d}_c,
\qquad \bar{d}_c = \frac{1}{n_c} \sum_{i \in c} d_i
$$

where $K$ is the number of classes and $n_c$ the number of observations in
class $c$. The variance of the class mean is
$\text{Var}(\bar{d}_c) = s_c^2 / n_c$, and

$$
SE_{\text{bacc}}(M_a, M_b) = \frac{1}{K} \sqrt{\sum_{c=1}^{K} \frac{s_c^2}{n_c}},
\qquad s_c^2 = \frac{1}{n_c - 1} \sum_{i \in c} \left(d_i - \bar{d}_c\right)^2
$$

with $s_c^2 := 0$ for any class holding a single observation.
Writing $b_c$ and $m_c$ for the number of
observations in class $c$ that only $M_a$, respectively only $M_b$, classifies
correctly, the class term is a McNemar-type variance of a paired difference of
proportions (McNemar, 1947),

$$
\frac{s_c^2}{n_c} = \frac{1}{n_c - 1}
  \left(\frac{b_c + m_c}{n_c} - \frac{(b_c - m_c)^2}{n_c^2}\right)
$$

For binary outcomes balanced accuracy is
$(\text{sensitivity} + \text{specificity}) / 2$, so the difference above
is the estimand of Newcombe (2001) at a mixing parameter of one half, and half
the difference of two Youden indices in a paired design (Chen et al., 2015).

To demonstrate the computation of the standard error of the difference for bacc,
we introduce another example where the outcome variable is binary.
We simulate 200 observations from a logistic model with a single continuous
predictor `x`, and fit two Bernoulli models: `m_a` uses the predictor `x` that
generated the data, while `m_b` uses a noisy version `w` of it and therefore
classifies part of the observations differently.

```{r bacc-fit}
fit_bacc_x <- readRDS("data-for-vignettes/fit_bacc_x.Rds")
fit_bacc_w <- readRDS("data-for-vignettes/fit_bacc_w.Rds")

y <- fit_bacc_x$data$y

m_a <- loo_pred_measure(
  y = y,
  mupred = brms::posterior_epred(fit_bacc_x),
  ylp = brms::log_lik(fit_bacc_x),
  measure = "bacc"
)
m_b <- loo_pred_measure(
  y = y,
  mupred = brms::posterior_epred(fit_bacc_w),
  ylp = brms::log_lik(fit_bacc_w),
  measure = "bacc"
)

comp_bacc <- model_compare(list(m_a = m_a, m_b = m_b))
print(comp_bacc, measures = "all", digits = 3)
```

To compute the standard error we need to know which class each observation
belongs to. This information is stored in the `extra` slot of the
`measure_info` attribute of the model object:

```{r bacc-extra}
str(attr(m_a, "measure_info")$bacc$extra)
```

Undoing the pointwise weighting recovers the 0/1 accuracies, after which the
class-wise variances can be computed as shown in the formula above:

```{r pairwise-bacc}
class_id <- attr(m_a, "measure_info")$bacc$extra$class_id
n_c <- tabulate(class_id)
K   <- length(n_c)

d <- (m_b$pointwise[, "bacc_loo"] - m_a$pointwise[, "bacc_loo"]) *
  (K * n_c[class_id])

var_c <- sapply(seq_len(K), function(k) var(d[class_id == k]) / n_c[k])

c(by_hand  = sqrt(sum(var_c)) / K,
  reported = comp_bacc$bacc_se_diff[comp_bacc$model == "m_b"])
```

## Custom measures

A custom measure is any function returning a list with `estimate`, `se`, and
`pointwise`. It carries two attributes: `measure_name`, its name, and
`measure_loss`, which declares whether lower values are better. It always gets
`diff_method = "custom"`, and the user needs to provide information about how to
compute the standard error of the difference via the `custom_se_fn` argument of
`model_compare()`. `custom_se_fn` accepts one of four things:

| value | meaning |
| :--- | :--- |
| `"mean"` | estimate is the mean of the pointwise values. SE is computed as `sd(d_i) / sqrt(N)` |
| `"sum"` | estimate is the sum of the pointwise values. SE is computed as `sqrt(N) * sd(d_i)` |
| a function | custom function for computing the SE, called as `custom_se_fn(ref, cmp)` |
| `NULL` | only the estimate is computed. SE is reported as `NA` |

### Example (one custom measure)

In the following, we provide a custom measure that is not part of the built-in
measures. The Huber loss is quadratic for small residuals and linear for
large residuals. We define it as follows:

```{r huber-measure}
huber_fn <- function(y, mupred) {
  delta <- 10
  r <- y - colMeans(mupred)
  l <- ifelse(abs(r) <= delta, 0.5 * r^2, delta * (abs(r) - 0.5 * delta))
  list(estimate = mean(l), se = sd(l) / sqrt(length(l)), pointwise = l)
}
attr(huber_fn, "measure_name") <- "huber"
attr(huber_fn, "measure_loss") <- TRUE
```

Furthermore, we provide the name of the measure (i.e., "huber") as an
attribute, and declare that it is a loss rather than a utility (i.e., lower
values are better). Note that a custom measure is by default considered to be a
utility.

::: {.callout .callout-warning}
A custom measure is by default considered to be a utility (i.e., higher values
are better).

This can be changed by setting the `measure_loss` attribute of the
custom-measure function to `TRUE` (see example above).
:::

First, we compute the predictive performance measures for two different models
using `"rmse"` and the custom Huber loss as performance criteria.

```{r custom-measure}
h1 <- fit_measure(1, measure = list("rmse", "huber" = huber_fn))
h3 <- fit_measure(3, measure = list("rmse", "huber" = huber_fn))
```

(Remember that we have created the `fit_measure` helper at the [beginning of
this tutorial](#example). It simply wraps `loo_pred_measure()`.)

Inspecting the attribute `measure_info` shows that the Huber loss
uses `diff_method = "custom"`.

```{r custom-diff-method}
str(attr(h1, "measure_info"))
```

Consequently, we need to provide information to `model_compare()` on how to
compute the standard error of the difference, which is done via the
`custom_se_fn` argument. The Huber estimate *is* the mean of its pointwise
values, so `custom_se_fn = "mean"` is the right declaration.

```{r custom-mean-se}
comp_h <- model_compare(list(m3 = h3, m1 = h1), custom_se_fn = "mean")
print(comp_h, measures = "huber")
```

The same standard error can be spelled out as a function instead.
This is the more general form, and the one to reach for when the measure
needs a custom function for computing the standard error.
It is called as `custom_se_fn(ref, cmp)`, where each argument is a list
describing **one** model with elements `estimate`, `se`, `pointwise`, and
`extra`:

```{r custom-fn-check}
huber_se_fn <- function(ref, cmp) {
  d <- cmp$pointwise - ref$pointwise
  sd(d) / sqrt(length(d))
}

comp_h_fn <- model_compare(list(m3 = h3, m1 = h1), custom_se_fn = huber_se_fn)
print(comp_h_fn, measures = "huber")
```

As discussed above, the difference estimates are always reported on the utility
scale. Therefore `huber_diff` for `m1` is negative.

### Extended example (multiple custom measures)

In the following, we extend the example above by adding one further custom
measure and show how this is implemented. As an additional custom measure we
use RMSE normalized by the standard deviation of the outcome:

```{r extended-example-def}
nrmse_fn <- function(y, mupred) {
  sqe <- (y - colMeans(mupred))^2
  list(
    estimate  = sqrt(mean(sqe)) / sd(y),
    se        = sqrt(var(sqe) / length(sqe)) / (2 * sqrt(mean(sqe))) / sd(y),
    pointwise = sqe,
    extra     = list(sd_y = sd(y))
  )
}
attr(nrmse_fn, "measure_name") <- "nrmse"
attr(nrmse_fn, "measure_loss") <- TRUE
```

Subsequently, we compute the predictive performance measures for two different
models using `"rmse"`, the custom Huber loss, and the custom normalized RMSE
(`"nrmse"`) as performance criteria.

```{r extended-example-fit}
n1 <- fit_measure(1, c("rmse", "huber" = huber_fn, "nrmse" = nrmse_fn))
n3 <- fit_measure(3, c("rmse", "huber" = huber_fn, "nrmse" = nrmse_fn))
```

Additionally, we provide a custom function for computing the standard error
of the difference for the new custom measure:

```{r custom-se-diff-fun}
nrmse_se_fn <- function(ref, cmp) {
  n <- length(ref$pointwise)
  mse_ref <- mean(ref$pointwise)
  mse_cmp <- mean(cmp$pointwise)
  se_ref  <- sqrt(var(ref$pointwise) / n)
  se_cmp  <- sqrt(var(cmp$pointwise) / n)
  cov_rc  <- sum((cmp$pointwise - mse_cmp) * (ref$pointwise - mse_ref)) /
    (n * (n - 1))

  rel_ref <- se_ref / sqrt(mse_ref)
  rel_cmp <- se_cmp / sqrt(mse_cmp)
  rho     <- cov_rc / (se_ref * se_cmp)

  # algebraically the same as the RMSE formula shown earlier, but written so
  # that a model compared against itself cancels to exactly zero
  se_rmse <- 0.5 * sqrt(
    (rel_cmp - rel_ref)^2 + 2 * rel_cmp * rel_ref * max(1 - rho, 0)
  )

  se_rmse / ref$extra$sd_y
}
```

Finally, we perform model comparison with multiple custom measures:

```{r extend-example-fit}
comp_n <- model_compare(
  list(m3 = n3, m1 = n1),
  custom_se_fn = list("huber" = huber_se_fn, "nrmse" = nrmse_se_fn)
)
print(comp_n, measures = "all")
```

## Selecting the reference model

Different measures do not have to agree on which model is best, and in the
[roaches comparison example](#running-example-count-data) they do not.
By default, for each measure the best model is used as reference model and
printed in the first row of the output.

```{r disagree-ranking}
print(
  model_compare(list(m1 = m1, m2 = m2, m3 = m3)),
  measures = "all"
)
```

We see that for `elpd` and `mae` the third model (`m3`) is best, while for
the other measures the second model (`m2`) is best.

We can change the reference model using the `rank_by` argument of
`model_compare()`. For example, if we want `m3` to be the reference model for
every measure, we use `rank_by = "m3"`.

```{r diff-ref-model}
print(
  model_compare(list(m1 = m1, m2 = m2, m3 = m3), rank_by = "m3"),
  measures = "all"
)
```

Note that the ordering remains the same: the estimates are displayed in
decreasing order (best model first).

It is also possible to select as reference model the best model according to a
specific measure. For example, suppose we want to use for all measures the same
reference model as the one used for `r2`:

```{r diff-ref-measure}
print(
  model_compare(list(m1 = m1, m2 = m2, m3 = m3), rank_by = "r2"),
  measures = "all"
)
```

## Summary

```{r summary-table, echo=FALSE}
knitr::kable(
  data.frame(
    Measure = c("elpd, ic", "mlpd, mae, mse, acc, brier, rps, srps",
                "rmse", "r2", "bacc", "custom"),
    `diff_method` = c("sum", "mean", "measure_specific",
                      "measure_specific", "measure_specific",
                      "custom"),
    `Standard error` = c(
      "sqrt(N) * sd(d_i)",
      "sd(d_i) / sqrt(N)",
      "delta method on the MSE scale",
      "delta method, including baseline uncertainty",
      "stratified by class, paired within class",
      "from custom_se_fn: your function, \"sum\", \"mean\", or NA"
    ),
    check.names = FALSE
  )
)
```

The things worth remembering:

1. The difference is always `estimate_cmp - estimate_ref`, reported so that
   negative means worse.
2. The standard error is always paired, because the models are evaluated on the
   same observations.
3. `NA` in a standard error column means "not available", not "zero" and not
   "an error occurred".
4. A custom measure is assumed to be a utility. If yours is a loss, say so with
   `attr(my_fun, "measure_loss") <- TRUE`, or its differences and its `rank_by`
   ordering will come out reversed.

## References

Chen, F., Xue, Y., Tan, M. T., and Chen, P. (2015). Efficient statistical tests
to compare Youden index: accounting for contingency correlation. *Statistics in
Medicine*, 34(9):1560–1576. doi:10.1002/sim.6432

Gelman, A., Goodrich, B., Gabry, J., and Vehtari, A. (2019). R-squared for
Bayesian regression models. *The American Statistician*, 73(3):307–309.
doi:10.1080/00031305.2018.1549100

Gelman, A., and Hill, J. (2007). *Data Analysis Using Regression and
Multilevel/Hierarchical Models*. Cambridge University Press.

McNemar, Q. (1947). Note on the sampling error of the difference between
correlated proportions or percentages. *Psychometrika*, 12(2):153–157.
doi:10.1007/BF02295996

Newcombe, R. G. (2001). Simultaneous comparison of sensitivity and specificity
of two tests in the paired design: a straightforward graphical approach.
*Statistics in Medicine*, 20(6):907–915. doi:10.1002/sim.906

Sivula, T., Magnusson, M., Matamoros, A. A., and Vehtari, A. (2025).
Uncertainty in Bayesian leave-one-out cross-validation based model comparison.
*Bayesian Analysis*. doi:10.1214/25-BA1569

Vehtari, A., Gelman, A., and Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics and
Computing*, 27(5):1413–1432. doi:10.1007/s11222-016-9696-4

Ver Hoef, J. M. (2012). Who invented the delta method? *The American
Statistician*, 66(2):124–127. doi:10.1080/00031305.2012.687494

## See also

- [Computing predictive performance measures](pred-measure-workflow.html) — how
  to produce the objects compared here.
- [Overview of scores and metrics](overview-measures.html) — definitions and
  formulas for each measure.
- `?"loo-glossary"` — the `measure_info` attribute in full.
