---
title: "Differences and their standard errors in model comparison"
author: "Florence Bockting"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: TRUE #!r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{comparison-standard-errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="../children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = params$EVAL,
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
devtools::load_all(".")
```

## Introduction

When you call `loo_compare()` on results from `loo_pred_measure()`, every
measure gets two columns: a difference and its standard error.
For example, `loo_compare` based on `loo_pred_measure` with measures `elpd`,
`mse`, and `rmse`, will produce:

```
model  elpd_diff  se_diff   mse_diff  mse_se_diff   rmse_diff  rmse_se_diff
```

The difference (`elpd_diff`, `mse_diff`, `rmse_diff`) tells you *which* model
did better and the standard error provides an estimate in how far you can trust
this estimate. In this vignette we explain:

1. The approaches for computing the point estimate and standard error of the
   difference for the different measures.
2. How to provide a formula for computing the standard error of a difference for
   a custom measure.

## A simple example

In the following, we want to compare three Poisson regression models on the same data set.
For this we use the `roaches` data from **rstanarm**. It records a pest-management trial in
262 apartment buildings. The outcome `y` is the number of roaches caught in
traps after treatment and ranges from 0 to 357. The predictors are:

| Predictor | Description |
| :-- | :---- |
| `roach1` | pre-treatment roach count (used as `sqrt_roach1 = sqrt(roach1)`) |
| `treatment` | whether the building received pest management |
| `senior` | whether the building is restricted to elderly residents |
| `exposure2` | number of days the traps were exposed |

The three models that we will use in the following comparison tasks are Poisson
regressions with different predictors:

+ `m1`: `y ~ treatment + senior`
+ `m2`: `y ~ sqrt_roach1 + senior`
+ `m3`: `y ~ sqrt_roach1 + treatment`

For each model we compute its predictive performance via `loo_pred_measure()` with
measures `elpd`, `mae`, `mse`, `rmse`, and `r2`.

```{r load-data}
res <- readRDS("../../tests/testthat/data-for-tests/test_data_roaches_compare.Rds")

fit_measure <- function(i, measure) {
  loo_pred_measure(
    y       = res$y,
    mupred  = res[[paste0("mupred_m", i)]],
    ylp     = res[[paste0("ylp_m", i)]],
    measure = measure
  )
}

m1 <- fit_measure(1, c("mae", "mse", "rmse", "r2"))
m2 <- fit_measure(2, c("mae", "mse", "rmse", "r2"))
m3 <- fit_measure(3, c("mae", "mse", "rmse", "r2"))

comp <- loo_compare(list(m1 = m1, m2 = m2, m3 = m3))
print(comp, measures = "all")
```

## A closer look at the model comparison output

### Estimated difference between two models
The estimates `elpd_diff`, `mae_diff`, `rmse_diff`, and `r2_diff` 
represent the differences of the estimates of the two models which enter the 
comparison (e.g., `m1$estimates["elpd_loo", "Estimate"] - m3$estimates["elpd_loo", "Estimate"]`). 
Whereby the first model in this difference is the comparison model and the
second model is the reference model.

```{r point-estimate}
est <- function(x, measure) x$estimates[measure, "Estimate"]

data.frame(
  measure  = c(
    "elpd_diff", "mse_diff", "mae_diff", "rmse_diff", "r2_diff"
  ),
  by_hand  = c(
    est(m1, "elpd_loo") - est(m3, "elpd_loo"),
    est(m1, "mse_loo")  - est(m3, "mse_loo"),
    est(m1, "mae_loo")  - est(m3, "mae_loo"),
    est(m1, "rmse_loo") - est(m3, "rmse_loo"),
    est(m1, "r2_loo")   - est(m3, "r2_loo")
  ),
  reported = c(
    comp$elpd_diff[comp$model == "m1"],
    comp$mse_diff[comp$model == "m1"],
    comp$mae_diff[comp$model == "m1"],
    comp$rmse_diff[comp$model == "m1"],
    comp$r2_diff[comp$model == "m1"]
  )
)
```

Note that the reported `mae_diff`, `mse_diff`, and `rmse_diff` rows differ in 
sign from the hand-computed value, and this is deliberate. `mae`, `mse`, and `rmse`
are **losses**, while `elpd` and `r2` are **utilities**. For losses, the stimated value
is interpreted as lower is better, while for utilities, higher is better.

However, `loo_compare()` reports every difference on a **utility scale** 
(higher is better). Measures that are on a loss scale are converted to a utility scale.
In consequnence, a negative number always means "this model is 
worse than the reference model", whatever the measure. 

`loo_compare()` informs the user when a measure is converted to a utility scale
for the model comparison task:

```{r sign-message, warning=TRUE, message=TRUE}
comp <- loo_compare(list(m1 = m1, m2 = m2, m3 = m3))
```

### Standard error of the difference between two models

Depending on the measure there are different approaches to compute 
the standard error (SE) of the difference. Which formula a measure uses 
is recorded on the result object, in the `measure_compare_meta` 
attribute (see `?"loo-glossary"`):

```{r diff-method}
meta <- attr(m1, "measure_compare_meta")
sapply(meta, function(x) x$diff_method)
```

There are four possible values:

| `diff_method` | SE of the difference | Measures |
| :--- | :--- | :--- |
| `"sum"` | `sqrt(N) * sd(d_i)` | `elpd`, `ic`, `mlpd` |
| `"mean"` | `sd(d_i) / sqrt(N)` | `mae`, `mse`, `acc`, `brier`, `rps`, `srps` |
| `"pairwise"` | the measure's own `se_diff_fun` | `rmse`, `r2`, `bacc` |
| `"auto"` | inferred at comparison time | custom measures |

whereby `N` is the number of observations and `d_i = pointwise_cmp - pointwise_ref`
is the paired difference of the pointwise estimates of the two models 
(e.g., `m1$pointwise[, "elpd_loo"]` and `m3$pointwise[, "elpd_loo"]`). 

A fifth value, `"estimates_only"`, means no standard error is available. In 
this case the estimate for the difference is still reported but the standard 
error gets `NA`. This can only happen when `"auto"` cannot work out how a
custom measure's pointwise values relate to its estimate (see here[TODO] for more
details on custom measures).

#### Method: "sum"

For `elpd`, `ic`, and `mlpd` the estimate is a sum of pointwise log predictive
densities, so the standard error of a difference scales up with `sqrt(N)`:

```{r sum-method}
n <- attr(m1, "dims")[2]
d_elpd <- m1$pointwise[, "elpd_loo"] - m3$pointwise[, "elpd_loo"]

c(by_hand     = sqrt(n) * sd(d_elpd),
  reported = comp$se_diff[comp$model == "m1"])
```

#### Method: "mean"

For `mae`, `mse`, and most scoring rules, the estimate is a mean of pointwise
squared errors, so the same standard deviation is divided by `sqrt(N)` instead:

```{r mean-method}
d_mse <- m1$pointwise[, "mse_loo"] - m3$pointwise[, "mse_loo"]

c(hand     = sd(d_mse) / sqrt(n),
  reported = comp$mse_se_diff[comp$model == "m1"])
```

#### Method: "pairwise"

**RMSE**

For RMSE, the delta method [REF-TODO] on the MSE scale is used 
and propagated through the square root:

$$
SE_{\Delta RMSE} = \frac{1}{2}\sqrt{
  \frac{SE_{MSE(a)}^2}{MSE(a)} + \frac{SE_{MSE(b)}^2}{MSE(b)}
  - \frac{2\,\mathrm{Cov}[MSE(a), MSE(b)]}{RMSE(a)\,RMSE(b)}
}
$$

Everything on the right hand side is computable from the stored pointwise
squared errors:

```{r pairwise-rmse}
sqe_ref <- m3$pointwise[, "rmse_loo"]
sqe_cmp <- m1$pointwise[, "rmse_loo"]

mse_ref <- mean(sqe_ref)
mse_cmp <- mean(sqe_cmp)
se_ref  <- sqrt(var(sqe_ref) / n)
se_cmp  <- sqrt(var(sqe_cmp) / n)
cov_rc  <- sum((sqe_cmp - mse_cmp) * (sqe_ref - mse_ref)) / (n * (n - 1))

se_rmse <- 0.5 * sqrt(
  se_cmp^2 / mse_cmp + se_ref^2 / mse_ref -
    2 * cov_rc / (sqrt(mse_cmp) * sqrt(mse_ref))
)

c(by_hand  = se_rmse,
  reported = comp$rmse_se_diff[comp$model == "m1"])
```

**R2**

For R2, the delta method with one more term is applied, because its 
denominator `MSE(y)` is estimated from the same data and carries its own
uncertainty. It needs the pointwise baseline `(y_i - mean(y))^2`, which
`loo_compare()` no longer has access to by the time it runs, so 
`measure_r2()` stores it on the object:

```{r r2-extra}
str(attr(m1, "measure_compare_meta")$r2$extra)
```

```{r pairwise-r2}
mse_y_i <- attr(m1, "measure_compare_meta")$r2$extra$mse_y_i
mse_y   <- mean(mse_y_i)

d_sqe  <- m1$pointwise[, "r2_loo"] - m3$pointwise[, "r2_loo"]
scaled <- d_sqe - (mean(d_sqe) / mse_y) * mse_y_i

c(by_hand  = sd(scaled) / sqrt(n) / mse_y,
  reported = comp$r2_se_diff[comp$model == "m1"])
```

**bacc** 
The balanced accuracy (bacc) is the average of the per-class accuracies, which 
gives each class equal weight no matter how rare it is. That makes it the 
standard choice for imbalanced outcomes.

To explain the standard error of the difference for bacc, we need to introduce
another example where the outcome variable is binary. Therfore, we simulate 200 
observations from a logistic model with a single continuous predictor `x`. 

```{r bacc-data}
set.seed(2024)
n_obs   <- 200
n_draws <- 400

x <- rnorm(n_obs)
y <- rbinom(n_obs, 1, plogis(-1.4 + 1.6 * x))
table(y)
```

TODO from here

## Custom measures

A custom measure is any function returning a list with `estimate`, `se`, and
`pointwise`. By default it gets `diff_method = "auto"`, and `loo_compare()`
works out what to do by checking how the estimate relates to the pointwise
values.

### When autodetection works

If your estimate really is the mean of your pointwise values, autodetection
finds that and you get a standard error for free. Here is a Huber-type loss:

```{r custom-auto}
huber <- function(y, mupred) {
  delta <- 10
  r <- y - colMeans(mupred)
  l <- ifelse(abs(r) <= delta, 0.5 * r^2, delta * (abs(r) - 0.5 * delta))
  list(estimate = mean(l), se = sd(l) / sqrt(length(l)), pointwise = l)
}
attr(huber, "measure_name") <- "huber"

h1 <- fit_measure(1, huber)
h3 <- fit_measure(3, huber)

attr(h1, "measure_compare_meta")$huber$diff_method

comp_h <- loo_compare(list(m3 = h3, m1 = h1))
comp_h[, c("model", "huber_diff", "huber_se_diff")]
```

```{r custom-auto-check}
d_huber <- h1$pointwise[, "huber_loo"] - h3$pointwise[, "huber_loo"]

c(hand     = sd(d_huber) / sqrt(n),
  reported = comp_h$huber_se_diff[comp_h$model == "m1"])
```

Two things to note. First, autodetection confirmed that `mean(pointwise)`
reproduces the estimate, and used the `"mean"` formula. Second, the reported
difference is **positive** even though a larger Huber loss is worse. Custom
measures are always treated as utilities — higher is better — because the
package has no way to know otherwise. If your measure is a loss, either return
its negative, or read the difference column with that in mind.

### When autodetection cannot work

Now a measure whose estimate is *not* a sum or mean of its pointwise values —
RMSE normalised by the standard deviation of the outcome:

```{r custom-estimates-only}
nrmse <- function(y, mupred) {
  sqe <- (y - colMeans(mupred))^2
  list(
    estimate  = sqrt(mean(sqe)) / sd(y),
    se        = sqrt(var(sqe) / length(sqe)) / (2 * sqrt(mean(sqe))) / sd(y),
    pointwise = sqe,
    extra     = list(sd_y = sd(y))
  )
}
attr(nrmse, "measure_name") <- "nrmse"

n1 <- fit_measure(1, nrmse)
n3 <- fit_measure(3, nrmse)

attr(n1, "measure_compare_meta")$nrmse$diff_method

comp_n <- loo_compare(list(m3 = n3, m1 = n1))
comp_n[, c("model", "nrmse_diff", "nrmse_se_diff")]
```

The stored `diff_method` is still `"auto"`, because `"auto"` is not a formula —
it is an instruction to work one out at comparison time. Here that attempt
failed: neither `sum(pointwise)` nor `mean(pointwise)` reproduces the estimate,
so `loo_compare()` resolved it to `"estimates_only"` and reported the difference
with an `NA` standard error.

This is not an error. It is the package declining to print a number it cannot
justify, which is the correct behaviour when the alternative is a confident
wrong one.

It is worth being aware of what autodetection can and cannot check. It verifies
that the estimate equals the sum or the mean of the pointwise values, which is
an *algebraic* fact about the numbers in front of it. The standard error it then
applies needs something stronger: that those pointwise values behave as `N`
exchangeable contributions whose sample standard deviation describes how much
the estimate would vary. Balanced accuracy is a measure where the first holds
and the second does not — its pointwise values do sum to its estimate, but
using the `"sum"` formula would throw away the class stratification, which is
why it registers its own function instead. If your measure has structure of that
kind, supply an `se_diff_fun` rather than relying on autodetection.

### Supplying your own `se_diff_fun`

To get a real standard error, attach a function to the measure. It is called as
`se_diff_fun(ref, cmp)`, where each argument is a list describing **one** model
with elements `estimate`, `se`, `pointwise`, and `extra`, and it returns a
single number.

Normalised RMSE is RMSE divided by a quantity that does not depend on the
model, so its delta-method standard error is the RMSE one divided by `sd(y)` —
which is exactly what the measure stashed in `extra` for this purpose:

```{r custom-se-diff-fun}
attr(nrmse, "se_diff_fun") <- function(ref, cmp) {
  n <- length(ref$pointwise)
  mse_ref <- mean(ref$pointwise)
  mse_cmp <- mean(cmp$pointwise)
  se_ref  <- sqrt(var(ref$pointwise) / n)
  se_cmp  <- sqrt(var(cmp$pointwise) / n)
  cov_rc  <- sum((cmp$pointwise - mse_cmp) * (ref$pointwise - mse_ref)) /
    (n * (n - 1))

  rel_ref <- se_ref / sqrt(mse_ref)
  rel_cmp <- se_cmp / sqrt(mse_cmp)
  rho     <- cov_rc / (se_ref * se_cmp)

  # algebraically the same as the RMSE formula shown earlier, but written so
  # that a model compared against itself cancels to exactly zero
  se_rmse <- 0.5 * sqrt(
    (rel_cmp - rel_ref)^2 + 2 * rel_cmp * rel_ref * max(1 - rho, 0)
  )

  se_rmse / ref$extra$sd_y
}

n1 <- fit_measure(1, nrmse)
n3 <- fit_measure(3, nrmse)

attr(n1, "measure_compare_meta")$nrmse$diff_method

comp_n <- loo_compare(list(m3 = n3, m1 = n1))
comp_n[, c("model", "nrmse_diff", "nrmse_se_diff")]
```

Attaching the function switched `diff_method` from `"auto"` to `"pairwise"`,
and the standard error column is populated.

Three practical points:

- **Write the formula so that identical inputs cancel exactly.** The literal
  three-term version of the RMSE delta method can produce a tiny negative
  number under the square root when both models have the same squared errors,
  which is the first row of every comparison table. The rearranged form above
  returns exactly `0` there.
- **Your function always sees the measure's natural scale.** If you set an
  orientation that flips the stored values, the package undoes it before
  calling you and reapplies it to the difference afterwards. You never have to
  think about sign conventions.
- **`extra` is how you get data to your function.** Anything your measure
  returns as `extra` is stored on the object and handed back to
  `se_diff_fun()`. This is the only way to reach data, like `y`, that is no
  longer available when `loo_compare()` runs.

One caveat about storage: a custom `se_diff_fun` is stored on the result object
as a closure, so `saveRDS()` will serialise its enclosing environment. If you
define it inline in a script where a large draws matrix is in scope, that matrix
goes into the file. Define such functions in a clean environment if object size
matters. Built-in measures avoid this by storing a name instead of a function.

## When measures disagree

Everything so far has been about computing the two columns. This section is
about what they are for.

Different measures do not have to agree on which model is best, and in the
roaches comparison they do not. Ranking the three models by each measure
separately:

```{r disagree-ranking}
estimates <- sapply(
  list(m1 = m1, m2 = m2, m3 = m3),
  function(x) x$estimates[, "Estimate"]
)
round(estimates, 3)
```

```{r disagree-best}
# elpd and r2 are utilities (higher is better); mae, mse and rmse are losses
higher_is_better <- c(elpd_loo = TRUE, p_loo = NA, mae_loo = FALSE,
                      mse_loo = FALSE, rmse_loo = FALSE, r2_loo = TRUE)

ranking <- function(measure) {
  v <- sort(estimates[measure, ], decreasing = higher_is_better[measure])
  paste(names(v), collapse = " > ")
}

measures <- c("elpd_loo", "mae_loo", "mse_loo", "rmse_loo", "r2_loo")
data.frame(measure = measures, ranking = sapply(measures, ranking),
           row.names = NULL)
```

Two different answers. `elpd` and `mae` put **m3** first; `mse`, `rmse`, and
`r2` put **m2** first. A reader who ran only one of these would come away with a
different "best model".

This is not a malfunction, and the pattern is a common one:

- `mse`, `rmse`, and `r2` agree with each other because they are all monotone
  functions of the same pointwise squared errors — they cannot disagree, and
  reporting all three is reporting one thing three times.
- Squared error is dominated by the largest residuals. `mae` weights every
  observation's error linearly and so is far less sensitive to a handful of
  badly predicted buildings, and `elpd` scores the whole predictive
  distribution rather than a point summary. When those two part company from
  the squared-error family, it usually means a few extreme observations are
  driving the ranking.

So which model is best? The standard error column answers that, and the answer
is that the question is not settled by this data:

```{r disagree-se}
row_m2 <- comp[comp$model == "m2", ]

data.frame(
  measure = c("elpd", "mae", "mse", "rmse", "r2"),
  diff    = c(row_m2$elpd_diff, row_m2$mae_diff, row_m2$mse_diff,
              row_m2$rmse_diff, row_m2$r2_diff),
  se_diff = c(row_m2$se_diff, row_m2$mae_se_diff, row_m2$mse_se_diff,
              row_m2$rmse_se_diff, row_m2$r2_se_diff)
) |>
  transform(ratio = round(abs(diff) / se_diff, 2))
```

Every one of these differences is **smaller than its own standard error**. The
`elpd` difference is about one standard error; the rest are a fifth of one. Both
orderings are noise. The measures are not really contradicting each other — they
are all saying that `m2` and `m3` are indistinguishable on 262 observations, and
the apparent ranking flip is which side of zero the noise happened to land on.

Note also that all three squared-error measures produce nearly the same ratio.
For `mse` and `rmse` that is exact up to the delta-method approximation: a
first-order expansion scales a difference and its standard error by the same
derivative, so the ratio between them survives the square root. `r2` is very
slightly different (`0.199` against `0.202` at full precision) because its
standard error also carries the uncertainty in the estimated baseline `MSE(y)`,
which is not simply a transformation of `MSE`.

`m1` is the one genuine finding here — it drops `sqrt_roach1`, and every measure
puts it last by a margin large relative to its standard error. That is a
conclusion the data supports; "m2 beats m3 on RMSE" is not.

The practical advice:

1. **Choose your measure before you look**, based on what you want the model to
   do well. Picking the measure that favours your preferred model after the fact
   is a way of fooling yourself, and with several measures on the table it is an
   easy mistake to make.
2. **Read the difference against its own standard error**, not against zero. A
   ranking is only meaningful if the differences that produce it are.
3. **Do not read `mse`, `rmse`, and `r2` as three independent votes.** They are
   one measure in three costumes.

## Summary

```{r summary-table, echo=FALSE}
knitr::kable(
  data.frame(
    Measure = c("elpd, ic, mlpd", "mae, mse, acc, brier, rps, srps",
                "rmse", "r2", "bacc", "custom (default)",
                "custom (with se_diff_fun)"),
    `diff_method` = c("sum", "mean", "pairwise", "pairwise", "pairwise",
                      "auto", "pairwise"),
    `Standard error` = c(
      "sqrt(N) * sd(d_i)",
      "sd(d_i) / sqrt(N)",
      "delta method on the MSE scale",
      "delta method, including baseline uncertainty",
      "stratified by class, paired within class",
      "detected as sum or mean, otherwise NA",
      "whatever your function returns"
    ),
    check.names = FALSE
  )
)
```

The things worth remembering:

1. The difference is always `estimate_cmp - estimate_ref`, reported so that
   negative means worse.
2. The standard error is always paired, because the models are evaluated on the
   same observations.
3. `NA` in a standard error column means "not available", not "zero" and not
   "an error occurred".
4. A custom measure gets a standard error for free if it is a mean or a sum of
   its pointwise values, and needs an `se_diff_fun` otherwise.

## References

Newcombe, R. G. (2001). Simultaneous comparison of sensitivity and specificity
of two tests in the paired design: a straightforward graphical approach.
*Statistics in Medicine*, 20(6):907–915.

Chen, F., Xue, Y., Tan, M. T., and Chen, P. (2015). Efficient statistical tests
to compare Youden index: accounting for contingency correlation. *Statistics in
Medicine*, 34(9):1560–1576.

Vehtari, A., Gelman, A., and Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics and
Computing*, 27(5):1413–1432.

## See also

- [Computing predictive performance measures](pred-measure-workflow.html) — how
  to produce the objects compared here.
- [Overview of scores and metrics](overview-measures.html) — definitions and
  formulas for each measure.
- `?"loo-glossary"` — the `measure_compare_meta` attribute in full.
