---
title: "Model comparison: Using the `model_compare()` function."
author: "Florence Bockting"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    css:
      - !expr system.file("rmarkdown/templates/html_vignette/resources/vignette.css", package = "rmarkdown")
      - styles.css
    toc: true
    toc_depth: 3
params:
  EVAL: TRUE #!r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Model comparison: Using the `model_compare()` function.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="../children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = TRUE,
  warning = TRUE,
  eval = params$EVAL,
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)

# brms::kfold() exports the fits to future.apply, which caps globals at 500 MiB
options(future.globals.maxSize = 4 * 1024^3)
```

```{r, child="../children/LOAD-BRMS-GITHUB.txt", eval=params$EVAL}
```

## Introduction
This vignette introduces the `loo::model_compare()` function.
Use `model_compare()` instead of the deprecated `loo::loo_compare()` (see the
note below).

This vignette shows how to use the function. It does not explain how to
interpret the results. It also does not explain what to do in the Bayesian
workflow when you get diagnostic warnings, for example high Pareto-k values.
For these topics, see the "TODO-case study-model_compare vignette" [Link-TODO].

This vignette does not derive the formulas behind `<measure>_diff` and
`<measure>_se_diff` (introduced below). For the derivations, see the supplement
(TODO-Vehtari et al., 2026).

`model_compare()` accepts the output of these functions:

+ `loo()`, `kfold()`, and `waic()`
+ `<type>_pred_measure()`, where `<type>` is `loo`, `kfold`, `test`, or `insample`

`model_compare()` is new in loo v3.0.0. It compares models on several
predictive performance measures, for example `elpd`, `mae`, `rmse`, and `acc`.
For the full list, see the [Overview of scores and metrics](overview-measures.html)
vignette. You can also define custom measures.

This vignette uses the output of `<type>_pred_measure()`. It also briefly
shows how it compares to the deprecated `loo_compare()` output.
The vignette explains:

1. The arguments and output of `model_compare()`.
1. Which standard error `model_compare()` uses for each built-in measure.
1. How to compute the difference and its standard error for a **custom**
   measure.

::: {.callout .callout-note}
**Relation between `model_compare` and `loo_compare`**

The `loo_compare` function has traditionally been used for model comparison in
loo, accepting any object that inherits from the `loo` class (such as those
produced by `loo()`, `kfold()`, and `waic()`).

With the introduction of the *`pred_measure` family*, we have deprecated
`loo_compare` and introduced `model_compare`. While `model_compare` retains full
backward compatibility, it also expands functionality to handle outputs from
the `pred_measure` family.

`loo_compare` still works and is still a generic, so methods registered by other
packages keep dispatching, but it now warns and is frozen at its previous
behavior: `loo` objects compared on ELPD only. The `pred_measure` sources and
custom measures described in this article require `model_compare`.
`loo_compare` will be removed in a future release. For more details, see
`vignette("migration-guide", package = "loo")`.
:::

## Running example (count data)

To illustrate the functionality of `model_compare()` we first introduce a
*running example*. In this example, we compare three Poisson regression
models fitted to the **`roaches` data** from `rstanarm` (Gelman and Hill, 2007).
This data set records a pest-management trial in 262 apartment buildings.
The outcome `y` is the number of roaches caught in traps after treatment and
ranges from 0 to 357. The predictors are:

+ `roach1`: pre-treatment roach count (used as `sqrt_roach1 = sqrt(roach1)`)
+ `treatment`: whether the building received pest management
+ `senior`: whether the building is restricted to elderly residents

The **three models** used in the following comparison tasks are defined as
follows:

```
  m1:      y ~ treatment + senior
  m2:      y ~ sqrt_roach1 + senior
  m3:      y ~ sqrt_roach1 + treatment
```

As **performance criteria** we select six predictive performance measures:
`elpd`, `mae`, `mse`, `rmse`, `r2`, and `rps` computed by `loo_pred_measure`.

Note that `rps` is the only one of these that needs the posterior *predictive*
draws `ypred`; the error measures are computed from the posterior expectation
`mupred` (see the input requirements table in `?loo_pred_measure`).

::: {.callout .callout-note}
**Running example**

The running example shows how to call `model_compare()`. It does not show
how to interpret the results. For that, see [TODO-LINK]. We therefore do not
discuss diagnostic flags or interpretation of the output here.

:::

```{r load-libraries}
library(loo)
```

```{r setup-example}
measures <- c("elpd", "mae", "mse", "rmse", "r2", "rps")
```

## The classical `loo_compare` output

As the actual model building workflow is not the focus of this tutorial, we fit
the three models here without further comment. `add_criterion()` attaches a
`loo` object (as returned by `loo::loo()`) to each fit.

```{r fit-models, results="hide", warning=FALSE, message=FALSE}
SEED <- 42

data(roaches, package = "rstanarm")
roaches$sqrt_roach1 <- sqrt(roaches$roach1)

fit_m1 <- brms::brm(
  y ~ treatment + senior,
  data = roaches,
  family = poisson,
  prior = brms::prior(normal(0, 1), class = b),
  chains = 2, iter = 400, seed = SEED, refresh = 0
)
fit_m2 <- update(fit_m1, formula = y ~ sqrt_roach1 + senior,
                 newdata = roaches, refresh = 0)
fit_m3 <- update(fit_m1, formula = y ~ sqrt_roach1 + treatment,
                 newdata = roaches, refresh = 0)

add_loo <- function(fit) {
  brms::add_criterion(fit, criterion = "loo", moment_match = TRUE,
                      save_psis = TRUE)
}
fit_m1 <- add_loo(fit_m1)
fit_m2 <- add_loo(fit_m2)
fit_m3 <- add_loo(fit_m3)

loos <- list(m1 = fit_m1$criteria$loo, m2 = fit_m2$criteria$loo,
             m3 = fit_m3$criteria$loo)
```

Passing these `loo` objects to the classical `loo_compare()` reproduces the
familiar ELPD-only comparison, and warns once per session that `model_compare()`
is now the preferred function:

```{r loo_compare}
loo_compare(loos)
```

When we use `model_compare` instead, we get the same output for `loo` objects:

```{r model_compare_wloo}
model_compare(loos)
```

It also supports the `simplify` argument in the print method:

```{r model_compare_wloo_simplify}
print(model_compare(loos), simplify = FALSE)
```

## Preview: Glimpse into the model comparison results

We now use `loo_pred_measure()` to compute the predictive performance for
each model (`m1, m2, m3`) and measure (see list above).
Finally, we compare the performance of the three models using `model_compare()`.
By default, only the first measure in the provided vector (here `measures`) is
shown. In this case the first measure is `elpd`.

<a id="example"></a>
```{r pred-measure}
fit_measure <- function(fit, measure) {
  loo_pred_measure(
    loo = fit$criteria$loo,
    y = fit$data$y,
    ypred = brms::posterior_predict(fit),
    mupred = brms::posterior_epred(fit),
    measure = measure
  )
}

set.seed(SEED)
m1 <- fit_measure(fit_m1, measures)
m2 <- fit_measure(fit_m2, measures)
m3 <- fit_measure(fit_m3, measures)

(comp <- model_compare(list(m1 = m1, m2 = m2, m3 = m3)))
```

When we want to see a **specific measure**, we can change this by passing the 
information to the `print` method:

```{r pred-measure-r2}
print(comp, measures = "r2")
```

If you want to see **all columns related to a measure** specified in
`measures` use `simplify = FALSE`:

```{r pred-measure-r2-simplify}
print(comp, measures = "r2", simplify = FALSE)
```

If you want to see **all measures** use `measures = "all"`:

```{r pred-measure-all}
print(comp, measures = "all")
```

### Output of `model_compare()`

Using `print(., measures = "all")` returns a subsection for each measure.
The *rows* represent the different models, with the best model always in the
first row, followed by the second best, and so on.
The *columns*:

  + `model`: Name of each model
  + `<measure>_diff`: The point estimate representing the difference in the measure between the current model and the reference model (calculated as model - reference). For the reference model itself, this value is always zero.
  + `<measure>_se_diff`$^{(1)}$: The standard error of the difference
  + `p_worse`: (only for `elpd`) Estimated probability that a model has worse predictive performance than the reference model
  + `diag_diff`: (only for `elpd`) Diagnostic column flagging miscalibration of the normal approximation behind `p_worse` and `se_diff`

Above the measure subsections, `print()` names the reference model each set of
differences was computed against, and, for `loo_pred_measure()` inputs, reports
the PSIS-LOO diagnostic once per model:

  + `bad_k`: Number of observations whose Pareto $\hat{k}$ exceeds the PSIS
    reliability threshold. Pareto $\hat{k}$ describes a model's PSIS-LOO
    approximation.

This block is omitted when no model is flagged, and for `kfold`, `test`, and
`insample` sources, which have no Pareto $\hat{k}$ diagnostic.

$^{(1)}$ The only measure that deviates from this naming convention is `elpd`,
where the column is called `se_diff` instead of `elpd_se_diff` for backwards
compatibility.

### The `measure_info` attribute

Each model has a `measure_info` attribute, which provides meta information for
each performance measure.

```{r meta-attribute, class.output="scroll-output"}
meta <- attr(m1, "measure_info")

# measure info for each single measure (e.g., R2)
names(meta$r2)
```

The information is stored in different *slots*:

| slot | description | functionality |
| :--- | :---- | :----- |
| `loss` | Whether the measure is a *loss* (`TRUE`; lower values of the measure are better) or a *utility* (`FALSE`, higher values are better) | During model comparison, loss measures have their sign flipped so that all differences are reported on a utility scale (higher is better) |
| `diff_method` | How paired pointwise differences are aggregated within model comparison: `"sum"`, `"mean"`, `"measure_specific"`, or `"custom"` | Selects the formula for `<measure>_se_diff` (see the table in Section [Computation of the standard error](#computation-of-the-standard-error-measure_se_diff)) |
| `se_diff_fun` | Name of the measure's own standard-error function for computing `<measure>_se_diff`, only set when `diff_method = "measure_specific"` (`"rmse"`, `"r2"`, `"bacc"`) | Points to the built-in function that computes `<measure>_se_diff` for measures whose estimate is not a plain sum or mean of pointwise values |
| `extra` | Optional auxiliary data the measure stored for its `se_diff_fun`, on the measure's natural scale (e.g., `mse_y_i` for `r2`, `class_id` for `bacc`); `NULL` for measures that need none | Carries the quantities the standard error needs but the pointwise values do not contain; passed to `se_diff_fun()` as part of each model's input |

## How `model_compare()` computes the output

This section relates `<measure>_diff` and `<measure>_se_diff` to the model
objects. All measures use the same point estimate. The standard error depends
on the measure. For the formulas, see the supplement
(TODO-Vehtari et al., 2026, Section TODO-ARXIV-SECTION).

### Computation of the point estimate `<measure>_diff`
The estimates `elpd_diff`, `mae_diff`, `rmse_diff`, and `r2_diff` in the
output of the running example represent the point estimates of the pairwise
model comparisons.
They are computed as the measure-specific difference between two model
estimates (comparison model - reference model).

For example, the difference in `elpd` between model `m1` (comparison) and model
`m3` (reference) is computed as follows:

```{r, eval=FALSE}
m1_elpd <- m1$estimates["elpd_loo", "Estimate"]
m3_elpd <- m3$estimates["elpd_loo", "Estimate"]

m1_elpd - m3_elpd
```

This computation is done for each measure and each pairwise model comparison.

By default, however, the reference model is chosen *per measure* as the best
model for that measure.
The reference used for each measure is recorded in the `compare_reference`
attribute:

```{r reference-models}
models <- list(m1 = m1, m2 = m2, m3 = m3)
ref <- attr(comp, "compare_reference")
ref
```

The following code block shows, for model `m1`, the hand-computed differences
against each measure's own reference model and the differences reported in the
`comp` object:

```{r point-estimate}
est <- function(x, measure) x$estimates[measure, "Estimate"]
measures_all <- c("elpd", "mse", "mae", "rmse", "r2")

data.frame(
  reference = ref[measures_all],
  by_hand   = sapply(measures_all, function(m) {
    est(m1, paste0(m, "_loo")) - est(models[[ref[[m]]]], paste0(m, "_loo"))
  }),
  reported  = sapply(measures_all, function(m) {
    comp[[paste0(m, "_diff")]][comp$model == "m1"]
  })
)
```

Note that in the above output, the sign for some measures is flipped between
the `by_hand` and `reported` column, specifically for `mse`, `mae`, and `rmse`.
These measures are *losses* (see callout box below).

However, when we compare all measures together, we want them all to be
on the same scale, which in our case is the **utility scale**. Therefore, losses
are converted to the utility scale. As such, higher values are always better.
Consequently, the best model has the highest value, and when the best model is
also the reference model, the estimated difference `<measure>_diff`
(computed as comp - reference) is always non-positive.

When a measure is converted to the utility scale, `print()` marks it with
"sign flipped" in the table header. It also names it below the tables:

```{r sign-note}
comp <- model_compare(list(m1 = m1, m2 = m2, m3 = m3))
print(comp, measures = c("r2", "mae"))
```

::: {.callout .callout-note}
**Utility vs. loss scale**

We call a measure a **loss** when it has, by default, the orientation
*lower is better*.

By contrast, we call a measure a **utility** if it has, by default, the
orientation *higher is better*.

Which orientation a measure has can be inspected from the `loss` slot of the
`measure_info` attribute of the model object:
```{r measure-meta}
sapply(meta, function(x) x$loss)
```

:::

### Computation of the standard error `<measure>_se_diff`

The standard error summarizes the uncertainty in the estimated difference.
Depending on the measure, different approaches for computing the standard error
are used. We summarize them in four different categories:

| `diff_method` | SE of the difference | Measures | Supplement |
| :--- | :--- | :--- |:--- |
| `"sum"` | `sqrt(N) * sd(d_i)` | `elpd`, `ic` | Eq. TODO-ARXIV |
| `"mean"` | `sd(d_i) / sqrt(N)` | `mlpd`, `mae`, `mse`, `acc`, `brier`, `rps`, `srps` | Eq. TODO-ARXIV |
| `"measure_specific"` | the measure's own `se_diff_fun` | `rmse`, `r2`, `bacc` | Eq. TODO-ARXIV |
| `"custom"` | declared via `custom_measure(se_diff_fun = )` | custom measures | -- |

where `N` is the number of observations and `d_i` the paired difference of
the pointwise estimates of the two models (`comparison - reference`).

The user can check which approach is used for each measure via the
`diff_method` slot in the `measure_info` attribute of the model object:

```{r diff-method}
sapply(meta, function(x) x$diff_method)
```

## Custom measures

A custom measure is any function returning a list with `estimate`, `se`, and
`pointwise`. Define it with `custom_measure()`, which attaches three
declarations to the function:

| argument of `custom_measure()` | attribute | meaning |
| :--- | :--- | :--- |
| `name` | `measure_name` | the name under which the measure is reported |
| `loss` | `measure_loss` | `TRUE` if lower values are better (default `FALSE`) |
| `se_diff_fun` | `measure_se_diff` | how `model_compare()` computes the SE of a difference |

A custom measure always gets `diff_method = "custom"`. Setting the attributes
directly with `attr()` gives the same result.

The argument `se_diff_fun` accepts one of three things:

| value | meaning |
| :--- | :--- |
| `"mean"` | estimate is the mean of the pointwise values. SE is computed as `sd(d_i) / sqrt(N)` |
| `"sum"` | estimate is the sum of the pointwise values. SE is computed as `sqrt(N) * sd(d_i)` |
| a function | custom function for computing the SE, called as `fn(ref, cmp)` |

With `se_diff_fun = NULL` (the default), the measure declares nothing.
`model_compare()` then reports the difference with an `NA` standard error and
a message. `model_compare()` needs no further argument in either case.

We elaborate on these points in the following examples.

### Example (one custom measure)

In the following, we provide a custom measure that is not part of the built-in
measures. The Huber loss is quadratic for small residuals and linear for
large residuals. We define it as follows:

```{r huber-measure}
huber_fn <- function(y, mupred) {
  delta <- 10
  r <- y - colMeans(mupred)
  l <- ifelse(abs(r) <= delta, 0.5 * r^2, delta * (abs(r) - 0.5 * delta))
  list(estimate = mean(l), se = sd(l) / sqrt(length(l)), pointwise = l)
}
huber_se_fn <- function(ref, cmp) {
  d <- cmp$pointwise - ref$pointwise
  sd(d) / sqrt(length(d))
}

huber_measure <- custom_measure(
  fun = huber_fn,
  name = "huber",
  se_diff_fun = huber_se_fn,
  loss = TRUE
)
```

With `name`, we report the measure as "huber". With `loss = TRUE`, we declare
that it is a loss rather than a utility (i.e., lower values are better). Note
that a custom measure is by default considered to be a utility. The argument
`se_diff_fun` declares the standard error of the difference. It is called as
`huber_se_fn(ref, cmp)`, where each argument is
a list describing **one** model with elements `estimate`, `se`, `pointwise`, and
`extra`. This is the general form. The shorthands `"mean"` and `"sum"` cover
the two common cases.

::: {.callout .callout-warning}
A custom measure is by default considered to be a utility (i.e., higher values
are better).

This can be changed with `custom_measure(loss = TRUE)` (see example above).
:::

First, we compute the predictive performance measures for two different models
using `"rmse"` and the custom Huber loss as performance criteria.

```{r custom-measure}
h1 <- fit_measure(fit_m1, measure = list("rmse", huber_measure))
h3 <- fit_measure(fit_m3, measure = list("rmse", huber_measure))
```

(Remember that we have created the `fit_measure` helper at the [beginning of
this tutorial](#example). It simply wraps `loo_pred_measure()`.)

Inspecting the attribute `measure_info` shows that the Huber loss uses
`diff_method = "custom"`, and that our declaration is recorded as
`se_diff_fun`.

```{r custom-diff-method}
str(attr(h1, "measure_info"))
```

`model_compare()` reads the declaration from the objects, so the comparison
needs no further argument:

```{r custom-declared-se}
comp_h <- model_compare(list(m3 = h3, m1 = h1))
print(comp_h, measures = "huber")
```

As discussed above, the difference estimates are always reported on the utility
scale. Therefore `huber_diff` for `m1` is negative.

### Extended example (multiple custom measures)

In the following, we extend the example above by adding one further custom
measure and show how this is implemented. As an additional custom measure we
use RMSE normalized by the standard deviation of the outcome:

```{r extended-example-def}
nrmse_fn <- function(y, mupred) {
  sqe <- (y - colMeans(mupred))^2
  list(
    estimate = sqrt(mean(sqe)) / sd(y),
    se = sqrt(var(sqe) / length(sqe)) / (2 * sqrt(mean(sqe))) / sd(y),
    pointwise = sqe,
    extra = list(sd_y = sd(y))
  )
}

nrmse_se_fn <- function(ref, cmp) {
  n <- length(ref$pointwise)
  mse_ref <- mean(ref$pointwise)
  mse_cmp <- mean(cmp$pointwise)
  se_ref <- sqrt(var(ref$pointwise) / n)
  se_cmp <- sqrt(var(cmp$pointwise) / n)
  cov_rc <- sum((cmp$pointwise - mse_cmp) * (ref$pointwise - mse_ref)) /
    (n * (n - 1))

  rel_ref <- se_ref / sqrt(mse_ref)
  rel_cmp <- se_cmp / sqrt(mse_cmp)
  rho <- cov_rc / (se_ref * se_cmp)

  # the RMSE formula of the supplement (Eq. TODO-ARXIV), written so
  # that a model compared against itself cancels to exactly zero
  se_rmse <- 0.5 * sqrt(
    (rel_cmp - rel_ref)^2 + 2 * rel_cmp * rel_ref * max(1 - rho, 0)
  )

  se_rmse / ref$extra$sd_y
}

nrmse_measure <- custom_measure(
  fun = nrmse_fn,
  name = "nrmse",
  se_diff_fun = nrmse_se_fn,
  loss = TRUE
)
```

Subsequently, we compute the predictive performance measures for two different
models using `"rmse"`, the custom Huber loss, and the custom normalized RMSE
(`"nrmse"`) as performance criteria.

```{r extended-example-fit}
n1 <- fit_measure(fit_m1, list("rmse", huber_measure, nrmse_measure))
n3 <- fit_measure(fit_m3, list("rmse", huber_measure, nrmse_measure))
```

Finally, we perform model comparison with multiple custom measures. Both
measures declare their own `se_diff_fun`, so no argument is needed:

```{r extend-example-fit}
comp_n <- model_compare(list(m3 = n3, m1 = n1))
print(comp_n, measures = "all")
```

## Comparing other predictive sources

`model_compare()` treats every `*_pred_measure()` result the same way. Only the
constructor changes. All models in one call must share one source. A mixed call
is an error.

### K-fold cross-validation

`brms::kfold()` refits each model `K` times, so this is the most expensive step
in this article.

```{r kfold-compare, message=FALSE}
set.seed(SEED)
folds <- kfold_split_random(K = 5, N = nrow(roaches))

kfold_measure <- function(fit, measure) {
  kf <- brms::kfold(fit, folds = folds, save_fits = TRUE)

  kfold_pred_measure(
    y = fit$data$y,
    ypred = brms::kfold_predict(kf, method = "predict")$yrep,
    mupred = brms::kfold_predict(kf, method = "fitted")$yrep,
    kfold = kf,
    measure = measure
  )
}

k1 <- kfold_measure(fit_m1, measures)
k2 <- kfold_measure(fit_m2, measures)
k3 <- kfold_measure(fit_m3, measures)

model_compare(list(m1 = k1, m2 = k2, m3 = k3))
```

`print()` names the source above the table. 
The PSIS-LOO block is gone, because Pareto $\hat{k}$ describes the LOO
approximation only.

### Held-out test data

Here we hold out 60 buildings, refit each model on the rest, and score the
held-out ones. `ylp_test` supplies `elpd_test`; `ylp` stays the training
log-likelihood.

```{r test-compare}
set.seed(SEED)
test_id <- sample(nrow(roaches), 60)
roaches_train <- roaches[-test_id, ]
roaches_test <- roaches[test_id, ]

test_measure <- function(fit, measure) {
  fit_train <- update(fit, newdata = roaches_train, refresh = 0)
  test_pred_measure(
    y = roaches_test$y,
    ypred = brms::posterior_predict(fit_train, newdata = roaches_test),
    mupred = brms::posterior_epred(fit_train, newdata = roaches_test),
    ylp = brms::log_lik(fit_train),
    ylp_test = brms::log_lik(fit_train, newdata = roaches_test),
    measure = measure
  )
}

t1 <- test_measure(fit_m1, measures)
t2 <- test_measure(fit_m2, measures)
t3 <- test_measure(fit_m3, measures)

model_compare(list(m1 = t1, m2 = t2, m3 = t3))
```

### In-sample

In-sample scores need no refit. They are optimistically biased and favor the
more complex model, so `model_compare()` warns:

```{r insample-compare}
insample_measure <- function(fit, measure) {
  insample_pred_measure(
    y = fit$data$y,
    ypred = brms::posterior_predict(fit),
    mupred = brms::posterior_epred(fit),
    ylp = brms::log_lik(fit),
    measure = measure
  )
}

i1 <- insample_measure(fit_m1, measures)
i2 <- insample_measure(fit_m2, measures)
i3 <- insample_measure(fit_m3, measures)

model_compare(list(m1 = i1, m2 = i2, m3 = i3))
```

Everything else in this article, custom measures, the `print`
arguments `measures` and `simplify`, works unchanged for all four sources.

## Summary

```{r summary-table, echo=FALSE}
knitr::kable(
  data.frame(
    Measure = c("elpd, ic", "mlpd, mae, mse, acc, brier, rps, srps",
                "rmse", "r2", "bacc", "custom"),
    `diff_method` = c("sum", "mean", "measure_specific",
                      "measure_specific", "measure_specific",
                      "custom"),
    `Standard error` = c(
      "sqrt(N) * sd(d_i)",
      "sd(d_i) / sqrt(N)",
      "delta method on the MSE scale",
      "delta method, including baseline uncertainty",
      "stratified by class, paired within class",
      "from se_diff_fun: your function, \"sum\", \"mean\", or NA"
    ),
    check.names = FALSE
  )
)
```

The things worth remembering:

1. The difference is always `estimate_cmp - estimate_ref`, reported so that
   negative means worse.
2. The standard error is always paired, because the models are evaluated on the
   same observations.
3. `NA` in a standard error column means "not available", not "zero" and not
   "an error occurred".
4. A custom measure is assumed to be a utility. If yours is a loss, say so with
   `custom_measure(loss = TRUE)`, or its differences and its ranking will
   come out reversed.

## References

Sivula, T., Magnusson, M., Matamoros, A. A., and Vehtari, A. (2025).
Uncertainty in Bayesian leave-one-out cross-validation based model comparison.
*Bayesian Analysis*. doi:10.1214/25-BA1569

Vehtari, A., Gelman, A., and Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics and
Computing*, 27(5):1413–1432. doi:10.1007/s11222-016-9696-4

<!-- TODO-ARXIV: add the title, the year and the arXiv ID before the release -->
Vehtari, A., Bockting, F., Axen, S., and Weber, F. (2026).
*TODO-ARXIV-TITLE*. arXiv preprint arXiv:TODO-ARXIV-ID.

## See also

- [Computing predictive performance measures](pred-measure-workflow.html) — how
  to produce the objects compared here.
- [Overview of scores and metrics](overview-measures.html) — definitions and
  formulas for each measure.
- `?"loo-glossary"` — the `measure_info` attribute in full.
- Vehtari et al. (2026), *TODO-ARXIV-TITLE*: the supplement with the formulas for every measure and standard error in this article.