---
title: "Predictive model performance: Overview of measures"
author: "Florence Bockting"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    css:
      - !expr system.file("rmarkdown/templates/html_vignette/resources/vignette.css", package = "rmarkdown")
      - styles.css
    toc: true
    toc_depth: 3
params:
  EVAL: true
vignette: >
  %\VignetteIndexEntry{Predictive model performance: Overview of measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="../children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = params$EVAL,
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

```{r, child="../children/LOAD-BRMS-GITHUB.txt", eval=params$EVAL}
```

## Introduction

A *predictive measure* summarizes how well a model predicts data. **loo**
provides twelve built-in measures and the possibility to create a custom
measure. Each built-in measure has its own `measure_*()` function.

This vignette introduces each built-in measure. For each measure it gives:

1. the motivation,
2. when to use it,
3. the key formula,
4. the standard error,
5. the implementation in **loo**, and
6. a plot and its interpretation, comparing a model with predictors to an intercept-only model.

This vignette states only the key formulas. The full definitions, the
estimators, and the standard errors are in our supplement
(Vehtari et al., 2026). <!-- TODO -->

This vignette hides the code for the plots. The full source code is in the
[`loo` GitHub repository](https://github.com/stan-dev/loo/blob/master/vignettes/articles-online-only/overview-measures.Rmd).

::: {.callout .callout-note}
**What this vignette does not cover**

+ **Predictive schemes**
  + Examples in this vignette use in-sample predictions.
  + All measures can be used with leave-one-out, K-fold, or test-data predictions; see [Predictive model performance: Predictive schemes with `pred_measure()`](pred-measure-workflow.html).
+ **Model comparison** 
  + Using the `model_compare()` function; see [Model comparison: Explanation of `model_compare()`](model-comparison.html).
  + A tutorial on using model comparison within the Bayesian workflow; see [Model comparison: Case Study with `model_compare()`]() <!-- TODO -->
:::

## Overview of built-in measures

| Keyword | Measure | Family | Input | Outcome | Natural scale | Aggregation |
| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
| `elpd` | expected log pointwise predictive density | [log score](#log-score) | `ylp` | any | utility | sum |
| `mlpd` | mean log predictive density | [log score](#log-score) | `ylp` | any | utility | mean |
| `ic` | information criterion, $-2\,\text{elpd}$ | [log score](#log-score) | `ylp` | any | loss | sum |
| `rps` | (continuous) ranked probability score | [ranked probability score](#ranked-probability-score) | `ypred` | continuous or ordered | loss | mean |
| `srps` | scaled (continuous) ranked probability score | [ranked probability score](#ranked-probability-score) | `ypred` | continuous or ordered | utility | mean |
| `brier` | Brier score | [Brier score](#brier-score) | `ypred` | binary | loss | mean |
| `mae` | mean absolute error | [point error](#point-error) | `mupred` | continuous | loss | mean |
| `mse` | mean squared error | [point error](#point-error) | `mupred` | continuous | loss | mean |
| `rmse` | root mean squared error | [point error](#point-error) | `mupred` | continuous | loss | function of means |
| `r2` | predictive $R^2$ | [predictive $R^2$](#predictive-r2) | `mupred` | continuous | utility | function of means |
| `acc` | classification accuracy | [classification accuracy](#classification-accuracy) | `mupred` | binary or nominal | utility | mean |
| `bacc` | balanced accuracy | [classification accuracy](#classification-accuracy) | `mupred` | binary or nominal | utility | weighted sum |

The columns mean:

+ **Natural scale:** A *loss* is better when it is lower. A *utility* is better
  when it is higher. `measure_*()` reports each measure on its natural scale.
+ **Aggregation:** Most measures are a sum or a mean of pointwise terms. RMSE
  and $R^2$ have no pointwise terms that add up to the estimate. Balanced
  accuracy weights each pointwise term by its class, so that the terms sum to
  the estimate. For more details see Vehtari et al. (2026). <!-- TODO -->

## Notation

This vignette uses one notation for all measures. The table maps each symbol to an
argument of `measure_*()` or to a field of its result.

| Symbol | Meaning | In **loo** |
| :--- | :--- | :--- |
| $y_i$ | observed outcome $i$ | `y` |
| $\theta^s$ | posterior draw $s$, $s = 1, \ldots, S$ | rows of the input matrices |
| $\ell_i^s = \log p(y_i \mid \theta^s)$ | log density of $y_i$ under draw $s$ | `ylp`, an $S \times n$ matrix |
| $\tilde y_i^s$ | posterior predictive draw | `ypred`, an $S \times n$ matrix; for `brier`, the event probabilities $p(\tilde y_i = 1 \mid \theta^s)$  |
| $\mathbb{E}[\tilde y_i \mid \theta^s]$ | predictive mean under draw $s$ | `mupred`, an $S \times n$ matrix, or $S \times n \times C$ for $C$ classes |
| $w_i^s$ | normalized weight of draw $s$ (Vehtari et al., 2026, Eq. 2.5) | `log_weights`, on the log scale and unnormalized; **loo** normalizes them. `NULL` gives $w_i^s = 1/S$ |
| $\text{elpd}_i$, $\text{rps}_i$, $\text{ae}_i$, … | pointwise term of a measure, on its natural scale | `$pointwise` |
| $\widehat{\text{elpd}}$, $\widehat{\text{rps}}$, $\widehat{\text{mae}}$, … | estimate of a measure | `$estimates[, "Estimate"]` |
| $\widehat{\text{SE}}$ | standard error of the estimate | `$estimates[, "SE"]` |
| $\widehat{\text{Var}}$ | sample variance over the observations | |

::: {.callout .callout-note}
**Weights**

Every estimator in this vignette uses weights $w_i^s$. With
`log_weights = NULL`, all draws get the weight $1/S$. This gives in-sample
predictions. Other weights give other predictive schemes, for example PSIS-LOO.
:::

## Data and models

```{r load-libraries}
library(loo)
library(ggplot2)
theme_set(theme_minimal())
```

For each outcome type we fit two `brms` models:

+ **with predictors**: a model that uses the predictors of the data,
+ **intercept only**: a model that predicts the same distribution for every
  observation.

The intercept-only model is a baseline. It shows how a measure changes when a
model uses no information from the predictors. The model with predictors is
better, but not necessarily good. Fitting is not the topic of this vignette. The
model code is in the [appendix](#appendix-model-code).

| Object | Data | Outcome | Predictors | Used for |
| :--- | :--- | :--- | :--- | :--- |
| `fits_cont` | `faithful` (`datasets`) | continuous, normal | `waiting` | log score, CRPS, point error*, $R^2$ |
| `fits_bin` | `Pima.te` (`MASS`) | binary, Bernoulli | `glu`, `bmi`, `age` | Brier score |
| `fits_ord` | `wine` (`ordinal`) | ordered, 5 categories | `temp`, `contact` | RPS |
| `fits_cat` | `penguins` (`datasets`) | nominal, 3 unbalanced classes | `bill_len` | accuracy |

*point error measures include MAE, MSE, and RMSE.

Each `fits_*` object is a list with the elements `"with predictors"` and
`"intercept only"`.

**Prediction tasks.** The models predict the following outcomes:

| Model | Outcome | Predictors | N |
|:--|:--|:--|:--|
| Eruption duration (`fits_cont`) | `eruptions`: duration of the eruption, in minutes | `waiting`: waiting time to the eruption, in minutes | 272 eruptions |
| Diabetes (`fits_bin`) | `diabetes`: 1 if the woman has diabetes (no: 223, yes: 109) | `glu`: plasma glucose concentration; `bmi`: body mass index; `age`: age in years | 332 women |
| Wine bitterness (`fits_ord`) | `rating`: bitterness of the wine, on an ordered scale from 1 to 5 (1: 5, 2: 22, 3: 26, 4: 12, 5: 7) | `temp`: temperature during the crushing of the grapes, cold or warm; `contact`: contact between juice and skins, no or yes | 72 ratings from 9 judges |
| Penguin species (`fits_cat`) | `species`: Adelie (151), Chinstrap (68), or Gentoo (123) | `bill_len`: bill length, in mm | 342 penguins |

```{r fit-models, echo=FALSE, results="hide"}
SEED <- 42

fit_pair <- function(formula, data, family) {
  fit <- brms::brm(formula, data = data, family = family,
                   chains = 2, iter = 1000, seed = SEED, refresh = 0)
  list(
    "with predictors" = fit,
    "intercept only" = update(fit, formula. = . ~ 1, seed = SEED, refresh = 0)
  )
}

# continuous
fits_cont <- fit_pair(eruptions ~ waiting, faithful, gaussian())

# binary
data(Pima.te, package = "MASS")
Pima.te$diabetes <- as.integer(Pima.te$type == "Yes")
fits_bin <- fit_pair(diabetes ~ glu + bmi + age, Pima.te, brms::bernoulli())

# ordered
data(wine, package = "ordinal")
fits_ord <- fit_pair(rating ~ temp + contact, wine, brms::cumulative("logit"))

# nominal: Adelie is the most frequent species
penguins_cc <- datasets::penguins[!is.na(datasets::penguins$bill_len), ]
fits_cat <- fit_pair(species ~ bill_len, penguins_cc, brms::categorical())
```

## Log score {#log-score}

**Motivation.** The log score evaluates the predictive density at the observed
value. A model gets a high score when it puts much probability near the
observation. It uses the whole predictive distribution, and it strongly
penalizes observations the model considers very unlikely.

**When to use.** The log score works for any outcome type. It is the default
measure in **loo**. It needs the log-likelihood draws `ylp`.

**Key formula.** The pointwise term and the estimate are
(Vehtari et al., 2026, Eqs. 2.16–2.18) <!-- TODO -->

$$
\text{elpd}_i = \log \sum_{s=1}^S \exp\!\left(\log w_i^s + \ell_i^s\right),
\qquad
\widehat{\text{elpd}} = \sum_{i=1}^n \text{elpd}_i .
$$

**loo** reports two rescaled versions: 

+ $\widehat{\text{mlpd}} =
\widehat{\text{elpd}} / n$ on the mean scale, and 
+ the information criterion
$\widehat{\text{ic}} = -2\,\widehat{\text{elpd}}$.

**Standard error.** The elpd is a sum, the mlpd is a mean
(Vehtari et al., 2026, Eqs. 2.8, 2.9): <!-- TODO -->

$$
\widehat{\text{SE}}(\widehat{\text{elpd}}) =
  \sqrt{n\,\widehat{\text{Var}}(\text{elpd}_i)},
\qquad
\widehat{\text{SE}}(\widehat{\text{mlpd}}) =
  \sqrt{\widehat{\text{Var}}(\text{elpd}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{ic}}) =
  2\,\widehat{\text{SE}}(\widehat{\text{elpd}}) .
$$

**Implementation.**

```{r log-score}
ylp_cont <- lapply(fits_cont, brms::log_lik)
elpd_cont <- lapply(ylp_cont, function(ylp) measure_elpd(ylp = ylp))
mlpd_cont <- lapply(ylp_cont, function(ylp) measure_mlpd(ylp = ylp))
ic_cont <- lapply(ylp_cont, function(ylp) measure_ic(ylp = ylp))

# output structure of elpd measure for model with predictors
str(elpd_cont$`with predictors`)
```

**Plot.** The models `fits_cont` predict `eruptions`, the duration of each of
272 eruptions of the Old Faithful geyser (`faithful` data). Each point is one
eruption. The x-axis shows the distance of the observed duration from the
predictive mean. The y-axis shows the pointwise log score $\text{elpd}_i$.
The dashed line is the mlpd, the mean of the points. Higher is better.

```{r log-score-plot, echo = FALSE}
mu_hat_cont <- lapply(fits_cont, function(fit) colMeans(brms::posterior_epred(fit)))
y_cont <- fits_cont[[1]]$data$eruptions
df_elpd <- do.call(rbind, lapply(names(ylp_cont), function(m) {
  data.frame(
    model = m,
    d = y_cont - mu_hat_cont[[m]],
    elpd_i = as.vector(elpd_cont[[m]]$pointwise)
  )
}))
df_lines <- data.frame(
  model = names(ylp_cont),
  elpd = sapply(elpd_cont, function(r) r$estimates[, "Estimate"]),
  mlpd = sapply(mlpd_cont, function(r) r$estimates[, "Estimate"]),
  ic = sapply(ic_cont, function(r) r$estimates[, "Estimate"])
)

ggplot(df_elpd, aes(d, elpd_i)) +
  geom_point(alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = mlpd), linetype = "dashed") +
  geom_text(data = df_lines,
            aes(x = -Inf, y = -Inf,
                label = sprintf("elpd = %.0f\nic = %.0f\nmlpd = %.2f", elpd, ic, mlpd)),
            hjust = -0.1, vjust = -0.3, size = 3) +
  facet_wrap(~model) +
  labs(x = "observed duration minus predictive mean", y = "pointwise log score")

```

**Interpretation.** Both models predict a normal distribution, so the points
lie on a parabola. The intercept-only model predicts the same wide
distribution for every eruption. Its parabola is flat: no eruption gets a very
high or a very low score. The predictor `waiting` makes the predictive
distribution narrower. Eruptions close to the predictive mean then get a
higher score, and eruptions far from it get a lower score. The log score
drops with the square of the distance, so these few eruptions pull the
estimate down. The mlpd is
`r round(mlpd_cont[["with predictors"]]$estimates[, "Estimate"], 2)` against
`r round(mlpd_cont[["intercept only"]]$estimates[, "Estimate"], 2)`.

## Ranked probability score {#ranked-probability-score}

**Motivation.** The ranked probability score compares the predictive
distribution with the observed value through absolute distances. Unlike the
log score, it rewards predictions that are *close* to the observation, not only
predictions that hit it. It penalizes outlying observations less than the log
score, but it is not robust to them (Bolin and Wallin, 2023).

**When to use.** Use `rps` for ordered categorical outcomes (RPS) and for
continuous outcomes (CRPS). It is on the scale of $y$, so a value can be
read in the units of the outcome. Use `srps` when outcomes differ strongly in
their predictive spread. Both need the posterior predictive draws `ypred`.

**Key formula.** Let $X$ and $X'$ be independent draws from the predictive
distribution of observation $i$ (Vehtari et al., 2026, Eqs. 2.20-2.22): <!-- TODO -->

$$
\text{rps}_i = \mathbb{E}|X - y_i| - \tfrac{1}{2}\,\mathbb{E}|X - X'|, \qquad
\text{srps}_i = -\frac{\mathbb{E}|X - y_i|}{\mathbb{E}|X - X'|}
  - \tfrac{1}{2}\log \mathbb{E}|X - X'| .
$$

The estimates are the means $\widehat{\text{rps}} = \frac{1}{n}\sum_i
\text{rps}_i$ and $\widehat{\text{srps}} = \frac{1}{n}\sum_i \text{srps}_i$.
**loo** estimates both expectations from one set of draws. The estimator of
$\mathbb{E}|X - X'|$ is a probability-weighted moment estimator
(Vehtari et al., 2026, Eq. 2.25). <!-- TODO -->
`rps` is a loss (lower is better), and `srps` is a utility (higher is better).

For ordered outcomes, code the categories $1, \ldots, K$. Let $F_i$ be the
predictive CDF of observation $i$. Then
$\text{rps}_i = \sum_{k=1}^{K-1} \left(F_i(k) - \mathbb{1}\{y_i \le k\}\right)^2$.
This is the RPS of Epstein (1969) without the factor $1/(K-1)$. Divide by $K-1$
to get values between 0 and 1.

**Standard error.** Both are means (Vehtari et al., 2026, Eq. 2.8): <!-- TODO -->

$$
\widehat{\text{SE}}(\widehat{\text{rps}}) =
  \sqrt{\widehat{\text{Var}}(\text{rps}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{srps}}) =
  \sqrt{\widehat{\text{Var}}(\text{srps}_i) / n} .
$$

**Implementation.** For the continuous outcome, `measure_rps()` gives the CRPS
and the SCRPS:

```{r crps}
y_cont <- fits_cont[[1]]$data$eruptions
ypred_cont <- lapply(fits_cont, brms::posterior_predict)

crps_cont <- lapply(ypred_cont, function(yp) measure_rps(y = y_cont, ypred = yp))
scrps_cont <- lapply(ypred_cont, function(yp) measure_srps(y = y_cont, ypred = yp))

# output structure of crps measure for model with predictors
str(crps_cont$`with predictors`)
```

For the ordered outcome, the same functions give the RPS and the SRPS:

```{r rps}
y_ord <- as.integer(fits_ord[[1]]$data$rating)
ypred_ord <- lapply(fits_ord, brms::posterior_predict)

rps_ord <- lapply(ypred_ord, function(yp) measure_rps(y = y_ord, ypred = yp))
srps_ord <- lapply(ypred_ord, function(yp) measure_srps(y = y_ord, ypred = yp))

# output structure of srps measure for model with predictors
str(srps_ord$`with predictors`)
```

**Plot.** The models `fits_cont` predict `eruptions`, the duration of each of
272 eruptions of the Old Faithful geyser (`faithful` data). Each point is one
eruption. The x-axis shows the distance of the observed duration from the
predictive mean, as in the log score plot.
The y-axis shows the pointwise CRPS $\text{rps}_i$, in minutes.
The dashed line is the estimate, the mean of the points. Lower is better.

```{r crps-plot, echo = FALSE}
df_crps <- do.call(rbind, lapply(names(ypred_cont), function(m) {
  data.frame(
    model = m,
    d = y_cont - mu_hat_cont[[m]],
    crps_i = as.vector(crps_cont[[m]]$pointwise)
  )
}))
df_lines <- data.frame(
  model = names(crps_cont),
  crps = sapply(crps_cont, function(r) r$estimates[, "Estimate"])
)

ggplot(df_crps, aes(d, crps_i)) +
  geom_point(alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = crps), linetype = "dashed") +
  facet_wrap(~model) +
  labs(x = "observed duration minus predictive mean", y = "pointwise CRPS")
```

**Interpretation.** In both panels the points form a V with a round bottom.
Far from the predictive mean, the CRPS grows about linearly with the distance,
like an absolute error. The log score, by contrast, drops with the square of
the distance. The narrower distribution of the model with predictors moves the
whole V down: even an exact hit gets a lower CRPS. Its estimate is
`r round(crps_cont[["with predictors"]]$estimates[, "Estimate"], 2)` against
`r round(crps_cont[["intercept only"]]$estimates[, "Estimate"], 2)`.

**Plot for the ordered outcome.** 
The models `fits_ord` predict `rating`, the bitterness of a wine on a scale
from 1 to 5. There are 72 ratings from 9 judges (`wine` data). Each point is one
rating. The x-axis shows the distance of the observed rating from the
predictive mean. The y-axis shows the pointwise RPS $\text{rps}_i$.
The color shows the observed rating. The dashed line is the estimate.
Lower is better.

```{r rps-plot, echo = FALSE}
df_rps <- do.call(rbind, lapply(names(ypred_ord), function(m) {
  data.frame(
    model = m,
    d = y_ord - colMeans(ypred_ord[[m]]),
    rps_i = as.vector(rps_ord[[m]]$pointwise),
    rating = factor(y_ord)
  )
}))
df_lines <- data.frame(
  model = names(rps_ord),
  rps = sapply(rps_ord, function(r) r$estimates[, "Estimate"])
)

ggplot(df_rps, aes(d, rps_i)) +
  geom_point(aes(color = rating), alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = rps), linetype = "dashed") +
  scale_color_viridis_d(end = 0.9) +
  facet_wrap(~model) +
  labs(x = "observed rating minus predictive mean", y = "pointwise RPS",
       color = "observed rating")
```

**Interpretation.** The intercept-only model gives every wine the same
predictive distribution. Its points therefore form five tight clusters, one for
each observed rating. The RPS uses the order of the categories: a rating far from
the predictive mean gets a higher RPS. The model with predictors gives each
combination of `temp` and `contact` its own predictive distribution. Each spot
therefore splits into smaller groups. Most points move closer to zero
distance, and together they form a V, as for the CRPS. Its estimate is lower:
`r round(rps_ord[["with predictors"]]$estimates[, "Estimate"], 2)` against
`r round(rps_ord[["intercept only"]]$estimates[, "Estimate"], 2)`.

## Brier score {#brier-score}

**Motivation.** The Brier score is the squared distance between the predicted
probability of an event and the observed outcome, 0 or 1. It rewards
probabilities that are both calibrated and sharp, that is, close to 0 or 1.

**When to use.** Use it for binary outcomes. **loo** implements the binary
Brier score only. `measure_brier()` needs the event probabilities of the
draws, passed as `ypred`.

**Key formula.** With the predicted event probability
$\hat p_i = \sum_s w_i^s\, p(\tilde y_i = 1 \mid \theta^s)$
(Vehtari et al., 2026, Eqs. 2.26–2.28), <!-- TODO -->

$$
\text{bs}_i = \left(\hat p_i - y_i\right)^2, \qquad
\widehat{\text{bs}} = \frac{1}{n}\sum_{i=1}^n \text{bs}_i .
$$

The Brier score is a loss (lower is better). It ranges from 0 to 1.

**Standard error.** The Brier score is a mean (Vehtari et al., 2026, Eq. 2.8): <!-- TODO -->

$$
\widehat{\text{SE}}(\widehat{\text{bs}}) =
  \sqrt{\widehat{\text{Var}}(\text{bs}_i) / n} .
$$

**Implementation.**

```{r brier}
y_bin <- fits_bin[[1]]$data$diabetes
p_bin <- lapply(fits_bin, brms::posterior_epred)
brier_bin <- lapply(p_bin, function(p) measure_brier(y = y_bin, ypred = p))

str(brier_bin$`with predictors`)
```

**Plot.** The models `fits_bin` predict `diabetes`, 1 if the woman has
diabetes, for 332 women (`Pima.te` data). Each point is one woman.
The x-axis shows the predicted probability $\hat p_i$ of diabetes.
Every point lies on one of two curves: $\hat p_i^2$ if the woman has no
diabetes, and $(1 - \hat p_i)^2$ if she has diabetes. Lower is better.

```{r brier-plot, echo = FALSE}
df_brier <- do.call(rbind, lapply(names(p_bin), function(m) {
  data.frame(
    model = m,
    p = colMeans(p_bin[[m]]),
    bs_i = as.vector(brier_bin[[m]]$pointwise),
    diabetes = factor(y_bin, levels = 0:1, labels = c("no", "yes"))
  )
}))

max_bs_yes <- max(df_brier$bs_i[df_brier$model == "with predictors" &
                                  df_brier$diabetes == "yes"])

ggplot(df_brier, aes(p, bs_i, color = diabetes)) +
  geom_function(fun = function(p) p^2, color = "grey70", linetype = "dotted") +
  geom_function(fun = function(p) (1 - p)^2, color = "grey70", linetype = "dotted") +
  geom_point(alpha = 0.5) +
  facet_wrap(~model) +
  labs(x = "predicted probability of diabetes",
       y = "pointwise Brier score", color = "diabetes")
```

**Interpretation.** The intercept-only model gives every woman the same
probability, the proportion of women with diabetes. All its points fall on two
spots, one on each curve. The model with predictors spreads the probabilities.
A woman moves down her curve when the model is confident and right. She moves
up her curve when the model is confident and wrong. Most women without
diabetes get a low probability and a score near 0. A few women with diabetes
get a low probability and a score up to `r round(max_bs_yes, 1)`.
Its estimate is lower:
`r round(brier_bin[["with predictors"]]$estimates[, "Estimate"], 2)` against
`r round(brier_bin[["intercept only"]]$estimates[, "Estimate"], 2)`.

## Point error {#point-error}

**Motivation.** The point-error measures reduce the predictive distribution to
one point prediction, the predictive mean. They then measure the distance
between that prediction and the observed value. They are easy to interpret,
but they ignore the predictive uncertainty.

**When to use.** Use them for continuous outcomes when the quality of the point
prediction matters. MAE is less sensitive to large errors than MSE and RMSE.
MAE and RMSE are on the scale of $y$. All three need the predictive means
`mupred`.

**Which point prediction.** Each loss rewards a different point prediction.
The squared error is smallest in expectation for the predictive mean. The
absolute error is smallest for the predictive median (Gneiting, 2011).
**loo** uses the predictive mean for all three measures. MSE and RMSE
therefore evaluate the prediction they reward. MAE evaluates the mean, not
the median. In the normal model here, the predictive distribution is close to
symmetric. Its mean and its median are then about equal.

**MAE or RMSE.** Neither is better in general. RMSE fits normal errors, and
MAE fits Laplace errors with heavier tails (Hodson, 2022). RMSE is never
smaller than MAE. They are equal only when all absolute errors are equal.

**Key formula.** With the point prediction
$\hat y_i = \sum_s w_i^s\, \mathbb{E}[\tilde y_i \mid \theta^s]$, the pointwise
errors are $\text{ae}_i = |y_i - \hat y_i|$ and
$\text{sqe}_i = (y_i - \hat y_i)^2$ (Vehtari et al., 2026, Eqs. 2.29-2.34): <!-- TODO -->

$$
\widehat{\text{mae}} = \frac{1}{n}\sum_{i=1}^n \text{ae}_i, \qquad
\widehat{\text{mse}} = \frac{1}{n}\sum_{i=1}^n \text{sqe}_i, \qquad
\widehat{\text{rmse}} = \sqrt{\widehat{\text{mse}}} .
$$

All three are losses.

**Standard error.** MAE and MSE are means (Vehtari et al., 2026, Eq. 2.8). <!-- TODO -->
RMSE is a function of a mean, so its standard error uses the delta method
(Vehtari et al., 2026, Eq. 2.33): <!-- TODO -->

$$
\widehat{\text{SE}}(\widehat{\text{mae}}) =
  \sqrt{\widehat{\text{Var}}(\text{ae}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{mse}}) =
  \sqrt{\widehat{\text{Var}}(\text{sqe}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{rmse}}) =
  \frac{\widehat{\text{SE}}(\widehat{\text{mse}})}{2\,\widehat{\text{rmse}}} .
$$

**Pointwise terms.** `measure_rmse()` has no pointwise terms of its own. Its
`$pointwise` field holds the squared errors $\text{sqe}_i$, the same as
`measure_mse()`.

**Implementation.**

```{r point-error}
mu_cont <- lapply(fits_cont, brms::posterior_epred)

m_mae <- lapply(mu_cont, function(mu) measure_mae(y = y_cont, mupred = mu))
m_mse <- lapply(mu_cont, function(mu) measure_mse(y = y_cont, mupred = mu))
m_rmse <- lapply(mu_cont, function(mu) measure_rmse(y = y_cont, mupred = mu))

str(m_mse$`with predictors`)
```

**Plot.** The models `fits_cont` predict `eruptions`, the duration of each of
272 eruptions of the Old Faithful geyser (`faithful` data).
Each point is one eruption. The x-axis shows the error of
the point prediction, $y_i - \hat y_i$. The y-axis shows the pointwise squared
error $\text{sqe}_i$. The dashed line is the MSE, the mean of the points. Lower
is better.

```{r point-error-plot, echo = FALSE}
df_pe <- do.call(rbind, lapply(names(mu_cont), function(m) {
  data.frame(
    model = m,
    e = y_cont - colMeans(mu_cont[[m]]),
    sqe_i = as.vector(m_mse[[m]]$pointwise)
  )
}))
df_lines <- data.frame(
  model = names(mu_cont),
  mae = sapply(m_mae, function(r) r$estimates[, "Estimate"]),
  mse = sapply(m_mse, function(r) r$estimates[, "Estimate"]),
  rmse = sapply(m_rmse, function(r) r$estimates[, "Estimate"])
)
top_share <- sapply(m_mse, function(r) {
  sqe <- sort(r$pointwise, decreasing = TRUE)
  sum(sqe[seq_len(ceiling(0.1 * length(sqe)))]) / sum(sqe)
})

ggplot(df_pe, aes(e, sqe_i)) +
  geom_point(alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = mse), linetype = "dashed") +
  geom_text(data = df_lines,
            aes(x = -Inf, y = Inf,
                label = sprintf("MSE = %.2f\nRMSE = %.2f", mse, rmse)),
            hjust = -0.1, vjust = 1.2, size = 3) +
  facet_wrap(~model) +
  labs(x = "observed duration minus point prediction",
       y = "pointwise squared error")
```

**Interpretation.** The points lie on a parabola. Compare this with the V of
the CRPS: the squared error grows with the square of the distance. The
intercept-only model predicts the mean duration for every eruption. Most
eruptions are either short or long, so few points lie near zero error. The
predictor `waiting` separates short and long eruptions. Its MSE is
`r round(m_mse[["with predictors"]]$estimates[, "Estimate"], 2)` against
`r round(m_mse[["intercept only"]]$estimates[, "Estimate"], 2)`.
For the model with predictors, the 10% largest errors make up
`r round(100 * top_share[["with predictors"]])`% of the MSE. The MSE is in
squared minutes. The RMSE, its square root, is back in minutes.

**Plot for MAE and RMSE.** The data are the same as in the plot above: the
duration of 272 eruptions. The histogram shows the pointwise absolute errors
$\text{ae}_i$. The dashed line is the MAE. The dotted line is the RMSE. Both
are in minutes.

```{r point-error-mae-rmse-plot, echo = FALSE}
df_ae <- do.call(rbind, lapply(names(mu_cont), function(m) {
  data.frame(model = m, ae_i = as.vector(m_mae[[m]]$pointwise))
}))
df_ref <- rbind(
  data.frame(model = df_lines$model, measure = "MAE", value = df_lines$mae),
  data.frame(model = df_lines$model, measure = "RMSE", value = df_lines$rmse)
)

ggplot(df_ae, aes(ae_i)) +
  geom_histogram(binwidth = 0.1, boundary = 0, fill = "grey70", color = "white") +
  geom_vline(data = df_ref, aes(xintercept = value, linetype = measure)) +
  geom_text(data = df_ref, aes(x = value, y = Inf, label = measure),
            angle = 90, vjust = -0.4, hjust = 1.1, size = 3) +
  scale_linetype_manual(values = c(MAE = "dashed", RMSE = "dotted")) +
  facet_wrap(~model) +
  labs(x = "pointwise absolute error (minutes)",
       y = "number of eruptions", linetype = NULL)
```

**Interpretation.** In both models the RMSE lies to the right of the MAE. The
squared error gives more weight to the large errors. The
model with predictors moves the histogram to the left, so both measures
decrease. Its MAE is
`r round(m_mae[["with predictors"]]$estimates[, "Estimate"], 2)` and its RMSE
is `r round(m_rmse[["with predictors"]]$estimates[, "Estimate"], 2)`.

## Predictive $R^2$ {#predictive-r2}

**Motivation.** The predictive $R^2$ compares the squared error of the model
with the squared error of the simplest prediction, the mean of the
observations. A value of 0 means the model predicts no better than the mean.

**When to use.** Use it for continuous outcomes, when a unitless summary of the
point prediction is needed. It is a predictive version of the classical $R^2$,
not the Bayesian $R^2$ of Gelman et al. (2019). It needs `mupred`.

**Key formula.** With $\text{sqe}_i$ and $\widehat{\text{mse}}$ from the
[point error](#point-error) section (Vehtari et al., 2026, Eqs. 2.35, 2.36), <!-- TODO -->

$$
\widehat{R^2} = 1 - \frac{\widehat{\text{mse}}}{\widehat{\text{mse}}_y},
\qquad
\widehat{\text{mse}}_y = \frac{1}{n}\sum_{i=1}^n (y_i - \bar y)^2 .
$$

$R^2$ is a utility.

**Standard error.** $R^2$ is a function of two means, so its standard error
uses the delta method. With $c = \widehat{\text{mse}} /
\widehat{\text{mse}}_y$ (Vehtari et al., 2026, Eq. 2.39), <!-- TODO -->

$$
\widehat{\text{SE}}(\widehat{R^2}) = \frac{1}{\widehat{\text{mse}}_y}
  \sqrt{\widehat{\text{Var}}\!\left(\text{sqe}_i - c\,(y_i - \bar y)^2\right) / n} .
$$

**Pointwise terms.** $R^2$ has no pointwise terms of its own. The
`$pointwise` field of `measure_r2()` holds the squared errors
$\text{sqe}_i$ of the model.

**Implementation.**

```{r r2}
m_r2 <- lapply(mu_cont, function(mu) measure_r2(y = y_cont, mupred = mu))

str(m_r2$`with predictors`)
```

**Plot.** The models `fits_cont` predict `eruptions`, the duration of each of
272 eruptions of the Old Faithful geyser (`faithful` data).
Each point is one eruption. The x-axis shows the distance
of the observed duration from the mean duration. The y-axis shows the
contribution of the eruption to $R^2$,

$$
r_i = \frac{(y_i - \bar y)^2 - \text{sqe}_i}{\widehat{\text{mse}}_y},
\qquad
\frac{1}{n}\sum_{i=1}^n r_i = \widehat{R^2} .
$$

A point above zero is an eruption for which the model predicts better than
the mean. The dashed line is $\widehat{R^2}$, the mean of the points. Higher is
better.

```{r r2-plot, echo = FALSE}
mse_y <- mean((y_cont - mean(y_cont))^2)
df_r2 <- do.call(rbind, lapply(names(m_r2), function(m) {
  data.frame(
    model = m,
    d = y_cont - mean(y_cont),
    r_i = ((y_cont - mean(y_cont))^2 - as.vector(m_mse[[m]]$pointwise)) / mse_y
  )
}))
df_lines <- data.frame(
  model = names(m_r2),
  r2 = sapply(m_r2, function(r) r$estimates[, "Estimate"])
)

ggplot(df_r2, aes(d, r_i)) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_point(alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = r2), linetype = "dashed") +
  geom_text(data = df_lines,
            aes(x = -Inf, y = Inf, label = sprintf("R2 = %.2f", r2)),
            hjust = -0.1, vjust = 1.2, size = 3) +
  facet_wrap(~model) +
  labs(x = "observed duration minus mean duration",
       y = "contribution to R2")
```

**Interpretation.** The intercept-only model predicts about the mean duration
for every eruption. All its contributions are close to zero, and so is its
$R^2$. For the model with predictors, the points form a U.
An eruption close to the mean duration contributes about zero or less:
the mean already predicts it well, and the model cannot do much better.
Most of the $R^2$ comes from the short and the long eruptions, which `waiting`
separates well.

## Classification accuracy {#classification-accuracy}

**Motivation.** Accuracy is the proportion of observations whose predicted class
is the observed class. It is easy to interpret, but it uses only the predicted
class, not the predictive probabilities. Balanced accuracy gives each class the
same weight. It prevents a model from scoring well by predicting only the most
frequent class.

**When to use.** Use them for binary or nominal outcomes. Prefer `bacc` when the
classes are unbalanced. Both need the class probabilities `mupred`: an
$S \times n$ matrix for binary outcomes, or an $S \times n \times C$ array for
$C$ classes.

**Key formula.** 
Let $\hat p_{ic} = \sum_s w_i^s\, p(\tilde y_i = c \mid \theta^s)$ be the
predicted probability of class $c$. For binary outcomes, the predicted class is
$\hat c_i = \mathbb{1}\{\hat p_{i1} > 0.5\}$. For $C$ classes, it is
$\hat c_i = \arg\max_c \hat p_{ic}$. The pointwise accuracy is
$\text{acc}_i = \mathbb{1}\{\hat c_i = y_i\}$.  With $n_c$ observations in
class $c$ (Vehtari et al., 2026, Eqs. 2.40-2.44), <!-- TODO -->

$$
\widehat{\text{acc}} = \frac{1}{n}\sum_{i=1}^n \text{acc}_i,
\qquad
\widehat{\text{bacc}} = \frac{1}{C}\sum_{c=1}^C \widehat{\text{acc}}_c,
\qquad
\widehat{\text{acc}}_c = \frac{1}{n_c}\sum_{i:\, y_i = c} \text{acc}_i .
$$

Both are utilities.

**Standard error.** Both use the binomial standard error, for balanced accuracy
within each class (Vehtari et al., 2026, Eqs. 2.43, 2.45): <!-- TODO -->

$$
\widehat{\text{SE}}(\widehat{\text{acc}}) =
  \sqrt{\frac{\widehat{\text{acc}}\,(1 - \widehat{\text{acc}})}{n}},
\qquad
\widehat{\text{SE}}(\widehat{\text{bacc}}) = \frac{1}{C}
  \sqrt{\sum_{c=1}^C \frac{\widehat{\text{acc}}_c\,(1 - \widehat{\text{acc}}_c)}{n_c}} .
$$

**Implementation.**

```{r accuracy}
species <- levels(fits_cat[[1]]$data$species)
y_cat <- as.integer(fits_cat[[1]]$data$species)
mu_cat <- lapply(fits_cat, brms::posterior_epred)

acc_cat <- lapply(mu_cat, function(mu) measure_acc(y = y_cat, mupred = mu))
bacc_cat <- lapply(mu_cat, function(mu) measure_bacc(y = y_cat, mupred = mu))

str(acc_cat$`with predictors`)
```

**Plot.** The models `fits_cat` predict `species`, the species of each of 342
penguins (`penguins` data). The plot shows the confusion matrix of each model.
Each row is an observed class, and each column is a predicted class.
Each cell gives the number of observations and their percentage of the row.
The cells on the diagonal are the correct classifications.
The other cells are the false classifications.

The two measures read the diagonal in different ways:

+ **Accuracy** is the sum of the diagonal counts divided by $n$. Each
  observation has the same weight, so a large class dominates.
+ **Balanced accuracy** is the mean of the diagonal percentages. Each class
  has the same weight, so a small class counts as much as a large class.

```{r accuracy-plot, echo = FALSE}
df_cm <- do.call(rbind, lapply(names(mu_cat), function(m) {
  pred <- max.col(apply(mu_cat[[m]], c(2, 3), mean))
  tab <- table(observed = factor(species[y_cat], species),
               predicted = factor(species[pred], species))
  data.frame(model = m, as.data.frame(tab), prop = as.vector(prop.table(tab, 1)))
}))
chin_gentoo <- with(df_cm, prop[model == "with predictors" &
                                observed == "Chinstrap" & predicted == "Gentoo"])
est_acc <- sapply(acc_cat, function(r) r$estimates[, "Estimate"])
est_bacc <- sapply(bacc_cat, function(r) r$estimates[, "Estimate"])

ggplot(df_cm, aes(predicted, observed, fill = prop)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%d\n(%.0f%%)", Freq, 100 * prop)), size = 3) +
  scale_fill_gradient(low = "grey95", high = "steelblue",
                      limits = c(0, 1), labels = scales::percent) +
  scale_y_discrete(limits = rev) +
  coord_equal() +
  facet_wrap(~model) +
  labs(x = "predicted species", y = "observed species",
       fill = "share of the\nobserved class")
```

**Interpretation.** The intercept-only model predicts Adelie, the most
frequent species, for every penguin. All its observations are in the first
column. Its diagonal percentages are 100%, 0%, and 0%. Its accuracy is
`r round(est_acc[["intercept only"]], 2)`, the share of Adelie. Its balanced
accuracy is `r round(est_bacc[["intercept only"]], 2)`, the mean of the three
percentages. This is the value of a random guess among three classes.

The model with predictors classifies most Adelie and Gentoo penguins
correctly. It predicts Gentoo for `r round(100 * chin_gentoo)`% of the
Chinstrap penguins, because their bills are about as long. Its accuracy is
`r round(est_acc[["with predictors"]], 2)`, and its balanced accuracy is
`r round(est_bacc[["with predictors"]], 2)`. Chinstrap has only 68 of 342
penguins, so the accuracy hides this failure. The balanced accuracy gives
Chinstrap the same weight as the other species, so it is lower.

## References

Bolin, D., and Wallin, J. (2023). Local scale invariance and robustness of
proper scoring rules. *Statistical Science*, 38(1):140–159.

Epstein, E. S. (1969). A scoring system for probability forecasts of ranked
categories. *Journal of Applied Meteorology*, 8, 985–987.

Gelman, A., Goodrich, B., Gabry, J., and Vehtari, A. (2019). R-squared for
Bayesian regression models. *The American Statistician*, 73(3):307–309.
doi:10.1080/00031305.2018.1549100

Gneiting, T. (2011). Making and evaluating point forecasts. *Journal of the
American Statistical Association*, 106(494):746–762.
doi:10.1198/jasa.2011.r10138

Hodson, T. O. (2022). Root-mean-square error (RMSE) or mean absolute error
(MAE): when to use them or not. *Geoscientific Model Development*,
15(14):5481–5487. doi:10.5194/gmd-15-5481-2022

Vehtari, A., Axen, S., Weber, F., and Bockting, F. (2026). Predictive measures
and model comparison in the loo package: A formula reference. <!-- TODO -->

## See also

- [Model comparison: Explanation of `model_compare()`](model-comparison.html)
- [Model comparison: Case Study with `model_compare()`]() <!-- TODO -->
- [Predictive model performance: Predictive schemes with `pred_measure()`](pred-measure-workflow.html)
- [Predictive measures and model comparison in the loo package: A formula reference]() <!-- TODO -->

## Appendix: model code {#appendix-model-code}

```{r fit-models-code, ref.label="fit-models", eval=FALSE}
```
