---
title: "Overview of predictive measures"
author: "Florence Bockting"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    css:
      - !expr system.file("rmarkdown/templates/html_vignette/resources/vignette.css", package = "rmarkdown")
      - styles.css
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Overview of predictive measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="../children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = params$EVAL,
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

```{r, child="../children/LOAD-BRMS-GITHUB.txt", eval=params$EVAL}
```

## Introduction

A *predictive measure* summarizes how well a model predicts data. **loo**
provides twelve built-in measures and the possibility to create a custom
measure. Each built-in measure has its own `measure_*()` function.

This vignette introduces each built-in measure. For each measure it gives:

1. the motivation,
2. when to use it,
3. the key formula and standard error,
4. the call in **loo**,
5. a plot and its interpretation, comparing a model with predictors to an intercept-only model.

This vignette states only the key formulas. The full definitions, the
estimators, and the standard errors are in our supplement 
(TODO-Vehtari et al., 2026).

This vignette hides the code for the plots. The full source code is in the
[`loo` GitHub repository](https://github.com/stan-dev/loo/blob/master/vignettes/articles-online-only/overview-measures.Rmd).

::: {.callout .callout-note}
**What this article does not cover**

+ **Predictive schemes**
  + Examples in this vignette use in-sample predictions.
  + All measures can be used with leave-one-out, K-fold, or test-data predictions. The use of these variants is introduced in [TODO-article]
+ **Model comparison** 
  + How to use the `loo::model_compare` function? See the [model comparison](model-comparison.html)
  vignette.
  + How to interpret the output of `model_compare`? See the tutorial [model-comparison TODO]
:::

## Overview of built-in measures

| Keyword | Measure | Family | Input | Outcome | Natural scale | Aggregation |
| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
| `elpd` | expected log pointwise predictive density | [log score](#log-score) | `ylp` | any | utility | sum |
| `mlpd` | mean log predictive density | [log score](#log-score) | `ylp` | any | utility | mean |
| `ic` | information criterion, $-2\,\text{elpd}$ | [log score](#log-score) | `ylp` | any | loss | sum |
| `rps` | (continuous) ranked probability score | [ranked probability score](#ranked-probability-score) | `ypred` | continuous or ordered | loss | mean |
| `srps` | scaled (continuous) ranked probability score | [ranked probability score](#ranked-probability-score) | `ypred` | continuous or ordered | utility | mean |
| `brier` | Brier score | [Brier score](#brier-score) | `ypred` | binary | loss | mean |
| `mae` | mean absolute error | [point error](#point-error) | `mupred` | continuous | loss | mean |
| `mse` | mean squared error | [point error](#point-error) | `mupred` | continuous | loss | mean |
| `rmse` | root mean squared error | [point error](#point-error) | `mupred` | continuous | loss | function of means |
| `r2` | predictive $R^2$ | [predictive $R^2$](#predictive-r2) | `mupred` | continuous | utility | function of means |
| `acc` | classification accuracy | [classification accuracy](#classification-accuracy) | `mupred` | binary or nominal | utility | mean |
| `bacc` | balanced accuracy | [classification accuracy](#classification-accuracy) | `mupred` | binary or nominal | utility | function of means |

The columns mean:

+ **Natural scale:** A *loss* is better when it is lower. A *utility* is better
  when it is higher. `measure_*()` reports each measure on its natural scale.
+ **Aggregation:** Most measures are a sum or a mean of pointwise terms. RMSE,
  $R^2$, and balanced accuracy are functions of means. RMSE and $R^2$ have no
  pointwise terms that add up to the estimate. The pointwise terms of balanced
  accuracy are weighted by class, so that their sum is the estimate. For more
  details see [TODO-article-suppl]

## Notation

This article uses one notation for all measures. The table maps each symbol to an
argument of `measure_*()` or to a field of its result.

| Symbol | Meaning | In **loo** |
| :--- | :--- | :--- |
| $y_i$ | observed outcome $i$ | `y` |
| $\theta^s$ | posterior draw $s$, $s = 1, \ldots, S$ | rows of the input matrices |
| $\ell_i^s = \log p(y_i \mid \theta^s)$ | log density of $y_i$ under draw $s$ | `ylp`, an $S \times n$ matrix |
| $\tilde y_i^s$ | posterior predictive draw | `ypred`, an $S \times n$ matrix; for `brier`, the event probabilities $p(\tilde y_i = 1 \mid \theta^s)$ |
| $\mathbb{E}[\tilde y_i \mid \theta^s]$ | predictive mean under draw $s$ | `mupred`, an $S \times n$ matrix, or $S \times n \times C$ for $C$ classes |
| $w_i^s$ | normalized weight of draw $s$ (Vehtari et al., 2026, Eq. 2.5) | `log_weights`, unnormalized; **loo** normalizes them. `NULL` gives $w_i^s = 1/S$ |
| $\text{elpd}_i$, $\text{rps}_i$, $\text{ae}_i$, … | pointwise term of a measure, on its natural scale | `$pointwise` |
| $\widehat{\text{elpd}}$, $\widehat{\text{rps}}$, $\widehat{\text{mae}}$, … | estimate of a measure | `$estimates[, "Estimate"]` |
| $\widehat{\text{SE}}$ | standard error of the estimate | `$estimates[, "SE"]` |
| $\widehat{\text{Var}}$ | sample variance over the observations | |

::: {.callout .callout-note}
**Weights**

Every estimator in this article uses weights $w_i^s$. With
`log_weights = NULL`, all draws get the weight $1/S$. This gives in-sample
predictions. Other weights give other predictive schemes, for example PSIS-LOO.
:::

## Data and models

```{r load-libraries}
library(loo)
library(ggplot2)
theme_set(theme_minimal())
```

For each outcome type we fit two `brms` models:

+ **with predictors**: a model that uses the predictors of the data,
+ **intercept only**: a model that predicts the same distribution for every
  observation.

The intercept-only model is a baseline. It shows how a measure changes when a
model uses no information from the predictors. The model with predictors is
better, but not necessarily good. Fitting is not the topic of this article. The
model code is in the [appendix](#appendix-model-code).

| Object | Data | Outcome | Predictors | Used for |
| :--- | :--- | :--- | :--- | :--- |
| `fits_count` | `roaches` (`rstanarm`) | count, Poisson | `sqrt_roach1`, `treatment`, `senior` | log score |
| `fits_cont` | `faithful` (`datasets`) | continuous, normal | `waiting` | CRPS, point error, $R^2$ |
| `fits_bin` | `wells` (`rstanarm`) | binary, Bernoulli | `dist100`, `arsenic` | Brier score |
| `fits_ord` | simulated | ordered, 4 categories | `x` | RPS |
| `fits_cat` | simulated | nominal, 3 unbalanced classes | `x` | accuracy |

Each object is a list with the elements `"with predictors"` and
`"intercept only"`.

```{r fit-models, echo=FALSE, results="hide"}
SEED <- 42

fit_pair <- function(formula, data, family) {
  fit <- brms::brm(formula, data = data, family = family,
                   chains = 2, iter = 1000, seed = SEED, refresh = 0)
  list(
    "with predictors" = fit,
    "intercept only" = update(fit, formula. = . ~ 1, seed = SEED, refresh = 0)
  )
}

# count
data(roaches, package = "rstanarm")
roaches$sqrt_roach1 <- sqrt(roaches$roach1)
fits_count <- fit_pair(y ~ sqrt_roach1 + treatment + senior, roaches, poisson())

# continuous
fits_cont <- fit_pair(eruptions ~ waiting, faithful, gaussian())

# binary
data(wells, package = "rstanarm")
wells$dist100 <- wells$dist / 100
fits_bin <- fit_pair(switch ~ dist100 + arsenic, wells, brms::bernoulli())

# ordered: latent logistic model with three cut points
set.seed(SEED)
n_sim <- 300
x <- rnorm(n_sim)
latent <- 1.2 * x + rlogis(n_sim)
df_ord <- data.frame(y = 1 + rowSums(outer(latent, c(-1, 0.3, 1.5), ">")), x = x)
fits_ord <- fit_pair(y ~ x, df_ord, brms::cumulative("logit"))

# nominal: class 1 is the most frequent
set.seed(SEED + 1)
x <- rnorm(n_sim)
prob <- cbind(1, exp(-3.5 + 5 * x), exp(-5.5 - 5 * x))
prob <- prob / rowSums(prob)
df_cat <- data.frame(y = apply(prob, 1, function(p) sample(3, 1, prob = p)), x = x)
fits_cat <- fit_pair(y ~ x, df_cat, brms::categorical())
```

**Prediction tasks.** The models predict the following outcomes:

| Model | Outcome | Predictors | N |
|:--|:--|:--|:--|
| Roach counts (`fits_count`) | `y`: roaches caught in traps after the treatment | `sqrt_roach1`: square root of the count before the treatment; `treatment`: the building received the treatment; `senior`: only elderly people live in the building | 262 buildings |
| Geyser eruptions (`fits_cont`) | `eruptions`: duration of an eruption of the Old Faithful geyser, in minutes | `waiting`: waiting time since the previous eruption, in minutes | 272 eruptions |
| Well switching (`fits_bin`) | `switch`: 1 if the household switched to a safe well | `dist100`: distance to the nearest safe well, in 100 m; `arsenic`: arsenic level of the current well | 3020 households |
| Ordered rating (`fits_ord`) | `y`: rating on an ordered scale from 1 to 4 | `x`: higher values make higher ratings more likely | 300 (simulated) |
| Class membership (`fits_cat`) | `y`: one of three unordered classes (class 1: `r sum(df_cat$y == 1)`, class 2: `r sum(df_cat$y == 2)`, class 3: `r sum(df_cat$y == 3)`) | `x`: shifts the probabilities between the classes | 300 (simulated) |

## Log score {#log-score}

**Motivation.** The log score evaluates the predictive density at the observed
value. A model gets a high score when it puts much probability near the
observation. It uses the whole predictive distribution, and it strongly
penalizes observations the model considers very unlikely.

**When to use.** The log score works for any outcome type. It is the default
measure in **loo**. It needs the log-likelihood draws `ylp`.

**Key formula.** The pointwise term and the estimate are
(Vehtari et al., 2026, Eqs. 2.16–2.18)

$$
\text{elpd}_i = \log \sum_{s=1}^S \exp\!\left(\log w_i^s + \ell_i^s\right),
\qquad
\widehat{\text{elpd}} = \sum_{i=1}^n \text{elpd}_i .
$$

**loo** reports two rescaled versions: 

+ $\widehat{\text{mlpd}} =
\widehat{\text{elpd}} / n$ on the mean scale, and 
+ the information criterion
$\widehat{\text{ic}} = -2\,\widehat{\text{elpd}}$.

**Standard error.** The elpd is a sum, the mlpd is a mean
(Vehtari et al., 2026, Eqs. 2.8, 2.9):

$$
\widehat{\text{SE}}(\widehat{\text{elpd}}) =
  \sqrt{n\,\widehat{\text{Var}}(\text{elpd}_i)},
\qquad
\widehat{\text{SE}}(\widehat{\text{mlpd}}) =
  \sqrt{\widehat{\text{Var}}(\text{elpd}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{ic}}) =
  2\,\widehat{\text{SE}}(\widehat{\text{elpd}}) .
$$

**Implementation.**

```{r log-score}
ylp_count <- lapply(fits_count, brms::log_lik)

elpd_count <- lapply(ylp_count, function(ylp) measure_elpd(ylp = ylp))
mlpd_count <- lapply(ylp_count, function(ylp) measure_mlpd(ylp = ylp))
ic_count <- lapply(ylp_count, function(ylp) measure_ic(ylp = ylp))

# output structure of elpd measure for model with predictors
str(elpd_count$`with predictors`)
```

**Plot.** Each point is one building. The y-axis shows its pointwise log score
$\text{elpd}_i$. The dashed line is the mlpd, the mean of the points. Higher is
better.

```{r log-score-plot, echo = FALSE}
df_elpd <- do.call(rbind, lapply(names(ylp_count), function(m) {
  data.frame(model = m, y = roaches$y, elpd_i = as.vector(elpd_count[[m]]$pointwise))
}))
df_lines <- data.frame(
  model = names(ylp_count),
  elpd = sapply(elpd_count, function(r) r$estimates[, "Estimate"]),
  mlpd = sapply(mlpd_count, function(r) r$estimates[, "Estimate"]),
  ic = sapply(ic_count, function(r) r$estimates[, "Estimate"])
)

ggplot(df_elpd, aes(y, elpd_i)) +
  geom_point(alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = mlpd), linetype = "dashed") +
  geom_text(data = df_lines,
            aes(x = 0, y = -Inf,
                label = sprintf("elpd = %.0f\nic = %.0f\nmlpd = %.1f", elpd, ic, mlpd)),
            hjust = 0, vjust = -0.3, size = 3) +
  scale_x_continuous(trans = "pseudo_log", breaks = c(0, 1, 3, 10, 30, 100, 300)) +
  facet_wrap(~model) +
  labs(x = "observed roach count", y = "pointwise log score")

```

**Interpretation.** The intercept-only model predicts the same distribution for
every building, with a mean of about `r round(mean(roaches$y))` roaches. All
its points lie on one curve, with its peak near that mean. Buildings with no
roaches and buildings with many roaches both get low scores. The model with
predictors uses the count before the treatment. Its points no longer lie on one
curve. Most buildings, also those with no roaches, get a score near zero. In
both models, a few buildings with large counts get very low scores. Because the
elpd is a sum, these few buildings contribute much of the total.

## Ranked probability score {#ranked-probability-score}

**Motivation.** The ranked probability score compares the predictive
distribution with the observed value through absolute distances. Unlike the
log score, it rewards predictions that are *close* to the observation, not only
predictions that hit it. It is less sensitive to single outliers.

**When to use.** Use `rps` for ordered categorical outcomes (RPS) and for
continuous outcomes (CRPS). It is on the scale of $y$, so a value can be
read in the units of the outcome. Use `srps` when outcomes differ strongly in
their predictive spread. Both need the posterior predictive draws `ypred`.

**Key formula.** Let $X$ and $X'$ be independent draws from the predictive
distribution of observation $i$ (Vehtari et al., 2026, Eqs. 2.20–2.22):

$$
\text{rps}_i = \mathbb{E}|X - y_i| - \tfrac{1}{2}\,\mathbb{E}|X - X'|, \qquad
\text{srps}_i = -\frac{\mathbb{E}|X - y_i|}{\mathbb{E}|X - X'|}
  - \tfrac{1}{2}\log \mathbb{E}|X - X'| .
$$

The estimates are the means $\widehat{\text{rps}} = \frac{1}{n}\sum_i
\text{rps}_i$ and $\widehat{\text{srps}} = \frac{1}{n}\sum_i \text{srps}_i$.
**loo** estimates both expectations from one set of draws. The estimator of
$\mathbb{E}|X - X'|$ is a probability-weighted moment estimator
(Vehtari et al., 2026, Eq. 2.25). `rps` is a loss, and `srps` is a utility.

**Standard error.** Both are means (Vehtari et al., 2026, Eq. 2.8):

$$
\widehat{\text{SE}}(\widehat{\text{rps}}) =
  \sqrt{\widehat{\text{Var}}(\text{rps}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{srps}}) =
  \sqrt{\widehat{\text{Var}}(\text{srps}_i) / n} .
$$

**Implementation.** For the continuous outcome, `measure_rps()` gives the CRPS:

```{r crps}
y_cont <- fits_cont[[1]]$data$eruptions
ypred_cont <- lapply(fits_cont, brms::posterior_predict)

crps_cont <- lapply(ypred_cont, function(yp) measure_rps(y = y_cont, ypred = yp))
scrps_cont <- lapply(ypred_cont, function(yp) measure_srps(y = y_cont, ypred = yp))

# output structure of crps measure for model with predictors
str(crps_cont$`with predictors`)
```

For the ordered outcome, the same functions give the RPS and the SRPS:

```{r rps}
y_ord <- fits_ord[[1]]$data$y
ypred_ord <- lapply(fits_ord, brms::posterior_predict)

rps_ord <- lapply(ypred_ord, function(yp) measure_rps(y = y_ord, ypred = yp))
srps_ord <- lapply(ypred_ord, function(yp) measure_srps(y = y_ord, ypred = yp))

# output structure of srps measure for model with predictors
str(srps_ord$`with predictors`)
```

**Plot.** Each point is one eruption. The x-axis shows the distance of the
observed duration from the predictive mean. The y-axis shows the pointwise CRPS
$\text{rps}_i$, in minutes. The dashed line is the estimate, the mean of the
points. Lower is better.

```{r crps-plot, echo = FALSE}
df_crps <- do.call(rbind, lapply(names(ypred_cont), function(m) {
  data.frame(
    model = m,
    d = y_cont - colMeans(ypred_cont[[m]]),
    crps_i = as.vector(crps_cont[[m]]$pointwise)
  )
}))
df_lines <- data.frame(
  model = names(crps_cont),
  crps = sapply(crps_cont, function(r) r$estimates[, "Estimate"])
)

ggplot(df_crps, aes(d, crps_i)) +
  geom_point(alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = crps), linetype = "dashed") +
  facet_wrap(~model) +
  labs(x = "observed duration minus predictive mean (minutes)", y = "pointwise CRPS")
```

**Interpretation.** In both panels the CRPS is smallest at zero distance.
Near zero the curve is round. Far from zero the CRPS grows about linearly with
the distance, like an absolute error. The left side of the intercept-only panel
shows this. The log score, by contrast, drops fast for an unlikely observation.
The model with predictors has a narrower predictive distribution. Its curve is
lower at the bottom, and more of its points lie near zero distance.

**Plot for the ordered outcome.** Each point is one observation. The x-axis
shows the distance of the observed rating from the predictive mean. The y-axis
shows the pointwise RPS $\text{rps}_i$. The color shows the observed rating.
The dashed line is the estimate. Lower is better.

```{r rps-plot, echo = FALSE}
df_rps <- do.call(rbind, lapply(names(ypred_ord), function(m) {
  data.frame(
    model = m,
    d = y_ord - colMeans(ypred_ord[[m]]),
    rps_i = as.vector(rps_ord[[m]]$pointwise),
    rating = factor(y_ord)
  )
}))
df_lines <- data.frame(
  model = names(rps_ord),
  rps = sapply(rps_ord, function(r) r$estimates[, "Estimate"])
)

ggplot(df_rps, aes(d, rps_i)) +
  geom_point(aes(color = rating), alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = rps), linetype = "dashed") +
  scale_color_viridis_d(end = 0.9) +
  facet_wrap(~model) +
  labs(x = "observed rating minus predictive mean", y = "pointwise RPS",
       color = "observed rating")
```

**Interpretation.** The intercept-only model gives every observation the same
predictive distribution. Its points therefore fall on four spots, one for each
observed rating. The RPS uses the order of the categories: a rating far from
the predictive mean gets a higher RPS than a rating close to it. The model
with predictors shifts the predictive distribution with `x`. Its points form a
V, as for the CRPS. Its estimate is lower:
`r round(rps_ord[["with predictors"]]$estimates[, "Estimate"], 2)` against
`r round(rps_ord[["intercept only"]]$estimates[, "Estimate"], 2)`.

**SRPS.** The SRPS divides the expected distance to the observation by the
predictive spread $\mathbb{E}|X - X'|$. A narrow predictive distribution
therefore gets a large penalty when it misses the observation. The SRPS is a
utility, so higher is better. The table gives the estimates of both models.

```{r srps-table, echo = FALSE}
est <- function(res) sapply(res, function(r) r$estimates[, "Estimate"])
knitr::kable(
  rbind("eruption duration (SCRPS)" = est(scrps_cont), "ordered rating (SRPS)" = est(srps_ord)),
  digits = 2
)
```

## Brier score {#brier-score}

**Motivation.** The Brier score is the squared distance between the predicted
probability of an event and the observed outcome, 0 or 1. It rewards
probabilities that are both calibrated and sharp, that is, close to 0 or 1.

**When to use.** Use it for binary outcomes. **loo** implements the binary
Brier score only. `measure_brier()` needs the event probabilities of the
draws, passed as `ypred`.

**Key formula.** With the predicted event probability
$\hat p_i = \sum_s w_i^s\, p(\tilde y_i = 1 \mid \theta^s)$
(Vehtari et al., 2026, Eqs. 2.26–2.28),

$$
\text{bs}_i = \left(\hat p_i - y_i\right)^2, \qquad
\widehat{\text{bs}} = \frac{1}{n}\sum_{i=1}^n \text{bs}_i .
$$

The Brier score is a loss. It ranges from 0 to 1.

**Standard error.** The Brier score is a mean
(Vehtari et al., 2026, Eq. 2.8):

$$
\widehat{\text{SE}}(\widehat{\text{bs}}) =
  \sqrt{\widehat{\text{Var}}(\text{bs}_i) / n} .
$$

**Implementation.**

```{r brier}
y_bin <- fits_bin[[1]]$data$switch
p_bin <- lapply(fits_bin, brms::posterior_epred)
brier_bin <- lapply(p_bin, function(p) measure_brier(y = y_bin, ypred = p))

str(brier_bin$`with predictors`)
```

**Plot.** Each point is one household. The x-axis shows the predicted
probability $\hat p_i$ that it switches wells. Every point lies on one of two
curves: $\hat p_i^2$ if the household did not switch, and $(1 - \hat p_i)^2$
if it switched. The dashed line is the estimate, the mean of the points. Lower
is better.

```{r brier-plot, echo = FALSE}
df_brier <- do.call(rbind, lapply(names(p_bin), function(m) {
  data.frame(
    model = m,
    p = colMeans(p_bin[[m]]),
    bs_i = as.vector(brier_bin[[m]]$pointwise),
    switched = factor(y_bin, levels = 0:1, labels = c("no", "yes"))
  )
}))
df_lines <- data.frame(
  model = names(brier_bin),
  brier = sapply(brier_bin, function(r) r$estimates[, "Estimate"])
)

ggplot(df_brier, aes(p, bs_i, color = switched)) +
  geom_function(fun = function(p) p^2, color = "grey70", linetype = "dotted") +
  geom_function(fun = function(p) (1 - p)^2, color = "grey70", linetype = "dotted") +
  geom_point(alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = brier), linetype = "dashed") +
  facet_wrap(~model) +
  labs(x = "predicted probability of switching",
       y = "pointwise Brier score", color = "switched")
```

**Interpretation.** The intercept-only model gives every household the same
probability, the proportion of households that switched. All its points fall
on two spots, one on each curve. The model with predictors spreads the
probabilities. A household moves down its curve when the model is confident
and right. It moves up its curve when the model is confident and wrong.

## Point error {#point-error}

**Motivation.** The point-error measures reduce the predictive distribution to
one point prediction, the predictive mean. They then measure the distance
between that prediction and the observed value. They are easy to interpret,
but they ignore the predictive uncertainty.

**When to use.** Use them for continuous outcomes when the quality of the point
prediction matters. MAE is less sensitive to large errors than MSE and RMSE.
MAE and RMSE are on the scale of $y$. All three need the predictive means
`mupred`.

**Which point prediction.** Each loss rewards a different point prediction.
The squared error is smallest in expectation for the predictive mean. The
absolute error is smallest for the predictive median (Gneiting, 2011).
**loo** uses the predictive mean for all three measures. MSE and RMSE
therefore evaluate the prediction they reward. MAE evaluates the mean, not
the median. In the normal model here, the predictive distribution is close to
symmetric. Its mean and its median are then about equal.

**MAE or RMSE.** Neither is better in general. RMSE fits normal errors, and
MAE fits Laplace errors with heavier tails (Hodson, 2022). RMSE is never
smaller than MAE. They are equal only when all absolute errors are equal.

**Key formula.** With the point prediction
$\hat y_i = \sum_s w_i^s\, \mathbb{E}[\tilde y_i \mid \theta^s]$, the pointwise
errors are $\text{ae}_i = |y_i - \hat y_i|$ and
$\text{sqe}_i = (y_i - \hat y_i)^2$ (Vehtari et al., 2026, Eqs. 2.29–2.34):

$$
\widehat{\text{mae}} = \frac{1}{n}\sum_{i=1}^n \text{ae}_i, \qquad
\widehat{\text{mse}} = \frac{1}{n}\sum_{i=1}^n \text{sqe}_i, \qquad
\widehat{\text{rmse}} = \sqrt{\widehat{\text{mse}}} .
$$

All three are losses.

**Standard error.** MAE and MSE are means (Vehtari et al., 2026, Eq. 2.8).
RMSE is a function of a mean, so its standard error uses the delta method
(Vehtari et al., 2026, Eq. 2.33):

$$
\widehat{\text{SE}}(\widehat{\text{mae}}) =
  \sqrt{\widehat{\text{Var}}(\text{ae}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{mse}}) =
  \sqrt{\widehat{\text{Var}}(\text{sqe}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{rmse}}) =
  \frac{\widehat{\text{SE}}(\widehat{\text{mse}})}{2\,\widehat{\text{rmse}}} .
$$

**Pointwise terms.** `measure_rmse()` has no pointwise terms of its own. Its
`$pointwise` field holds the squared errors $\text{sqe}_i$, the same as
`measure_mse()`.

**Implementation.**

```{r point-error}
mu_cont <- lapply(fits_cont, brms::posterior_epred)

m_mae <- lapply(mu_cont, function(mu) measure_mae(y = y_cont, mupred = mu))
m_mse <- lapply(mu_cont, function(mu) measure_mse(y = y_cont, mupred = mu))
m_rmse <- lapply(mu_cont, function(mu) measure_rmse(y = y_cont, mupred = mu))

str(m_mse$`with predictors`)
```

**Plot.** Each point is one eruption. The x-axis shows the error of the point
prediction, $y_i - \hat y_i$. The y-axis shows the pointwise squared error
$\text{sqe}_i$. The dashed line is the MSE, the mean of the points. Lower is
better.

```{r point-error-plot, echo = FALSE}
df_pe <- do.call(rbind, lapply(names(mu_cont), function(m) {
  data.frame(
    model = m,
    e = y_cont - colMeans(mu_cont[[m]]),
    sqe_i = as.vector(m_mse[[m]]$pointwise)
  )
}))
df_lines <- data.frame(
  model = names(mu_cont),
  mae = sapply(m_mae, function(r) r$estimates[, "Estimate"]),
  mse = sapply(m_mse, function(r) r$estimates[, "Estimate"]),
  rmse = sapply(m_rmse, function(r) r$estimates[, "Estimate"])
)
top_share <- sapply(m_mse, function(r) {
  sqe <- sort(r$pointwise, decreasing = TRUE)
  sum(sqe[seq_len(ceiling(0.1 * length(sqe)))]) / sum(sqe)
})

ggplot(df_pe, aes(e, sqe_i)) +
  geom_point(alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = mse), linetype = "dashed") +
  geom_text(data = df_lines,
            aes(x = -Inf, y = Inf,
                label = sprintf("MSE = %.2f\nRMSE = %.2f", mse, rmse)),
            hjust = -0.1, vjust = 1.2, size = 3) +
  facet_wrap(~model) +
  labs(x = "observed duration minus point prediction (minutes)",
       y = "pointwise squared error")
```

**Interpretation.** The points lie on a parabola. Compare the left side of
the intercept-only panel with the CRPS plot: the squared error grows with the
square of the distance, not linearly. Most
eruptions lie below the dashed line, and a few large errors pull the MSE up.
For the model with predictors, the 10% largest errors make up
`r round(100 * top_share[["with predictors"]])`% of the MSE. The MSE is in
squared minutes. The RMSE, its square root, is back in minutes.

**Plot for MAE and RMSE.** The histogram shows the pointwise absolute errors
$\text{ae}_i$. The dashed line is the MAE. The dotted line is the RMSE. Both
are in minutes.

```{r point-error-mae-rmse-plot, echo = FALSE}
df_ae <- do.call(rbind, lapply(names(mu_cont), function(m) {
  data.frame(model = m, ae_i = as.vector(m_mae[[m]]$pointwise))
}))
df_ref <- rbind(
  data.frame(model = df_lines$model, measure = "MAE", value = df_lines$mae),
  data.frame(model = df_lines$model, measure = "RMSE", value = df_lines$rmse)
)

ggplot(df_ae, aes(ae_i)) +
  geom_histogram(binwidth = 0.1, boundary = 0, fill = "grey70", color = "white") +
  geom_vline(data = df_ref, aes(xintercept = value, linetype = measure)) +
  geom_text(data = df_ref, aes(x = value, y = Inf, label = measure),
            angle = 90, vjust = -0.4, hjust = 1.1, size = 3) +
  scale_linetype_manual(values = c(MAE = "dashed", RMSE = "dotted")) +
  facet_wrap(~model) +
  labs(x = "pointwise absolute error (minutes)",
       y = "number of eruptions", linetype = NULL)
```

**Interpretation.** In both models the RMSE lies to the right of the MAE. The
squared error gives more weight to the largest errors. The
model with predictors moves the histogram to the left, so both measures
decrease. Its MAE is
`r sprintf("%.2f", m_mae[["with predictors"]]$estimates[, "Estimate"])` minutes and
its RMSE is
`r sprintf("%.2f", m_rmse[["with predictors"]]$estimates[, "Estimate"])` minutes.

## Predictive $R^2$ {#predictive-r2}

**Motivation.** The predictive $R^2$ compares the squared error of the model
with the squared error of the simplest prediction, the mean of the
observations. A value of 0 means the model predicts no better than the mean.
A negative value means the model predicts worse than the mean.

**When to use.** Use it for continuous outcomes, when a unitless summary of the
point prediction is needed. It is a predictive version of the classical $R^2$,
not the Bayesian $R^2$ of Gelman et al. (2019). It needs `mupred`.

**Key formula.** With $\text{sqe}_i$ and $\widehat{\text{mse}}$ from the
[point error](#point-error) section (Vehtari et al., 2026, Eqs. 2.35, 2.36),

$$
\widehat{R^2} = 1 - \frac{\widehat{\text{mse}}}{\widehat{\text{mse}}_y},
\qquad
\widehat{\text{mse}}_y = \frac{1}{n}\sum_{i=1}^n (y_i - \bar y)^2 .
$$

$R^2$ is a utility.

**Standard error.** $R^2$ is a function of two means, so its standard error
uses the delta method. With $c = \widehat{\text{mse}} /
\widehat{\text{mse}}_y$ (Vehtari et al., 2026, Eq. 2.39),

$$
\widehat{\text{SE}}(\widehat{R^2}) = \frac{1}{\widehat{\text{mse}}_y}
  \sqrt{\widehat{\text{Var}}\!\left(\text{sqe}_i - c\,(y_i - \bar y)^2\right) / n} .
$$

**Pointwise terms.** $R^2$ has no pointwise terms of its own. The
`$pointwise` field of `measure_r2()` holds the squared errors
$\text{sqe}_i$ of the model.

**Implementation.**

```{r r2}
m_r2 <- lapply(mu_cont, function(mu) measure_r2(y = y_cont, mupred = mu))

str(m_r2$`with predictors`)
```

**Plot.** Each point is one eruption. The x-axis shows the distance of the
observed duration from the mean duration. The y-axis shows the contribution of
the eruption to $R^2$,

$$
r_i = \frac{(y_i - \bar y)^2 - \text{sqe}_i}{\widehat{\text{mse}}_y},
\qquad
\frac{1}{n}\sum_{i=1}^n r_i = \widehat{R^2} .
$$

A point above zero is an eruption for which the model predicts better than
the mean. The dashed line is $\widehat{R^2}$, the mean of the points. Higher is
better.

```{r r2-plot, echo = FALSE}
mse_y <- mean((y_cont - mean(y_cont))^2)
df_r2 <- do.call(rbind, lapply(names(m_r2), function(m) {
data.frame(
    model = m,
    d = y_cont - mean(y_cont),
    r_i = ((y_cont - mean(y_cont))^2 - as.vector(m_mse[[m]]$pointwise)) / mse_y
  )
}))
df_lines <- data.frame(
  model = names(m_r2),
  r2 = sapply(m_r2, function(r) r$estimates[, "Estimate"])
)

ggplot(df_r2, aes(d, r_i)) +
  geom_hline(yintercept = 0, color = "grey70") +
  geom_point(alpha = 0.5) +
  geom_hline(data = df_lines, aes(yintercept = r2), linetype = "dashed") +
  geom_text(data = df_lines,
            aes(x = -Inf, y = Inf, label = sprintf("R2 = %.2f", r2)),
            hjust = -0.1, vjust = 1.2, size = 3) +
  facet_wrap(~model) +
  labs(x = "observed duration minus mean duration (minutes)",
       y = "contribution to R2")
```

**Interpretation.** The intercept-only model predicts about the mean duration
for every eruption. All its contributions are close to zero, and so is its
$R^2$.
For the model with predictors, the points lie on or below a parabola. The
contribution of an
eruption is never larger than $(y_i - \bar y)^2 / \widehat{\text{mse}}_y$. An
eruption close to the mean duration can therefore gain little: the mean already
predicts it well. An eruption far from the mean can gain much. Any eruption
loses much when its point prediction is far from its duration. Most eruptions
are either short or long, so most points lie far from the mean. The waiting
time predicts these eruptions well. This gives the high $R^2$.

## Classification accuracy {#classification-accuracy}

**Motivation.** Accuracy is the proportion of observations whose predicted class
is the observed class. It is easy to interpret, but it uses only the predicted
class, not the predictive probabilities. Balanced accuracy gives each class the
same weight. It prevents a model from scoring well by predicting only the most
frequent class.

**When to use.** Use them for binary or nominal outcomes. Prefer `bacc` when the
classes are unbalanced. Both need the class probabilities `mupred`: an
$S \times n$ matrix for binary outcomes, or an $S \times n \times C$ array for
$C$ classes.

**Key formula.** The predicted class is
$\hat c_i = \arg\max_c \sum_s w_i^s\, p(\tilde y_i = c \mid \theta^s)$, and the
pointwise accuracy is $\text{acc}_i = \mathbb{1}\{\hat c_i = y_i\}$. With $n_c$
observations in class $c$ (Vehtari et al., 2026, Eqs. 2.40–2.44),

$$
\widehat{\text{acc}} = \frac{1}{n}\sum_{i=1}^n \text{acc}_i,
\qquad
\widehat{\text{bacc}} = \frac{1}{C}\sum_{c=1}^C \widehat{\text{acc}}_c,
\qquad
\widehat{\text{acc}}_c = \frac{1}{n_c}\sum_{i:\, y_i = c} \text{acc}_i .
$$

Both are utilities.

**Pointwise terms.** The `$pointwise` field of `measure_acc()` holds
$\text{acc}_i$. The `$pointwise` field of `measure_bacc()` holds
$\text{acc}_i / (C\, n_c)$, with $c$ the class of observation $i$. These terms
sum to $\widehat{\text{bacc}}$.

**Standard error.** Both use the binomial standard error, for balanced accuracy
within each class (Vehtari et al., 2026, Eqs. 2.43, 2.45):

$$
\widehat{\text{SE}}(\widehat{\text{acc}}) =
  \sqrt{\frac{\widehat{\text{acc}}\,(1 - \widehat{\text{acc}})}{n}},
\qquad
\widehat{\text{SE}}(\widehat{\text{bacc}}) = \frac{1}{C}
  \sqrt{\sum_{c=1}^C \frac{\widehat{\text{acc}}_c\,(1 - \widehat{\text{acc}}_c)}{n_c}} .
$$

**Implementation.**

```{r accuracy}
y_cat <- fits_cat[[1]]$data$y
mu_cat <- lapply(fits_cat, brms::posterior_epred)

acc_cat <- lapply(mu_cat, function(mu) measure_acc(y = y_cat, mupred = mu))
bacc_cat <- lapply(mu_cat, function(mu) measure_bacc(y = y_cat, mupred = mu))

str(acc_cat$`with predictors`)
```

**Plot.** The plot shows the confusion matrix of each model. Each row is an
observed class, and each column is a predicted class. Each cell gives the
number of observations and their percentage of the row. The cells on the
diagonal are the correct classifications. The other cells are the false
classifications.

The two measures read the diagonal in different ways:

+ **Accuracy** is the sum of the diagonal counts divided by $n$. Each
  observation has the same weight, so a large class dominates.
+ **Balanced accuracy** is the mean of the diagonal percentages. Each class
  has the same weight, so a small class counts as much as a large class.

```{r accuracy-plot, echo = FALSE}
df_cm <- do.call(rbind, lapply(names(mu_cat), function(m) {
  pred <- max.col(apply(mu_cat[[m]], c(2, 3), mean))
  tab <- table(observed = factor(y_cat, 1:3), predicted = factor(pred, 1:3))
  data.frame(model = m, as.data.frame(tab), prop = as.vector(prop.table(tab, 1)))
}))
est_acc <- sapply(acc_cat, function(r) r$estimates[, "Estimate"])
est_bacc <- sapply(bacc_cat, function(r) r$estimates[, "Estimate"])

ggplot(df_cm, aes(predicted, observed, fill = prop)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%d\n(%.0f%%)", Freq, 100 * prop)), size = 3) +
  scale_fill_gradient(low = "grey95", high = "steelblue",
                      limits = c(0, 1), labels = scales::percent) +
  scale_y_discrete(limits = rev) +
  coord_equal() +
  facet_wrap(~model) +
  labs(x = "predicted class", y = "observed class",
       fill = "share of the\nobserved class")
```

**Interpretation.** The intercept-only model predicts class 1, the most
frequent class, for every observation. All its observations are in the first
column. Its diagonal percentages are 100%, 0%, and 0%. Its accuracy is
`r round(est_acc[["intercept only"]], 2)`, the share of class 1. Its balanced
accuracy is `r round(est_bacc[["intercept only"]], 2)`, the mean of the three
percentages. This is the value of a random guess among three classes.

The model with predictors also classifies part of classes 2 and 3 correctly.
Its accuracy is `r round(est_acc[["with predictors"]], 2)`, and its balanced
accuracy is `r round(est_bacc[["with predictors"]], 2)`. The balanced accuracy
shows the improvement in the small classes more clearly than the accuracy.

## References

Gelman, A., Goodrich, B., Gabry, J., and Vehtari, A. (2019). R-squared for
Bayesian regression models. *The American Statistician*, 73(3):307–309.
doi:10.1080/00031305.2018.1549100

Gneiting, T. (2011). Making and evaluating point forecasts. *Journal of the
American Statistical Association*, 106(494):746–762.
doi:10.1198/jasa.2011.r10138

Hodson, T. O. (2022). Root-mean-square error (RMSE) or mean absolute error
(MAE): when to use them or not. *Geoscientific Model Development*,
15(14):5481–5487. doi:10.5194/gmd-15-5481-2022

Vehtari, A., Axen, S., Weber, F., and Bockting, F. (2026). Predictive measures
and model comparison in the loo package: A formula reference. [TODO: arXiv ID]

## See also

- [Model comparison](model-comparison.html) — differences between models and
  their standard errors.
- [TODO: schemes and `pred_measure()` article]

## Appendix: model code {#appendix-model-code}

```{r fit-models-code, ref.label="fit-models", eval=FALSE}
```
