---
title: "Overview of predictive measures"
author: "Florence Bockting"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    css:
      - !expr system.file("rmarkdown/templates/html_vignette/resources/vignette.css", package = "rmarkdown")
      - styles.css
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Overview of predictive measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="../children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = params$EVAL,
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

```{r, child="../children/LOAD-BRMS-GITHUB.txt", eval=params$EVAL}
```

## Introduction

A *predictive measure* summarizes how well a model predicts data. **loo**
implements twelve of them. Each one has its own `measure_*()` function.

This article introduces each measure. For each measure it gives:

1. the motivation: what the measure rewards,
2. when to use it: the outcome type and the required input,
3. the key formula,
4. the standard error,
5. the call in **loo**,
6. a plot and its interpretation, for a model that performs well and a model
   that performs badly.

The article states only the key formulas. The full definitions, the
estimators, and the standard errors are in the formula reference
(Vehtari et al., 2026). Equation numbers in this article refer to that paper.

::: {.callout .callout-note}
**What this article does not cover**

+ **Predictive schemes.** All examples use in-sample predictions. In-sample
  estimates are optimistic for new data. The same measure can use leave-one-out,
  K-fold, or test-data predictions instead. [TODO: link to the schemes and
  `pred_measure()` article.]
+ **Model comparison.** See the [model comparison](model-comparison.html)
  article.
:::

## Overview of the measures

| Keyword | Measure | Family | Input | Outcome | Natural scale | Aggregation |
| :--- | :--- | :--- | :--- | :--- | :--- | :--- |
| `elpd` | expected log pointwise predictive density | [log score](#log-score) | `ylp` | any | utility | sum |
| `mlpd` | mean log predictive density | [log score](#log-score) | `ylp` | any | utility | mean |
| `ic` | information criterion, $-2\,\text{elpd}$ | [log score](#log-score) | `ylp` | any | loss | sum |
| `rps` | (continuous) ranked probability score | [ranked probability score](#ranked-probability-score) | `ypred` | continuous or ordered | loss | mean |
| `srps` | scaled (continuous) ranked probability score | [ranked probability score](#ranked-probability-score) | `ypred` | continuous or ordered | utility | mean |
| `brier` | Brier score | [Brier score](#brier-score) | `ypred` | binary | loss | mean |
| `mae` | mean absolute error | [point error](#point-error) | `mupred` | continuous | loss | mean |
| `mse` | mean squared error | [point error](#point-error) | `mupred` | continuous | loss | mean |
| `rmse` | root mean squared error | [point error](#point-error) | `mupred` | continuous | loss | function of means |
| `r2` | predictive $R^2$ | [predictive $R^2$](#predictive-r2) | `mupred` | continuous | utility | function of means |
| `acc` | classification accuracy | [classification accuracy](#classification-accuracy) | `mupred` | binary or nominal | utility | mean |
| `bacc` | balanced accuracy | [classification accuracy](#classification-accuracy) | `mupred` | binary or nominal | utility | function of means |

The columns mean:

+ **Natural scale.** A *loss* is better when it is lower. A *utility* is better
  when it is higher. `measure_*()` reports each measure on its natural scale.
  Model comparison converts every loss to a utility.
+ **Aggregation.** Most measures are a sum or a mean of pointwise terms. RMSE,
  $R^2$, and balanced accuracy are functions of means. They have no pointwise
  terms that add up to the estimate.

## Notation

The paper uses one notation for all measures. The table maps each symbol to an
argument of `measure_*()` or to a field of its result.

| Symbol | Meaning | In **loo** |
| :--- | :--- | :--- |
| $y_i$ | observed outcome $i$ | `y` |
| $\theta^s$ | posterior draw $s$, $s = 1, \ldots, S$ | rows of the input matrices |
| $\ell_i^s = \log p(y_i \mid \theta^s)$ | log density of $y_i$ under draw $s$ | `ylp`, an $S \times n$ matrix |
| $\tilde y_i^s$ | posterior predictive draw | `ypred`, an $S \times n$ matrix |
| $\mathbb{E}[\tilde y_i \mid \theta^s]$ | predictive mean under draw $s$ | `mupred`, an $S \times n$ matrix, or $S \times n \times C$ for $C$ classes |
| $w_i^s$ | normalized weight of draw $s$ (Vehtari et al., 2026, Eq. 2.5) | `log_weights`; `NULL` gives $w_i^s = 1/S$ |
| $\text{elpd}_i$, $\text{rps}_i$, $\text{ae}_i$, … | pointwise term of a measure, on its natural scale | `$pointwise` |
| $\widehat{\text{elpd}}$, $\widehat{\text{rps}}$, $\widehat{\text{mae}}$, … | estimate of a measure | `$estimates[, "Estimate"]` |
| $\widehat{\text{SE}}$ | standard error of the estimate | `$estimates[, "SE"]` |
| $\widehat{\text{Var}}$ | sample variance over the observations | |

::: {.callout .callout-note}
**Weights**

Every estimator in this article uses weights $w_i^s$. With
`log_weights = NULL`, all draws get the weight $1/S$. This gives in-sample
predictions. Other weights give other predictive schemes, for example PSIS-LOO.
:::

## Data and models

```{r load-libraries}
library(loo)
library(ggplot2)
theme_set(theme_minimal())
```

For each outcome type we fit two `brms` models:

+ **with predictors**: a model that uses the predictors of the data,
+ **intercept only**: a model that predicts the same distribution for every
  observation.

The intercept-only model shows how a measure looks for a model that performs
badly. Fitting is not the topic of this article. The model code is in the
[appendix](#appendix-model-code).

| Object | Data | Outcome | Predictors | Used for |
| :--- | :--- | :--- | :--- | :--- |
| `fits_count` | `roaches` (`rstanarm`) | count, Poisson | `sqrt_roach1`, `treatment`, `senior` | log score |
| `fits_cont` | `kidiq` (`rstanarm`) | continuous, normal | `mom_iq`, `mom_hs` | CRPS, point error, $R^2$ |
| `fits_bin` | `wells` (`rstanarm`) | binary, Bernoulli | `dist100`, `arsenic` | Brier score |
| `fits_ord` | simulated | ordered, 4 categories | `x` | RPS |
| `fits_cat` | simulated | nominal, 3 unbalanced classes | `x` | accuracy |

Each object is a list with the elements `"with predictors"` and
`"intercept only"`.

**Prediction tasks.** The models predict the following outcomes:

+ **Roach counts** (`fits_count`). A pest-management trial in 262 apartment
  buildings. The outcome `y` is the number of roaches caught in traps after the
  treatment. The predictors are the square root of the count before the
  treatment (`sqrt_roach1`), whether the building received the treatment
  (`treatment`), and whether only elderly people live in the building
  (`senior`).
+ **Child IQ** (`fits_cont`). A survey of 434 American mothers and their
  children. The outcome `kid_score` is the IQ score of the child. The
  predictors are the IQ of the mother (`mom_iq`) and whether the mother
  finished high school (`mom_hs`).
+ **Well switching** (`fits_bin`). A survey of 3020 households in Bangladesh
  whose wells had high arsenic levels. The outcome `switch` is 1 if the
  household switched to a safe well. The predictors are the distance to the
  nearest safe well in 100 meters (`dist100`) and the arsenic level of the
  current well (`arsenic`).
+ **Ordered rating** (`fits_ord`). Simulated data with 300 observations. The
  outcome `y` is a rating on an ordered scale from 1 to 4. Higher values of the
  predictor `x` make higher ratings more likely.
+ **Class membership** (`fits_cat`). Simulated data with 300 observations. The
  outcome `y` is one of three unordered classes. Class 1 is the most frequent
  class (169 observations), and class 3 the least frequent (35 observations).
  The predictor `x` shifts the probabilities between the classes.

```{r fit-models, echo=FALSE, results="hide"}
SEED <- 42

fit_pair <- function(formula, data, family) {
  fit <- brms::brm(formula, data = data, family = family,
                   chains = 2, iter = 1000, seed = SEED, refresh = 0)
  list(
    "with predictors" = fit,
    "intercept only" = update(fit, formula. = . ~ 1, seed = SEED, refresh = 0)
  )
}

# count
data(roaches, package = "rstanarm")
roaches$sqrt_roach1 <- sqrt(roaches$roach1)
fits_count <- fit_pair(y ~ sqrt_roach1 + treatment + senior, roaches, poisson())

# continuous
data(kidiq, package = "rstanarm")
fits_cont <- fit_pair(kid_score ~ mom_iq + mom_hs, kidiq, gaussian())

# binary
data(wells, package = "rstanarm")
wells$dist100 <- wells$dist / 100
fits_bin <- fit_pair(switch ~ dist100 + arsenic, wells, brms::bernoulli())

# ordered: latent logistic model with three cut points
set.seed(SEED)
n_sim <- 300
x <- rnorm(n_sim)
latent <- 1.2 * x + rlogis(n_sim)
df_ord <- data.frame(y = 1 + rowSums(outer(latent, c(-1, 0.3, 1.5), ">")), x = x)
fits_ord <- fit_pair(y ~ x, df_ord, brms::cumulative("logit"))

# nominal: class 1 is the most frequent
set.seed(SEED + 1)
x <- rnorm(n_sim)
prob <- cbind(1, exp(-1 + 1.5 * x), exp(-2 - x))
prob <- prob / rowSums(prob)
df_cat <- data.frame(y = apply(prob, 1, function(p) sample(3, 1, prob = p)), x = x)
fits_cat <- fit_pair(y ~ x, df_cat, brms::categorical())
```

## Log score {#log-score}

**Motivation.** The log score evaluates the predictive density at the observed
value. A model gets a high score when it puts much probability near the
observation. It uses the whole predictive distribution, and it strongly
penalizes observations the model considers very unlikely.

**When to use.** The log score works for any outcome type. It is the default
measure in **loo**. It needs the log-likelihood draws `ylp`.

**Key formula.** The pointwise term and the estimate are
(Vehtari et al., 2026, Eqs. 2.16–2.18)

$$
\text{elpd}_i = \log \sum_{s=1}^S \exp\!\left(\log w_i^s + \ell_i^s\right),
\qquad
\widehat{\text{elpd}} = \sum_{i=1}^n \text{elpd}_i .
$$

**loo** reports two rescaled versions: $\widehat{\text{mlpd}} =
\widehat{\text{elpd}} / n$ on the mean scale, and the information criterion
$\widehat{\text{ic}} = -2\,\widehat{\text{elpd}}$.

**Standard error.** The elpd is a sum, the mlpd is a mean
(Vehtari et al., 2026, Eqs. 2.8, 2.9):

$$
\widehat{\text{SE}}(\widehat{\text{elpd}}) =
  \sqrt{n\,\widehat{\text{Var}}(\text{elpd}_i)},
\qquad
\widehat{\text{SE}}(\widehat{\text{mlpd}}) =
  \sqrt{\widehat{\text{Var}}(\text{elpd}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{ic}}) =
  2\,\widehat{\text{SE}}(\widehat{\text{elpd}}) .
$$

**Implementation.**

```{r log-score}
y_count <- fits_count[[1]]$data$y
elpd_count <- lapply(fits_count, function(fit) measure_elpd(ylp = brms::log_lik(fit)))
elpd_count

ylp_count <- brms::log_lik(fits_count[["with predictors"]])
measure_mlpd(ylp = ylp_count)
measure_ic(ylp = ylp_count)
```

**Plot.** Each point is one building. The y-axis shows its pointwise log score
$\text{elpd}_i$. Higher is better.

```{r log-score-plot}
df_log <- do.call(rbind, lapply(names(elpd_count), function(m) {
  data.frame(model = m, y = y_count, elpd_i = as.vector(elpd_count[[m]]$pointwise))
}))

ggplot(df_log, aes(y, elpd_i)) +
  geom_point(alpha = 0.5) +
  scale_x_sqrt() +
  facet_wrap(~model) +
  labs(x = "observed roach count (square-root scale)",
       y = "pointwise log score")
```

**Interpretation.** The intercept-only model predicts the same distribution for
every building. Its log score falls steeply as the observed count moves away
from the average count. The model with predictors uses the count before the
treatment. It moves the predictive distribution toward the observed count, so
its points lie higher for most buildings.

## Ranked probability score {#ranked-probability-score}

**Motivation.** The ranked probability score compares the predictive
distribution with the observed value through absolute distances. Unlike the
log score, it rewards predictions that are *close* to the observation, not only
predictions that hit it. It is less sensitive to single outliers.

**When to use.** Use `rps` for ordered categorical outcomes (RPS) and for
continuous outcomes (CRPS). It is on the scale of $y$, so a value can be
read in the units of the outcome. Use `srps` when outcomes differ strongly in
their predictive spread. Both need the posterior predictive draws `ypred`.

**Key formula.** Let $X$ and $X'$ be independent draws from the predictive
distribution of observation $i$ (Vehtari et al., 2026, Eqs. 2.20–2.22):

$$
\text{rps}_i = \mathbb{E}|X - y_i| - \tfrac{1}{2}\,\mathbb{E}|X - X'|, \qquad
\text{srps}_i = -\frac{\mathbb{E}|X - y_i|}{\mathbb{E}|X - X'|}
  - \tfrac{1}{2}\log \mathbb{E}|X - X'| .
$$

The estimates are the means $\widehat{\text{rps}} = \frac{1}{n}\sum_i
\text{rps}_i$ and $\widehat{\text{srps}} = \frac{1}{n}\sum_i \text{srps}_i$.
**loo** estimates both expectations from one set of draws. The estimator of
$\mathbb{E}|X - X'|$ is a probability-weighted moment estimator
(Vehtari et al., 2026, Eq. 2.25). `rps` is a loss, and `srps` is a utility.

**Standard error.** Both are means (Vehtari et al., 2026, Eq. 2.8):

$$
\widehat{\text{SE}}(\widehat{\text{rps}}) =
  \sqrt{\widehat{\text{Var}}(\text{rps}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{srps}}) =
  \sqrt{\widehat{\text{Var}}(\text{srps}_i) / n} .
$$

**Implementation.** For the continuous outcome, `measure_rps()` gives the CRPS:

```{r crps}
y_cont <- fits_cont[[1]]$data$kid_score
ypred_cont <- lapply(fits_cont, brms::posterior_predict)

(crps_cont <- lapply(ypred_cont, function(yp) measure_rps(y = y_cont, ypred = yp)))
lapply(ypred_cont, function(yp) measure_srps(y = y_cont, ypred = yp))
```

For the ordered outcome, the same functions give the RPS and the SRPS:

```{r rps}
y_ord <- fits_ord[[1]]$data$y
ypred_ord <- lapply(fits_ord, brms::posterior_predict)

lapply(ypred_ord, function(yp) measure_rps(y = y_ord, ypred = yp))
lapply(ypred_ord, function(yp) measure_srps(y = y_ord, ypred = yp))
```

**Plot.** Both panels show the same child. We chose the child for which the
predictors reduce the CRPS the most. The solid line is the predictive CDF. The
dashed line is the step function of the observed score. The CRPS integrates
the squared vertical gap between the two lines, so a smaller shaded area is
better.

```{r crps-plot}
crps_i <- sapply(crps_cont, function(r) as.vector(r$pointwise))
i <- which.max(crps_i[, "intercept only"] - crps_i[, "with predictors"])
grid <- seq(min(sapply(ypred_cont, function(yp) min(yp[, i])), y_cont[i]),
            max(sapply(ypred_cont, function(yp) max(yp[, i])), y_cont[i]),
            length.out = 400)

df_cdf <- do.call(rbind, lapply(names(ypred_cont), function(m) {
  data.frame(
    model = sprintf("%s (CRPS = %.1f)", m, crps_i[i, m]),
    x = grid,
    cdf = ecdf(ypred_cont[[m]][, i])(grid),
    step = as.numeric(grid >= y_cont[i])
  )
}))

ggplot(df_cdf, aes(x)) +
  geom_ribbon(aes(ymin = pmin(cdf, step), ymax = pmax(cdf, step)),
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = cdf)) +
  geom_line(aes(y = step), linetype = "dashed") +
  facet_wrap(~model) +
  labs(x = "kid_score", y = "cumulative probability")
```

**Interpretation.** The intercept-only model predicts the same distribution for
every child, centered at the average score. For this child, its CDF rises far
from the observed score, and the shaded area is large. The model with
predictors uses the mother's IQ and schooling. Its CDF rises closer to the
observed score, and the shaded area is smaller.

## Brier score {#brier-score}

**Motivation.** The Brier score is the squared distance between the predicted
probability of an event and the observed outcome, 0 or 1. It rewards
probabilities that are both calibrated and sharp.

**When to use.** Use it for binary outcomes. **loo** implements the binary
Brier score only. `measure_brier()` needs the event probabilities of the
draws, passed as `ypred`.

**Key formula.** With the predicted event probability
$\hat p_i = \sum_s w_i^s\, p(\tilde y_i = 1 \mid \theta^s)$
(Vehtari et al., 2026, Eqs. 2.26–2.28),

$$
\text{bs}_i = \left(\hat p_i - y_i\right)^2, \qquad
\widehat{\text{bs}} = \frac{1}{n}\sum_{i=1}^n \text{bs}_i .
$$

The Brier score is a loss. It ranges from 0 to 1.

**Standard error.** The Brier score is a mean
(Vehtari et al., 2026, Eq. 2.8):

$$
\widehat{\text{SE}}(\widehat{\text{bs}}) =
  \sqrt{\widehat{\text{Var}}(\text{bs}_i) / n} .
$$

**Implementation.**

```{r brier}
y_bin <- fits_bin[[1]]$data$switch
p_bin <- lapply(fits_bin, brms::posterior_epred)

(brier_bin <- lapply(p_bin, function(p) measure_brier(y = y_bin, ypred = p)))
```

**Plot.** Each point is one household. The x-axis shows the predicted
probability $\hat p_i$ that it switches wells. Every point lies on one of two
curves: $\hat p_i^2$ if the household did not switch, and $(1 - \hat p_i)^2$
if it switched. Lower is better.

```{r brier-plot}
df_brier <- do.call(rbind, lapply(names(p_bin), function(m) {
  data.frame(
    model = m,
    p = colMeans(p_bin[[m]]),
    bs_i = as.vector(brier_bin[[m]]$pointwise),
    switched = factor(y_bin, levels = 0:1, labels = c("no", "yes"))
  )
}))

ggplot(df_brier, aes(p, bs_i, color = switched)) +
  geom_function(fun = function(p) p^2, color = "grey70", linetype = "dotted") +
  geom_function(fun = function(p) (1 - p)^2, color = "grey70", linetype = "dotted") +
  geom_point(alpha = 0.5) +
  facet_wrap(~model) +
  labs(x = "predicted probability of switching",
       y = "pointwise Brier score", color = "switched")
```

**Interpretation.** The intercept-only model gives every household the same
probability, the proportion of households that switched. All its points fall
on two spots, one on each curve. The model with predictors spreads the
probabilities. A household moves down its curve when the model is confident
and right. It moves up its curve when the model is confident and wrong.

## Point error {#point-error}

**Motivation.** The point-error measures reduce the predictive distribution to
one point prediction, the predictive mean. They then measure the distance
between that prediction and the observed value. They are easy to interpret,
but they ignore the predictive uncertainty.

**When to use.** Use them for continuous outcomes when the quality of the point
prediction matters. MAE is less sensitive to large errors than MSE and RMSE.
MAE and RMSE are on the scale of $y$. All three need the predictive means
`mupred`.

**Key formula.** With the point prediction
$\hat y_i = \sum_s w_i^s\, \mathbb{E}[\tilde y_i \mid \theta^s]$, the pointwise
errors are $\text{ae}_i = |y_i - \hat y_i|$ and
$\text{sqe}_i = (y_i - \hat y_i)^2$ (Vehtari et al., 2026, Eqs. 2.29–2.34):

$$
\widehat{\text{mae}} = \frac{1}{n}\sum_{i=1}^n \text{ae}_i, \qquad
\widehat{\text{mse}} = \frac{1}{n}\sum_{i=1}^n \text{sqe}_i, \qquad
\widehat{\text{rmse}} = \sqrt{\widehat{\text{mse}}} .
$$

All three are losses.

**Standard error.** MAE and MSE are means (Vehtari et al., 2026, Eq. 2.8).
RMSE is a function of a mean, so its standard error uses the delta method
(Vehtari et al., 2026, Eq. 2.33):

$$
\widehat{\text{SE}}(\widehat{\text{mae}}) =
  \sqrt{\widehat{\text{Var}}(\text{ae}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{mse}}) =
  \sqrt{\widehat{\text{Var}}(\text{sqe}_i) / n},
\qquad
\widehat{\text{SE}}(\widehat{\text{rmse}}) =
  \frac{\widehat{\text{SE}}(\widehat{\text{mse}})}{2\,\widehat{\text{rmse}}} .
$$

**Implementation.**

```{r point-error}
mu_cont <- lapply(fits_cont, brms::posterior_epred)

lapply(mu_cont, function(mu) measure_mae(y = y_cont, mupred = mu))
lapply(mu_cont, function(mu) measure_mse(y = y_cont, mupred = mu))
lapply(mu_cont, function(mu) measure_rmse(y = y_cont, mupred = mu))
```

**Plot.** Each point is one child. Each vertical segment is the error of the
point prediction, the distance to the dashed diagonal $y = \hat y$. Shorter
segments are better.

```{r point-error-plot}
df_pe <- do.call(rbind, lapply(names(mu_cont), function(m) {
  data.frame(model = m, y = y_cont, yhat = colMeans(mu_cont[[m]]))
}))

ggplot(df_pe, aes(yhat, y)) +
  geom_segment(aes(xend = yhat, yend = yhat), alpha = 0.3) +
  geom_abline(linetype = "dashed") +
  geom_point(size = 0.8) +
  facet_wrap(~model) +
  labs(x = "point prediction", y = "observed kid_score")
```

**Interpretation.** The intercept-only model predicts the average score for
every child. All its points form one vertical column, and each error is the
distance of the child's score from the average. The model with predictors
spreads the predictions along the diagonal. Its segments are shorter on
average, so its MAE and RMSE are lower.

## Predictive $R^2$ {#predictive-r2}

**Motivation.** The predictive $R^2$ compares the squared error of the model
with the squared error of the simplest prediction, the mean of the
observations. A value of 0 means the model predicts no better than the mean.

**When to use.** Use it for continuous outcomes, when a unitless summary of the
point prediction is needed. It is a predictive version of the classical $R^2$,
not the Bayesian $R^2$ of Gelman et al. (2019). It needs `mupred`.

**Key formula.** With $\text{sqe}_i$ and $\widehat{\text{mse}}$ from the
[point error](#point-error) section (Vehtari et al., 2026, Eqs. 2.35, 2.36),

$$
\widehat{R^2} = 1 - \frac{\widehat{\text{mse}}}{\widehat{\text{mse}}_y},
\qquad
\widehat{\text{mse}}_y = \frac{1}{n}\sum_{i=1}^n (y_i - \bar y)^2 .
$$

$R^2$ is a utility.

**Standard error.** $R^2$ is a function of two means, so its standard error
uses the delta method. With $c = \widehat{\text{mse}} /
\widehat{\text{mse}}_y$ (Vehtari et al., 2026, Eq. 2.39),

$$
\widehat{\text{SE}}(\widehat{R^2}) = \frac{1}{\widehat{\text{mse}}_y}
  \sqrt{\widehat{\text{Var}}\!\left(\text{sqe}_i - c\,(y_i - \bar y)^2\right) / n} .
$$

**Implementation.**

```{r r2}
lapply(mu_cont, function(mu) measure_r2(y = y_cont, mupred = mu))
```

**Plot.** Each point is one child. It compares the squared error of the model
with the squared error of the mean. Points below the dashed line are children
for whom the model is better than the mean.

```{r r2-plot}
df_r2 <- do.call(rbind, lapply(names(mu_cont), function(m) {
  data.frame(
    model = m,
    sqe_model = (y_cont - colMeans(mu_cont[[m]]))^2,
    sqe_mean = (y_cont - mean(y_cont))^2
  )
}))

ggplot(df_r2, aes(sqe_mean, sqe_model)) +
  geom_abline(linetype = "dashed") +
  geom_point(alpha = 0.5) +
  scale_x_sqrt() +
  scale_y_sqrt() +
  facet_wrap(~model) +
  labs(x = "squared error of the mean (square-root scale)",
       y = "squared error of the model (square-root scale)")
```

**Interpretation.** The intercept-only model predicts the average score, so
its squared errors are the squared errors of the mean. Its points lie on the
dashed line, and its $R^2$ is close to 0. The model with predictors moves
points off the line. Points below the line outweigh points above it, so its
$R^2$ is positive.

## Classification accuracy {#classification-accuracy}

**Motivation.** Accuracy is the proportion of observations whose predicted class
is the observed class. It is easy to interpret, but it uses only the predicted
class, not the predictive probabilities. Balanced accuracy gives each class the
same weight. It prevents a model from scoring well by predicting only the most
frequent class.

**When to use.** Use them for binary or nominal outcomes. Prefer `bacc` when the
classes are unbalanced. Both need the class probabilities `mupred`: an
$S \times n$ matrix for binary outcomes, or an $S \times n \times C$ array for
$C$ classes.

**Key formula.** The predicted class is
$\hat c_i = \arg\max_c \sum_s w_i^s\, p(\tilde y_i = c \mid \theta^s)$, and the
pointwise accuracy is $\text{acc}_i = \mathbb{1}\{\hat c_i = y_i\}$. With $n_c$
observations in class $c$ (Vehtari et al., 2026, Eqs. 2.40–2.44),

$$
\widehat{\text{acc}} = \frac{1}{n}\sum_{i=1}^n \text{acc}_i,
\qquad
\widehat{\text{bacc}} = \frac{1}{C}\sum_{c=1}^C \widehat{\text{acc}}_c,
\qquad
\widehat{\text{acc}}_c = \frac{1}{n_c}\sum_{i:\, y_i = c} \text{acc}_i .
$$

Both are utilities.

**Standard error.** Both use the binomial standard error, for balanced accuracy
within each class (Vehtari et al., 2026, Eqs. 2.43, 2.45):

$$
\widehat{\text{SE}}(\widehat{\text{acc}}) =
  \sqrt{\frac{\widehat{\text{acc}}\,(1 - \widehat{\text{acc}})}{n}},
\qquad
\widehat{\text{SE}}(\widehat{\text{bacc}}) = \frac{1}{C}
  \sqrt{\sum_{c=1}^C \frac{\widehat{\text{acc}}_c\,(1 - \widehat{\text{acc}}_c)}{n_c}} .
$$

**Implementation.**

```{r accuracy}
y_cat <- fits_cat[[1]]$data$y
mu_cat <- lapply(fits_cat, brms::posterior_epred)

(acc_cat <- lapply(mu_cat, function(mu) measure_acc(y = y_cat, mupred = mu)))
(bacc_cat <- lapply(mu_cat, function(mu) measure_bacc(y = y_cat, mupred = mu)))
```

**Plot.** The bars show the accuracy within each observed class. The labels
give the class sizes. The dashed line is the accuracy, and the dotted line is
the balanced accuracy. Higher is better.

```{r accuracy-plot}
df_acc <- do.call(rbind, lapply(names(mu_cat), function(m) {
  acc_i <- as.vector(acc_cat[[m]]$pointwise)
  data.frame(
    model = m,
    class = factor(sort(unique(y_cat))),
    acc_c = as.vector(tapply(acc_i, y_cat, mean)),
    n_c = as.vector(table(y_cat))
  )
}))
df_lines <- data.frame(
  model = names(mu_cat),
  acc = sapply(acc_cat, function(r) r$estimates[, "Estimate"]),
  bacc = sapply(bacc_cat, function(r) r$estimates[, "Estimate"])
)

ggplot(df_acc, aes(class, acc_c)) +
  geom_col(fill = "grey70") +
  geom_text(aes(label = paste("n =", n_c)), vjust = -0.4) +
  geom_hline(data = df_lines, aes(yintercept = acc), linetype = "dashed") +
  geom_hline(data = df_lines, aes(yintercept = bacc), linetype = "dotted") +
  scale_y_continuous(limits = c(0, 1.05)) +
  facet_wrap(~model) +
  labs(x = "observed class", y = "accuracy within the class")
```

**Interpretation.** The intercept-only model predicts class 1, the most
frequent class, for every observation. Its accuracy is 1 in class 1 and 0 in
the other classes. Its accuracy equals the share of class 1, but its balanced
accuracy is only $1/3$. The model with predictors also classifies part of
classes 2 and 3 correctly. Its balanced accuracy is higher, although its
accuracy can be close to that of the intercept-only model.

## References

Gelman, A., Goodrich, B., Gabry, J., and Vehtari, A. (2019). R-squared for
Bayesian regression models. *The American Statistician*, 73(3):307–309.
doi:10.1080/00031305.2018.1549100

Vehtari, A., Axen, S., Weber, F., and Bockting, F. (2026). Predictive measures
and model comparison in the loo package: A formula reference. [TODO: arXiv ID]

## See also

- [Model comparison](model-comparison.html) — differences between models and
  their standard errors.
- [TODO: schemes and `pred_measure()` article]

## Appendix: model code {#appendix-model-code}

```{r fit-models-code, ref.label="fit-models", eval=FALSE}
```
