---
title: "Overview of scores and metrics"
author: "Florence Bockting"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: FALSE
vignette: >
  %\VignetteIndexEntry{Overview of scores and metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Introduction

In Bayesian model evaluation, predictive performance is assessed by comparing a model's posterior predictive distribution against observed data. Depending on whether the outcome is continuous, binary, or categorical, and whether we evaluate the entire probability distribution or a specific point estimate, different scoring rules and performance metrics are appropriate.

The table below summarizes the core metrics implemented in our framework. 

| Metric Keyword | Full Name | Metric Type | Orientation | Target Outcome Type |
| :--- | :--- | :--- | :--- | :--- |
| `lppd` / `elppd` | (Expected) Log Pointwise Predictive Density | Scoring Rule (Distributional) | Higher is Better | Any |
| `acc` | Accuracy | Proper Scoring Rule | Higher is Better | Nominal Categorical |
| `bacc` | Balanced Accuracy | Adjusted Classification Metric | Higher is Better | Imbalanced Categorical |
| `brier` | Brier Score | Proper Scoring Rule | Lower is Better | Binary |
| `crps2` | Continuous Ranked Probability Score | Proper Scoring Rule (Distributional) | Lower is Better | Continuous |
| `scrps2` | Scaled Continuous Ranked Probability Score | Scale-Invariant Proper Scoring Rule | Higher is Better | Continuous |
| `rps` | Ranked Probability Score | Proper Scoring Rule | Lower is Better | Ordered Categorical |
| `srps` | Scaled Ranked Probability Score | Scale-Invariant Proper Scoring Rule | Higher is Better | Ordered Categorical |
| `mae` | Mean Absolute Error | Point Estimation Metric | Lower is Better | Continuous |
| `mse` | Mean Squared Error | Point Estimation Metric | Lower is Better | Continuous |
| `rmse` | Root Mean Squared Error | Point Estimation Metric | Lower is Better | Continuous |
| `r2` | Coefficient of Determination | Summary Fit Metric | Higher is Better | Continuous |

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warnings = FALSE)
```

## Data Examples
```{r, message=FALSE, warnings=FALSE, include = FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
# devtools::load_all("../brms")
```

```{r, include = FALSE}
plot_ptw <- function(ws, loo, kfold, label, show_estimate = TRUE){
  
  # 1. Safe estimation extractor
  ptw_est <- function(x) {
    if (is.null(x)) return(NA)
    if (!is.null(x$estimates[1])) {
      return(x$estimates[1])
    }
    x$estimates["Estimate"]
  }

  # 2. Build estimates and create the formatted labels right away
  estimates <- tibble(
    source = c("ws", "loo", "kfold"),
    estimate = c(ptw_est(ws), ptw_est(loo), ptw_est(kfold))
  ) |> 
    mutate(source_label = if_else(
      is.na(estimate), 
      paste0(source, " (no data)"), 
      paste0(source, " (est. = ", round(estimate, 2), ")")
    ))
  
  # Turn into a factor to preserve the ws -> loo -> kfold order
  estimates$source_label <- factor(estimates$source_label, levels = estimates$source_label)
  
  # 3. Safe data binding for the dots
  g_data <- dplyr::bind_rows(
    tibble(source = "ws", value = if (!is.null(ws)) ws$pointwise else as.numeric(NA)),
    tibble(source = "loo", value = if (!is.null(loo)) loo$pointwise else as.numeric(NA)),
    tibble(source = "kfold", value = if (!is.null(kfold)) kfold$pointwise else as.numeric(NA))
  )
  
  # 4. Align facet names for the dots data
  plot_df <- g_data |> 
    left_join(estimates, by = "source") |>
    mutate(source = source_label)
  
  # 5. Align facet names for the line data so ggplot matches them up perfectly
  line_data <- estimates |> 
    mutate(source = source_label)
  
  # Plotting
  g <- ggplot(plot_df, mapping = aes(x = value)) +
    geom_dotplot()
  
  if (isTRUE(show_estimate)) {
    g <- g +
      geom_vline(
        data = line_data, # Uses the aligned facet labels now
        aes(xintercept = estimate), color = "red", linetype = "solid", linewidth = 1
      )
  }

  g + 
    facet_wrap(~ source, nrow = 1) +
    theme_classic() +
    xlab(label)
}
```

```{r}
# see vignettes/articles-online-only/test-data/generate-data.R for code 
# to generate these objects

# poisson model; count data
res_roaches <- readRDS("../../tests/testthat/data-for-tests/res_roaches.Rds")
# normal model; continuous data
res_sleep <- readRDS("../../tests/testthat/data-for-tests/res_sleep.Rds")
# bernoulli model; binary data
res_binary <- readRDS("../../tests/testthat/data-for-tests/res_binary.Rds")
# categorical model; nominal categories
res_penguins <- readRDS("../../tests/testthat/data-for-tests/res_penguins.Rds")  
# binomial model; bounded count data 
res_binomial <- readRDS("../../tests/testthat/data-for-tests/res_binomial.Rds")
```

## Log pointwise predictive density (lppd) 
The log pointwise predictive density aggregates the log posterior predictive
densities over a pointwise partition of the data $y = (y_1,\ldots,y_n)$:
$$
\text{lppd} = \sum_{i=1}^n \log \int p(y_i \mid \theta) p(\theta \mid y) d\theta 
$$
Monte Carlo estimator with posterior draws:
$$
\widehat{\text{lppd}} = \sum_{i=1}^n \log \left(\frac{1}{S}\sum_{s=1}^S 
p(y_i\mid \theta^{(s)})\right)
= \sum_{i=1}^n \underbrace{\Big[\text{log-sum-exp}_s\{\log p(y_i\mid 
\theta^{(s)})\} - \log S\Big]}_{\text{lppd}_i}
$$

We use $\log p(y_i\mid \theta^{(s)})$ for numerical stability.

### Conceptual variants
We define pointwise terms $\text{lppd}_i$ and their sums. 
For standard errors we use the sample variance across $i$:

- $\widehat{\text{Var}}_i(\text{lppd}_i) = \frac{1}{n-1} \sum_{i=1}^n 
(\text{lppd}_i - \text{lppd})^2 \text{, with } \text{lppd} = 
\frac{1}{n}\sum_i \text{lppd}_i$
- $se(\text{sum}) \approx \sqrt{n \cdot \widehat{\text{Var}}(\text{lppd}_i)}$
- $se(\text{mean}) \approx \sqrt{\frac{\widehat{\text{Var}}(\text{lppd}_i)}{n}}$

#### Within-sample (training data; optimistic for new data)
+ Draws: $\theta^{(s)} \sim p(\theta \mid y)$
+ Pointwise term: 
$$
\text{lppd}_i^\text{ws} = \text{log-sum-exp}_s \big(\log p(y_i 
\mid \theta^s)\big) - \log(S)
$$
+ Estimate and Standard error:
  + $\text{lppd}^\text{sw} = \sum_i^n \text{lppd}_i^\text{ws}$
  + $\widehat{\text{elppd}}^\text{sw} = \text{lppd}^\text{sw}$ 
  (optimistically biased for future data)
  + $se(\text{lppd}^\text{sw}) \approx \sqrt{n \cdot \widehat{\text{Var}}(\text{lppd}_i^\text{ws})}$

#### Leave-one-out cross-validation (LOO-CV; held-out)
+ Draws: $\theta_{−i}^{(s)} \sim p(\theta \mid y_{−i})$
+ Pointwise term: 
$$
\text{lppd}_i^\text{loo} = \text{log-sum-exp}_s(\log p(y_i \mid \theta_{-i}^s)) - \log(S)
$$
+ Estimate and Standard Error:
  + $\text{lppd}^\text{loo} = \sum_i^n \text{lppd}_i^\text{loo}$, 
  an estimate of elppd
  + $se(\text{lppd}^\text{loo}) \approx \sqrt{n \cdot \widehat{\text{Var}}(\text{lppd}_i^\text{loo})}$

#### PSIS-LOO (held-out approximation)
+ Reweights full-data draws $\theta^{(s)} \sim p(\theta \mid y)$ to approximate 
$p(\theta \mid y_{−i})$ with normalized PSIS weights $\tilde w_i^{(s)}$ ($\sum \tilde w_i^{(s)} = 1$)
+ Pointwise term:
$$
\text{lppd}_i^\text{psis} = \text{log-sum-exp}_s\big(\log p(y_i \mid \theta^s) + \log \tilde w_i^{(s)}\big)
$$
+ Estimate and Standard Error:
  + $\text{lppd}^\text{psis} = \sum_i^n \text{lppd}_i^\text{psis}$, an estimate 
  of elppd
  + $se(\text{lppd}^\text{psis}) \approx \sqrt{n \cdot \widehat{\text{Var}}(\text{lppd}_i^\text{psis})}$

#### K-fold cross-validation
Partition indices into $K$ disjoint folds, $F_1,\ldots, F_K$. For each fold 
$m$, fit the model on the training data $y_{-F_m}$, draw parameters from the 
posterior distribution $\theta^{(m,s)} \sim p(\theta \mid y_{-F_m})$ and 
evaluate fold $F_m$.

+ Draws: $\theta^{(m,s)} \sim p(\theta \mid y_{-F_m})$
+ Pointwise term:
$$
\text{lppd}_i^\text{kfold} = \text{log-sum-exp}_s\big(\log p(y_i\mid \theta^{(m,s)})\big) - \log S,\quad i \in F_m
$$
+ Estimate and Standard Error:
  + $\text{lppd}^\text{kfold} = \sum_i \text{lppd}_i^\text{kfold}$, an estimate
   for elppd
  + $se(\text{lppd}^\text{kfold}) \approx \sqrt{n \cdot \widehat{\text{Var}}(\text{lppd}_i^\text{kfold})}$

## Expected log pointwise predictive density (elppd)
**elppd** (expected log pointwise predictive density) is the sum, over a chosen 
pointwise partition of the data, of the expected log posterior predictive 
density for future observations drawn from the true data-generating 
distribution $p_t$. Formally, 
$$ 
\mathrm{elppd}=\sum_{i=1}^n \mathbb{E}_{\tilde y_i \sim p_t} \left[\log \int p(\tilde y_i \mid \theta) p(\theta \mid y) d\theta \right] =\sum_{i=1}^n \mathbb{E}_{\tilde y_i \sim p_t} \left[\text{lppd}_i \right] 
$$ 

Because $p_t$ is unknown, **elppd** is estimated from data-either by 
bias-correcting the in-sample lppd (e.g., WAIC) or via cross-validation 
(e.g., LOO/Psis-LOO/K-fold) that evaluates predictive densities on held-out 
data.
$$
\widehat{\text{elppd}} \approx \sum_i^n\text{lppd}_i
$$


```{r, warnings=FALSE, message=FALSE}
# within-sample elpd
elpd_ws <- elpd(ylp = res_roaches$ylp)
message("within-sample elpd:")
invisible(str(elpd_ws))

# psis-loo elpd
elpd_psis1 <- elpd(
  ylp = res_roaches$ylp, 
  log_weights = res_roaches$log_weights
)
elpd_psis2 <- elpd(
  ylp = NULL, 
  pointwise = res_roaches$loo$pointwise[, "elpd_loo"]
)
# expect_equal(elpd_psis1, elpd_psis2)
message("psis-loo elpd:")
invisible(str(elpd_psis1))

# kfold elpd
elpd_kfold <- elpd(
  ylp = NULL, 
  pointwise = res_roaches$kfold$pointwise[, "elpd_kfold"]
)
message("kfold elpd:")
invisible(str(elpd_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(elpd_ws, elpd_psis1, elpd_kfold, "lppd_i", show_estimate = FALSE)
```

### Mean log pointwise predictive density
The mean log predictive density (**mlpd**) is the average (across observations) 
of the model's log predictive density.

$$
\text{mlpd} = \frac{1}{n} \widehat{\text{elppd}} = \frac{1}{n} \sum_i^n\text{lppd}_i
$$
The standard error of the mean is then
$$
se(\text{mlpd}) = \frac{se(\text{elppd})}{n}
$$

```{r}
mlpd_ws <- mlpd(
  ylp = res_roaches$ylp
)
mlpd_loo <- mlpd(
  ylp = res_roaches$ylp, 
  log_weights = res_roaches$log_weights
)
mlpd_kfold <- mlpd(
  ylp = NULL, 
  pointwise = res_roaches$kfold$pointwise[, "elpd_kfold"]
)

message("within-sample mlpd:")
invisible(str(mlpd_ws))
message("psis-loo mlpd:") 
invisible(str(mlpd_loo))
message("kfold mlpd:")
invisible(str(mlpd_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(mlpd_ws, mlpd_loo, mlpd_kfold, "lppd_i")
```

## Log score vs. elpd
The log score is the log of the predictive density (or probability) evaluated 
at the observed value. With $S$ posterior draws $\theta^1, \ldots, \theta^S$ 
from $p(\theta \mid y)$, the posterior predictive density is approximated by 
Monte Carlo averaging
$$
\log p(y_i \mid y) \approx \log \Big[\frac{1}{S} \sum_{s=1}^S p(y_i \mid \theta^s) \Big]
$$
or using the log-sum-exp trick:
$$
\log p(y_i \mid y) \approx \Big[\text{log-sum-exp}_s\big(\log p(y_i \mid \theta^{(s)})\big) - \log S\Big]
$$

This definition looks equivalent to that we used in order to introduce the 
elpd. What is then the difference between log score and elpd?

+ **Log score**: The realized value $\log p(y \mid p(\theta \mid y))$ for an 
observed data point $y$ under a model's predictive distribution. It is a proper,
 local scoring rule used to evaluate probabilistic predictions.
+ **elpd**: The expectation of the log score for a future, as-yet-unobserved 
data point drawn from the data-generating process. It averages the log 
predictive density over the distribution of future data, and its pointwise 
sum over $n$ items is elppd; in practice we estimate it via cross-validation 
or using information criteria that correct in-sample bias.

## Accuracy (`acc`)

Accuracy here is the posterior predictive probability assigned to the observed category, averaged across (weighted) posterior draws.
Accuracy is a proper scoring rule suitable for nominal categorical outcomes and is positively oriented (higher is better). Accuracy ranges between 0 (all predictions are wrong) and 1 (all predictions are correct).

- Predictive probability of the observed class, Monte Carlo:
  $$
  \hat p_i = \frac{1}{S}\sum_{s=1}^S \Pr(Y_i = y_i \mid \theta^{(s)}),\quad
  \hat p_i^w = \sum_{s=1}^S \tilde w_i^{(s)}\,\Pr(Y_i = y_i \mid \theta^{(s)})
  $$
- Pointwise:
  $$
  \text{acc}_i = \hat p_i \ \text{ or }\ \hat p_i^w
  $$
- Estimate and Standard Error:
  $$
  \widehat{\text{acc}} = \frac{1}{n}\sum_{i=1}^n \text{acc}_i,\quad
  se(\widehat{\text{acc}}) = \sqrt{\frac{\widehat{\text{Var}}(\text{acc}_i)}{n}}
  $$

::: {.callout-tip}

Instead of drawing a single label per sample (`ypred`), we use the expected probability under the posterior (`mupred`). This is the Bayes-optimal decision rule under 0-1 loss (0-1 loss = 1 - Accuracy).
:::

```{r}
acc_ws <- acc(
  y = as.numeric(res_penguins$y), 
  mupred = res_penguins$mupred
)
acc_loo <- acc(
  y = as.numeric(res_penguins$y), 
  mupred = res_penguins$mupred,
  log_weights = res_penguins$log_weights
)
# acc_kfold <- acc(
#   y = as.numeric(res_penguins$y), 
#   mupred = res_penguins$mupred_kfold
# )
  
message("within-sample acc:")
invisible(str(acc_ws))
message("psis-loo acc:") 
invisible(str(acc_loo))
# message("kfold acc:") 
# invisible(str(acc_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(acc_ws, acc_loo, NULL, "acc_i")
```

## Balanced Accuracy (`bacc`)

Balanced accuracy averages the within-class mean of the predicted probability assigned to the correct class, giving each class equal weight.

Let $\mathcal{C} = \{1, \ldots, K\}$ be the set of classes and 
$n_c = |\{i : y_i = c\}|$ the number of observations in class $c$. The 
class-specific accuracy is the mean predicted probability of the correct 
category among observations that truly belong to class $c$:

- Class-specific accuracy:
  $$
  \text{acc}_c = \frac{1}{n_c} \sum_{i: y_i=c} \hat p_i^w
  $$
- Balanced accuracy:
  $$
  \widehat{\text{bacc}} = \frac{1}{K}\sum_{c=1}^K \text{acc}_c
  $$
- A convenient pointwise contribution that yields the same mean:
$$
\text{bacc}_i = \frac{\hat p_i^w}{n_{y_i}}
$$
- Estimate and Standard Error:
  $$
  \widehat{\text{bacc}} = \frac{1}{K}\sum_{c=1}^K \frac{1}{n_c}\sum_{i: y_i=c} \hat p_i^w,\quad
  se(\widehat{\text{bacc}}) = \sqrt{\frac{\widehat{\text{Var}}(\text{bacc}_i)}{n}}
  $$

::: {.callout-note}
When class frequencies are equal ($n_c = n/K$), balanced accuracy 
reduces to standard accuracy. For binary outcomes ($K=2$), it equals the mean 
of sensitivity and specificity.
:::

```{r}
print(table(as.numeric(res_penguins$y)))

bacc_ws <- bacc(
  y = as.numeric(res_penguins$y), 
  mupred = res_penguins$mupred
)
bacc_loo <- bacc(
  y = as.numeric(res_penguins$y), 
  mupred = res_penguins$mupred,
  log_weights = res_penguins$log_weights
)
# bacc_kfold <- bacc(
#   y = as.numeric(res_penguins$y), 
#   mupred = res_penguins$mupred_kfold
# )
  
message("within-sample bacc:")
invisible(str(bacc_ws))
message("psis-loo bacc:") 
invisible(str(bacc_loo))
# message("kfold bacc:") 
# invisible(str(bacc_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(bacc_ws, bacc_loo, NULL, "bacc_i")
```

## Brier Score (`brier`)
The Brier score is a strictly proper scoring rule for binary outcomes; it is the mean
 squared error between the predicted probability and the observed binary outcome. The Brier score is 
 negatively oriented (lower is better).
 The brier score ranges within [0, 1].

- Pointwise:
  $$
  \widehat{\text{brier}}_i^{\text{raw}} = \big(p_i - y_i\big)^2,\quad
  p_i = 
  \begin{cases}
  \frac{1}{S}\sum_{s=1}^S \Pr(y_i=1 \mid \theta^{(s)}) & \text{(unweighted)}\\[4pt]
  \sum_{s=1}^S \tilde w_i^{(s)}\,\Pr(y_i=1 \mid \theta^{(s)}) & \text{(PSIS weights)}
  \end{cases}
  $$
+ Estimate and Standard Error:
  $$
  \widehat{\text{bs}} = \frac{1}{n}\sum_{i=1}^n \text{bs}_i,\quad
  se(\widehat{\text{bs}}) = \sqrt{\frac{\widehat{\text{Var}}(\text{bs}_i)}{n}}
  $$

```{r}
brier_ws <- brier(
  y = res_binary$y, 
  ypred = res_binary$ypred
)
brier_loo <- brier(
  y = res_binary$y, 
  ypred = res_binary$ypred, 
  log_weights = res_binary$log_weights
)
brier_kfold <- brier(
  y = res_binary$y, 
  ypred = res_binary$ypred_kfold
)

message("within-sample brier:")
invisible(str(brier_ws))
message("psis-loo brier:") 
invisible(str(brier_loo))
message("kfold brier:") 
invisible(str(brier_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(brier_ws, brier_loo, brier_kfold, "brier_i")
```

## Continuous Ranked Probability Score (`crps`)
CRPS is a strictly proper scoring rule for continuous outcomes; it measures the squared CDF distance between the predictive CDF $F$ and the observed step function integrated over the support. It is the continuous limit of rps as $K \rightarrow \infty$. As for rps, also crps is negatively oriented (lower = better). The crps is on the same units as $y$.

With predictive draws $Y^s \sim p(\tilde y \mid \theta^s)$:
 
- Pointwise terms:
$$
\text{crps}_i = \int_{-\infty}^{\infty} \left(F_i(x) - \mathbf{1}(x \geq y_i)\right)^2 dx
$$
Equivalently, via the energy score representation (when $F_i$ has a finite 
first moment):
$$
\text{crps}_i = \mathbb E_F\big[Y - y_i\big] - \frac{1}{2}\mathbb E_F \big[Y-Y'\big]
$$
where $Y, Y'$ are independent draws from the predictive distribution $F_i$. 
Because computing $\mathbb{E}_F[|Y - Y'|]$ requires two independent sets of 
predictive draws, we instead use the identity (valid when $F$ is continuous):
$$
\text{crps}_i = \mathbb E_F\big[Y - y_i\big] + \mathbb E_F\big[Y\big] - 2\mathbb E_F \big[Y\cdot F(Y)\big]
$$
which requires only a single set of draws.
- Weighted Monte Carlo estimators (PSIS weights $\tilde w_i^{(s)}$; unweighted is the special case $\tilde w = 1/S$):
$$
\hat F_i^w(x) = \sum_{s=1}^S \tilde w_i^s \mathbf{1}(Y_i^s \leq x)
$$
- Computation:
$$
\mathbb{E}_F\big[|Y - y_i|\big] \approx \sum_{s=1}^S \tilde w_i^s \big|Y_i^s - y_i\big|, \qquad
\mathbb{E}_F[Y] \approx \sum_{s=1}^S \tilde w_i^s Y_i^s,
$$
$$
\mathbb{E}_F\big[Y \cdot F(Y)\big] \approx \sum_{s=1}^S \tilde w_i^s Y_i^s \hat F_i^w\!\left(Y_i^s\right)
$$
giving the pointwise estimator
$$
\widehat{\text{crps}}_i = \sum_s \tilde w_i^s \big|Y_i^s - y_i\big|
  + \sum_s \tilde w_i^s Y_i^s
  - 2 \sum_s \tilde w_i^s Y_i^s \hat F_i^w\!\left(Y_i^s\right)
$$
+ Estimate and Standard Error:
  $$
  \widehat{\text{crps}} = \frac{1}{n}\sum_{i=1}^n \text{crps}_i,\quad
  se(\widehat{\text{crps}}) = \sqrt{\frac{\widehat{\text{Var}}(\text{crps}_i)}{n}}
  $$

```{r}
crps_ws <- crps2(
  y = res_sleep$y, 
  ypred = res_sleep$ypred
)
crps_loo <- crps2(
  y = res_sleep$y, 
  ypred = res_sleep$ypred, 
  log_weights = res_sleep$log_weights
)
crps_kfold <- crps2(
  y = res_sleep$y, 
  ypred = res_sleep$ypred_kfold
)

message("within-sample crps:")
invisible(str(crps_ws))
message("psis-loo crps:")
invisible(str(crps_loo))
message("kfold crps:")
invisible(str(crps_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(crps_ws, crps_loo, crps_kfold, "crps_i")
```

### Scaled CRPS (`scrps`)

SCRPS is a locally scale-invariant proper scoring rule that normalizes CRPS by predictive spread (denominator) and adds a log penalty on spread (second term) [Bolin & Wallin, 2023]:

$$
\text{scrps}_i = -\frac{\mathbb{E}_F\big[|Y - y_i|\big]}{\mathbb{E}_F\big[|Y - Y'|\big]}
- \frac{1}{2}\log \mathbb{E}_F\big[|Y - Y'|\big]
$$
- Use the same weighted estimators for the two expectations as above.
- Rather than requiring two independent draw matrices, we recover
$\mathbb{E}_F[|Y-Y'|]$ from quantities already computed for the crps via the 
identity
$$
\mathbb{E}_F\big[|Y - Y'|\big] = 4\,\mathbb{E}_F\big[Y \cdot F(Y)\big] - 2\,\mathbb{E}_F[Y]
$$
- With importance weights this becomes
$$
\widehat{\mathbb{E}}_F\big[|Y - Y'|\big] =
  4\sum_s \tilde w_i^s Y_i^s \hat F_i^w\!\left(Y_i^s\right)
  - 2\sum_s \tilde w_i^s Y_i^s
$$
so the pointwise scrps estimator is
$$
\widehat{\text{scrps}}_i =
  -\frac{\displaystyle\sum_s \tilde w_i^s \big|Y_i^s - y_i\big|}
        {\widehat{\mathbb{E}}_F\big[|Y-Y'|\big]}
  - \frac{1}{2}\log\widehat{\mathbb{E}}_F\big[|Y-Y'|\big]
$$
- Estimate and Standard Error:
  $$
  \widehat{\text{scrps}} = \frac{1}{n}\sum_{i=1}^n \text{scrps}_i,\quad
  se(\widehat{\text{scrps}}) = \sqrt{\frac{\widehat{\text{Var}}(\text{scrps}_i)}{n}}
  $$

```{r}
scrps_ws <- scrps2(
  y = res_sleep$y, 
  ypred = res_sleep$ypred
)
scrps_loo <- scrps2(
  y = res_sleep$y, 
  ypred = res_sleep$ypred, 
  log_weights = res_sleep$log_weights
)
scrps_kfold <- scrps2(
  y = res_sleep$y, 
  ypred = res_sleep$ypred_kfold
)

message("within-sample scrps:")
invisible(str(scrps_ws))
message("psis-loo scrps:")
invisible(str(scrps_loo))
message("kfold scrps:")
invisible(str(scrps_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(scrps_ws, scrps_loo, scrps_kfold, "scrps_i")
```

## Ranked Probability Score (`rps`)
RPS is a strictly proper scoring rule for ordered categorical outcomes with $K$ categories. It compares the predictive CDF to the observed 
step CDF and generalizes the Brier score ($K=2$ recovers Brier). The standard 
definition is negatively oriented (lower is better). The range of rps is $[0, K-1]$.

- Predictive class probabilities and CDFs:
  $$
  \hat p_{i,k}^w = \sum_{s=1}^S \tilde w_i^{(s)}\,\Pr(Y_i=k\mid \theta^{(s)}),\quad
  \hat F_{i,k}^w = \sum_{j=1}^k \hat p_{i,j}^w
  $$
- Observed step CDF $O_{i,k} = \mathbf{1}(y_i \le k)$.
- Pointwise term:
  $$
  \widehat{\text{rps}}_i^{\text{raw}} = \sum_{k=1}^{K-1} (\hat F_{i,k}^w - O_{i,k})^2
  $$
- Estimate and Standard Error:
  $$
  \widehat{\text{rps}} = \frac{1}{n}\sum_{i=1}^n \text{rps}_i,\quad
  se(\widehat{\text{rps}}) = \sqrt{\frac{\widehat{\text{Var}}(\text{rps}_i)}{n}}
  $$

```{r}
rps_ws <- rps(
  y = res_binomial$y, 
  ypred = res_binomial$ypred,
  size = 10
)
rps_loo <- rps(
  y = res_binomial$y, 
  ypred = res_binomial$ypred, 
  size = 10,
  log_weights = res_binomial$log_weights
)
rps_kfold <- rps(
  y = res_binomial$y, 
  ypred = res_binomial$ypred_kfold, 
  size = 10
)
  
message("within-sample rps:")
invisible(str(rps_ws))
message("psis-loo rps:")
invisible(str(rps_loo))
message("kfold rps:")
invisible(str(rps_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(rps_ws, rps_loo, rps_kfold, "rps_i")
```

### Scaled RPS (`srps`)

By analogy with SCRPS [Bolin & Wallin, 2023], the scaled RPS is

$$
\text{srps}_i = -\frac{\mathbb{E}_F\big[|Y - y_i|\big]}{\mathbb{E}_F\big[|Y - Y'|\big]}
- \frac{1}{2}\log \mathbb{E}_F\big[|Y - Y'|\big]
$$

For ordered discrete support:

- Weighted mean absolute deviation:
  $$
  \widehat{\mathbb{E}}_F[|Y-y_i|] = \sum_{k=1}^K \hat p_{i,k}^w\,|k - y_i|
  $$
- Predictive spread:
  $$
  \widehat{\mathbb{E}}_F[|Y - Y'|] = 2\sum_{k=1}^{K-1} \hat F_{i,k}^w\big(1 - \hat F_{i,k}^w\big)
  $$
Estimation and SE follow the same mean/variance recipe as above. The srps has a positive direction, i.e., "higher is better".

```{r}
srps_ws <- srps(
  y = res_binomial$y, 
  ypred = res_binomial$ypred,
  size = 10
)
srps_loo <- srps(
  y = res_binomial$y, 
  ypred = res_binomial$ypred, 
  size = 10,
  log_weights = res_binomial$log_weights
)
srps_kfold <- srps(
  y = res_binomial$y, 
  ypred = res_binomial$ypred_kfold,
  size = 10
)
  
message("within-sample srps:")
invisible(str(srps_ws))
message("psis-loo srps:")
invisible(str(srps_loo))
message("kfold srps:")
invisible(str(srps_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(srps_ws, srps_loo, srps_kfold, "srps_i")
```

## Mean absolute error (`mae`)
MAE evaluates point predictions $\hat \mu_i$ against outcomes $y_i$ via 
absolute loss. It has the same units as $y$ and has orientation "lower is better". 

In our setup, we take `mupred` to be the vector of point predictions (one per 
observation), obtained from posterior expectations (e.g., via `posterior_epred`)
 or from PSIS-LOO expected predictions (via `E_loo`), as illustrated below.

- Pointwise: 
$$
\text{mae}_i = \mid y_i − \hat \mu_i \mid
$$ 
- Point prediction:
$$
\hat \mu_i \approx
\begin{cases}
\frac{1}{S}\sum_{s=1}^S \mu_i^{(s)}, & \mu_i^{(s)} = \mathbb{E}[\tilde y_i \mid \theta^{(s)}] \quad \text{(within-sample)}\\[6pt]
\sum_{s=1}^S \tilde w_i^{(s)}\,\mu_i^{(s)}, & \text{(PSIS-LOO held-out approximation)}
\end{cases}
$$
- Estimate and Standard Error: 
$$
\widehat{\text{mae}} = \frac{1}{n} \sum_{i=1}^n \text{mae}_i, \quad 
se(\widehat{\text{mae}}) = \sqrt{\frac{\widehat{\text{Var}}(\text{mae}_i)}{n}}
$$

```{r}
mae_ws <- mae(
  y = res_sleep$y, 
  mupred = res_sleep$mupred
)
mae_loo <- mae(
  y = res_sleep$y, 
  mupred = res_sleep$mupred_loo
)
mae_kfold <- mae(
  y = res_sleep$y, 
  mupred = res_sleep$mupred_kfold
)
  
message("within-sample mae:")
invisible(str(mae_ws))
message("psis-loo mae:")
invisible(str(mae_loo))
message("kfold mae:")
invisible(str(mae_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(mae_ws, mae_loo, mae_kfold, "mae_i")
```

## (Root) mean squared error (`rmse`/`mse`)
RMSE uses squared loss, has the same units as `y` and has a positive direction 
("lower is better").
RMSE evaluates point predictions $\hat\mu_i$ against observed outcomes $y_i$ 
using squared loss and then takes the square root of the average. It has the 
same units as $y$ and the direction "lower is better". In our setup, `mupred` 
is the point prediction per observation, computed as the posterior predictive 
mean (e.g., via `posterior_epred`) or its PSIS-LOO analogue via `E_loo`. RMSE 
is the square root of the mean squared error (MSE).

- Pointwise squared error: 
  $$
  \text{sqe}_i^{\text{raw}} = (y_i - \hat \mu_i)^2
  $$
- Point prediction:
$$
\hat \mu_i \approx
\begin{cases}
\frac{1}{S}\sum_{s=1}^S \mu_i^{(s)}, & \mu_i^{(s)} = \mathbb{E}[\tilde y_i \mid \theta^{(s)}] \quad \text{(within-sample)}\\[6pt]
\sum_{s=1}^S \tilde w_i^{(s)}\,\mu_i^{(s)}, & \text{(PSIS-LOO held-out approximation)}
\end{cases}
$$
- Estimates:
$$
\text{mse}^{\text{raw}} = \frac{1}{n}\sum_{i=1}^n \text{sqe}_i^{\text{raw}},\quad
\text{rmse}^{\text{raw}} = \sqrt{\text{mse}^{\text{raw}}}
$$
- Standard Errors:
  - For MSE:
    $$
    se(\text{mse}^{\text{raw}}) =
    \sqrt{\frac{\widehat{\text{Var}}\big(\text{sqe}_i^{\text{raw}}\big)}{n}}
    $$
  - For RMSE (delta method):
    $$
    se(\text{rmse}^{\text{raw}}) \approx \frac{se(\text{mse}^{\text{raw}})}{2\,\text{rmse}^{\text{raw}}}
    $$
    If $\text{rmse} = 0$, set $se(\text{rmse}) = 0$.

```{r}
rmse_ws <- rmse(
  y = res_sleep$y, 
  mupred = res_sleep$mupred
)
rmse_loo <- rmse(
  y = res_sleep$y, 
  mupred = res_sleep$mupred_loo
)
rmse_kfold <- rmse(
  y = res_sleep$y, 
  mupred = res_sleep$mupred_kfold
)

message("within-sample rmse:")
invisible(str(rmse_ws))
message("psis-loo rmse:")
invisible(str(rmse_loo))
message("kfold rmse:")
invisible(str(rmse_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(rmse_ws, rmse_loo, rmse_kfold, "squared-error_i", show_estimate = FALSE)
```

```{r}
mse_ws <- mse(
  y = res_sleep$y, 
  mupred = res_sleep$mupred
)
mse_loo <- mse(
  y = res_sleep$y, 
  mupred = res_sleep$mupred_loo
)
mse_kfold <- mse(
  y = res_sleep$y, 
  mupred = res_sleep$mupred_kfold
)
  
message("within-sample mse:")
invisible(str(mse_ws))
message("psis-loo mse:")
invisible(str(mse_loo))
message("kfold mse:")
invisible(str(mse_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(mse_ws, mse_loo, mse_kfold, "squared-error_i")
```

## Coefficient of determination (`r2`)
R2 summarizes how well point predictions $\hat\mu_i$ explain the variation in 
observed outcomes $y_i$. We use the same point predictions, $\hat\mu_i$, as in 
the mae/rmse sections (i.e., the posterior predictive mean per observation). We 
define a predictive R2 compatible with cross-validation, not Bayesian-`R^2` 
based on a model’s residual variance parameter.

- Definition:
  $$
  R^2 = 1 - \frac{\sum_{i=1}^n (y_i - \hat y_i)^2}{\sum_{i=1}^n (y_i - \bar y)^2},\quad \bar y = \frac{1}{n}\sum_{i=1}^n y_i
  $$
  with $\hat y_i = \hat \mu_i$ formed as within-sample means, PSIS-LOO 
  reweighted means, or K-fold held-out means.
- Equivalently in mean form:
  $$
  \widehat{\text{mse}} = \frac{1}{n}\sum_i (y_i - \hat y_i)^2,\quad
  \widehat{\text{mse}}(y) = \frac{1}{n}\sum_i (y_i - \bar y)^2,\quad
  \widehat{R}^2 = 1 - \frac{\widehat{\text{mse}}}{\widehat{\text{mse}}(y)}
  $$
- Standard error (first‑order delta method):
  $$
  se(\widehat{R}^2) = \left[
  se(\widehat{\text{mse}})^2
  - 2\,\frac{\widehat{\text{mse}}}{\widehat{\text{mse}}(y)}\,\frac{\widehat{\text{Cov}}\big((y_i-\hat y_i)^2,\,(y_i-\bar y)^2\big)}{n}
  + \frac{\widehat{\text{mse}}^2}{\widehat{\text{mse}}(y)^2}\,\frac{\widehat{\text{Var}}\big((y_i-\bar y)^2\big)}{n}
  \right]^{1/2}\cdot \frac{1}{\widehat{\text{mse}}(y)}
  $$
  where $\widehat{\text{mse}} = \frac{1}{n}\sum_{i=1}^n \text{sqe}_i$ and $\widehat{\text{mse}}(y) = \frac{1}{n}\sum_{i=1}^n (y_i - \bar y)^2$.

```{r}
r2_ws <- r2(
  y = res_sleep$y, 
  mupred = res_sleep$mupred
)
r2_loo <- r2(
  y = res_sleep$y, 
  mupred = res_sleep$mupred_loo
)
r2_kfold <- r2(
  y = res_sleep$y, 
  mupred = res_sleep$mupred_kfold
)

message("within-sample r2:")
invisible(str(r2_ws))
message("psis-loo r2:")
invisible(str(r2_loo))
message("kfold r2:")
invisible(str(r2_kfold))
```

```{r, fig.height=3, fig.width=7, fig.align="center"}
plot_ptw(r2_ws, r2_loo, r2_kfold, "squared-error_i", show_estimate = FALSE)
```