---
title: "Computing predictive performance measures"
author: "Florence Bockting"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Computing predictive performance measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = params$EVAL,
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)

library(loo)
library(ggplot2)
```

```{r load-libraries, eval=FALSE}
library(loo)
library(ggplot2)
```

```{r, child="../children/LOAD-BRMS-GITHUB.txt", eval=params$EVAL}
```

## Introduction

After fitting a Bayesian model, you often want predictive performance measures
that are comparable across evaluation strategies: in-sample, PSIS-LOO,
k-fold, and holdout data. The **loo** package provides a unified
workflow through four entry points:

- `insample_pred_measure()` for in-sample evaluation
- `loo_pred_measure()` for PSIS-LOO cross-validation
- `kfold_pred_measure()` for k-fold cross-validation
- `test_pred_measure()` for holdout / new data

Each entry point can compute any supported measure in a single call when you
supply the required inputs and pass the `measure` argument. 
For the argument `measure` you can either specify a vector with predefined measure names whereby supported measures
are:

- **Density scores:** `elpd`, `mlpd`, `ic`
- **Distributional scores:** `rps`, `srps`, `brier`
- **Point metrics:** `mae`, `mse`, `rmse`, `r2`
- **Classification metrics:** `acc`, `bacc`

Base measure `elpd` is always included automatically when `ylp` is supplied;
request `ic`, `mlpd`, or other density scores via `measure`. Pass any
additional names from the list above through the `measure` argument. You can 
also obtain a list of supported measures via `supported_measures_list`:

```{r supported-measures}
supported_measures_list
```

When you want to compute a measure that is not yet supported, you can pass a
custom function to the `measure` argument (see [Passing a custom function for a
measure](#passing-a-custom-function-for-a-measure) below). In this case, the
argument `measure` should be a list, whereby you can pass both predefined
measure names and custom functions (e.g., `list("rps", "my_fun" = my_fun)`).
A custom function declares its inputs by name. `loo` supplies whichever of `y`,
`ypred`, `mupred`, `ylp` and `log_weights` appear in its formals
(see [Input requirements by measure](#input-requirements-by-measure) below),
and you can pass additional arguments through `control`.

For measure definitions and formulas, see the companion article
[Overview of scores and metrics](overview-measures.html).

Besides computing all desired measures in one call, `pred_measure()` lets you 
**add** measures to an existing result without recomputing what is already 
stored (see [Adding measures incrementally](#adding-measures-incrementally-optional) below).

**Prerequisites.** 

This tutorial assumes a fitted **brms** model and familiarity
with `brms::log_lik()`, `brms::posterior_predict()`, and
`brms::posterior_epred()`.

**What you will learn.**

1. Compute all desired measures in one call to an entry point.
2. Optionally accumulate measures incrementally with `pred_measure()`.
3. Repeat the workflow under LOO, k-fold, and out-of-sample evaluation.
4. Compare results across evaluation strategies.

This article fits five small **brms** models. Each uses 2 chains and 400
iterations, so the whole article renders in a few minutes.

```{r fit-models, results="hide"}
SEED <- 42
set.seed(SEED)

# Continuous outcome: reaction times from lme4::sleepstudy
data("sleepstudy", package = "lme4")
sleepstudy <- subset(sleepstudy, Days >= 2)
sleepstudy$Days <- sleepstudy$Days - 2
sleepstudy$y <- sleepstudy$Reaction

prior_sleep <- brms::prior(normal(200, 100), class = b, coef = "Intercept") +
  brms::prior(normal(0, 20), class = b, coef = "Days") +
  brms::prior(exponential(0.02), class = sigma)

fit_sleep <- brms::brm(
  y ~ 0 + Intercept + Days,
  data = sleepstudy,
  family = gaussian(),
  prior = prior_sleep,
  chains = 2, iter = 400, seed = SEED, refresh = 0
)

# The same model, fitted without five held-out subjects
test_subjects <- sample(unique(sleepstudy$Subject), size = 5)
train_data <- subset(sleepstudy, !Subject %in% test_subjects)
test_data <- subset(sleepstudy, Subject %in% test_subjects)

fit_sleep_train <- brms::brm(
  y ~ 0 + Intercept + Days,
  data = train_data,
  family = gaussian(),
  prior = prior_sleep,
  chains = 2, iter = 400, seed = SEED, refresh = 0
)

# Binary outcome
df_binary <- data.frame(y = rbinom(50, 1, 0.3))

fit_binary <- brms::brm(
  y ~ 1,
  data = df_binary,
  family = brms::bernoulli(),
  chains = 2, iter = 400, seed = SEED, refresh = 0
)

# Count outcome: rstanarm::roaches
data("roaches", package = "rstanarm")
roaches$sqrt_roach1 <- sqrt(roaches$roach1)

fit_roaches <- brms::brm(
  y ~ sqrt_roach1 + treatment + senior + offset(log(exposure2)),
  data = roaches,
  family = poisson(),
  prior = brms::prior(normal(0, 1), class = b),
  chains = 2, iter = 400, seed = SEED, refresh = 0
)

# Categorical outcome: datasets::iris
data("iris")
iris$y <- iris$Species

fit_iris <- brms::brm(
  y ~ Sepal.Length + Sepal.Width,
  data = iris,
  family = brms::categorical(),
  chains = 2, iter = 400, seed = SEED, refresh = 0
)
```

## Quick start

The sleep-study model from **lme4** predicts reaction time (ms) from days of
sleep deprivation. The shortest path from a fitted model to printed measures is
a single call to an entry point:

```{r quick-start}
sleep_measure <- insample_pred_measure(
  ylp = brms::log_lik(fit_sleep),
  y = fit_sleep$data$y,
  ypred = brms::posterior_predict(fit_sleep),
  measure = c("rps", "srps")
)

print(sleep_measure)
```

Base measure `elpd` is always computed when `ylp` is supplied; request `ic`
or other density scores via `measure`. The rest of this tutorial explains how inputs map to measures and
how to switch evaluation modes.

## Core concepts

### Object structure

Every function returns a **pred_measure** object which is a list with:

| Component | Description |
| :--- | :--- |
| `$estimates` | Summary estimate and SE for each measure |
| `$pointwise` | Observation-level contributions (one column per measure) |
| `$diagnostics` | Pareto-k diagnostics (LOO always; k-fold only when the `kfold` object supplies them) |
| `$log_weights` | Log weights used for LOO (LOO-CV only) |

The `print()` method shows `$estimates` in a readable table. Use
`str(result)` when you need the full structure.

Every object also carries attributes that identify the evaluation mode and data
dimensions. The table below summarizes what differs across approaches; all
objects share `attr(result, "dims")` (posterior draws × observations) and
`attr(result, "source")` (one of `"insample"`, `"loo"`, `"kfold"`, `"test"`).

| | In-sample | PSIS-LOO | K-fold | Holdout (Test) |
| :--- | :--- | :--- | :--- | :--- |
| **Class** | `insample_pred_measure` | `loo_pred_measure` (+ classes from `loo` object) | `kfold_pred_measure` (+ classes from `kfold` object) | `test_pred_measure` |
| **List slots** | `$estimates`, `$pointwise` | `$estimates`, `$pointwise`, `$diagnostics`, `$log_weights`; (`$psis_object`) | `$estimates`, `$pointwise`; (`$diagnostics`) | `$estimates`, `$pointwise` |
| **Measure suffix** | none (`elpd`) | `_loo` (`elpd_loo`, `p_loo`, …) | `_kfold` (`elpd_kfold`, `p_kfold`) | `_test` (`elpd_test`, …) |
| **Inherited attrs** | -- | `yhash`, `model_name` from `loo` | `K`, `Ksub`, `folds`, `fold_type`, `joint`, `yhash`, `model_name` from `kfold` | -- |

For holdout objects, `dims` reflects the test-set size (from `ylp_test`), not
the training data. LOO objects expose Pareto-k values in
`$diagnostics$pareto_k` to assess the reliability of the PSIS weights. K-fold
objects expose the same slot only when the `kfold` object carries one; recent
`brms::kfold()` versions do. That k diagnoses the Monte Carlo estimate of each
fold-wise log predictive density, because k-fold uses no importance weights.

### Input requirements by measure

Supply the inputs each measure needs in a single entry-point call (see the
[workflow overview](#workflow-overview) below). When adding measures
incrementally, pass an existing object as `predperf` to `pred_measure()` so
base estimates and weights are reused.

| Measure | `ylp` | `y` | `ypred` | `mupred` |
| :---: | :---: | :---: | :---: | :---: |
| `elpd`, `mlpd`, `ic` | ✓ | | | |
| `rps`, `srps` | | ✓ | ✓ | |
| `acc`, `bacc` | | ✓ | | ✓ |
| `mae`, `mse`, `rmse`, `r2` | | ✓ | | ✓ |

- **`ylp`**: pointwise log predictive density matrix (`S` draws × `n` observations),
  typically from `brms::log_lik(fit)`.
- **`ypred`**: posterior predictive draws from `brms::posterior_predict()`.
- **`mupred`**: posterior expected values from `brms::posterior_epred()`.

### Two ways to compute measures

| Approach | When to use | Example |
| :--- | :--- | :--- |
| **One-shot** | You know all measures upfront | `loo_pred_measure(..., measure = c("rps", "r2"))` |
| **Incremental** | Explore interactively or add measures later | `pred_measure(..., predperf = existing, measure = "r2")` |

### Workflow overview

| Evaluation mode | Entry point | Primary inputs |
| :--- | :--- | :--- |
| In-sample | `insample_pred_measure()` | `ylp`, plus `y`, `ypred`, `mupred` as needed | 
| PSIS-LOO | `loo_pred_measure()` | `loo`, or `ylp` + `psis_object`, or `ylp` alone |
| K-fold | `kfold_pred_measure()` | `kfold`, plus `y`, `ypred`, `mupred` as needed |
| Holdout | `test_pred_measure()` | `ylp_test` (for `elpd_test`), plus `y`, `ypred`, `mupred` as needed |

Pass all required inputs and the full `measure` vector to the entry point for
your evaluation mode. Use `pred_measure()` only when you want to extend an
existing result (see below).

## Continuous outcome: sleep study

We use the pre-fitted Gaussian model on the sleep study throughout this section.

### In-sample evaluation

Pass all required inputs and the measures you want to `insample_pred_measure()`:

```{r insample-all}
ylp_sleep <- brms::log_lik(fit_sleep)
ypred_sleep <- brms::posterior_predict(fit_sleep)
mupred_sleep <- brms::posterior_epred(fit_sleep)

sleep_measure <- insample_pred_measure(
  ylp = ylp_sleep,
  y = fit_sleep$data$y,
  ypred = ypred_sleep,
  mupred = mupred_sleep,
  measure = c("rps", "srps", "r2")
)
print(sleep_measure)
```

`elpd` is always computed when `ylp` is supplied; it sums the expected log
pointwise predictive density across observations. Request `ic` via `measure`
(`ic` equals `-2 × elpd`). Additional rows in the table correspond to the measures
passed in `measure`.

```{r pointwise-crps, fig.cap="Pointwise CRPS contributions across observations (in-sample)."}
crps_pw <- sleep_measure$pointwise[, "rps"]
ggplot(
  data = data.frame(obs = seq_along(crps_pw), crps = crps_pw)
) +
  geom_dotplot(aes(crps), dotsize = 0.5) +
  labs(
    x = "CRPS_i",
    title = "Dotplot for in-sample pointwise CRPS"
  ) +
  theme_minimal()
```

The entire structure of the returned object with corresponding 
attributes is shown in the following

```{r}
str(sleep_measure)
```

#### Passing a custom function for a measure

It is also possible to pass a custom function for any measure. Name its
arguments from `y`, `ypred`, `mupred`, `ylp` and `log_weights`; `loo` supplies
each name that appears in the formals. You can pass additional arguments
through `control`:

```{r}
# custom scoring rule
hamming_loss <- function(y, ypred, log_weights = NULL) {
  if (is.null(log_weights)) {
    hamming_i <- colMeans(sweep(ypred, 2, y, "!="))
  } else {
    lw <- sweep(log_weights, 2, matrixStats::colLogSumExps(log_weights), "-")
    w <- exp(lw)
    hamming_i <- colSums(w * sweep(ypred, 2, y, "!="))
  }

  list(
    estimate = mean(hamming_i),
    se = sqrt(var(hamming_i) / length(hamming_i)),
    pointwise = hamming_i
  )
}
# Hamming loss is on a loss scale (lower is better). Declaring that lets
# `model_compare()` report its differences on the utility scale, as it does for
# built-in loss measures, and rank models in the right direction.
attr(hamming_loss, "measure_loss") <- TRUE
```

```{r insample-custom-function}
(insample_pred_measure(
  ylp = brms::log_lik(fit_binary),
  y = fit_binary$data$y,
  ypred = brms::posterior_predict(fit_binary),
  measure = list("brier", hamming = hamming_loss)
))
```

#### Adding measures incrementally

If you already computed measures and want to add more without
recomputing them, pass the existing object as `predperf` (**pred**icted 
**perf**ormance measure) to `pred_measure()`:

```{r insample-incremental}
(pred_measure(
  y = fit_sleep$data$y,
  mupred = mupred_sleep,
  predperf = sleep_measure,
  measure = c("rmse", "mse")
))
```

This two-step pattern is useful when exploring measures interactively.

### PSIS-LOO cross-validation

In-sample `elpd` uses posterior draws from $p(\theta \mid y)$. Comparing
those predictions to the same data is **optimistically biased** for future or
otherwise unseen observations: the model has already seen the data it is
predicting ([Vehtari's CV-FAQ](https://users.aalto.fi/~ave/CV-FAQ.html),
Section 4).

**Cross-validation** estimates out-of-sample predictive performance by
holding out part of the data, refitting (or reweighting) without that part,
and scoring the held-out points. In **leave-one-out cross-validation (LOO-CV)**,
each observation $y_i$ is predicted using a model fit to $y_{-i}$. The target
quantity is the expected log pointwise predictive density (elppd), estimated by
summing held-out log predictive densities across observations (see the
[overview of scores and metrics](overview-measures.html) article for the full
definitions of within-sample, LOO, PSIS-LOO, and k-fold variants).

**PSIS-LOO** approximates exact LOO-CV without refitting the model $n$ times.
It reweights the full-data posterior draws $\theta^{(s)} \sim p(\theta \mid y)$
with Pareto-smoothed importance sampling (PSIS) weights so that each
observation $i$ is scored as if it had been left out. The resulting
`elpd_loo` is an estimate of how well the model would predict new reaction
times.

Compute LOO measures in one call:

```{r loo-all}
loo_sleep <- brms::loo(fit_sleep, save_psis = TRUE)

loo_measure <- loo_pred_measure(
  loo = loo_sleep,
  y = fit_sleep$data$y,
  ypred = ypred_sleep,
  mupred = mupred_sleep,
  measure = c("rps", "srps", "r2")
)
print(loo_measure)
```

The estimates carry a `_loo` suffix (e.g. `elpd_loo`). Base LOO summaries
include `elpd_loo` and `p_loo` (effective number of parameters: the
non-cross-validated log posterior predictive density minus `elpd_loo`).
Request `ic_loo` via `measure = "ic"`.
Furthermore, we get an additional entry with PSIS diagnostics in `$diagnostics`.

```{r loo-pareto-k}
str(loo_measure)
```

**Alternative LOO inputs.** 

`loo_pred_measure()` accepts three equivalent input patterns:

| Input | Example |
| :--- | :--- |
| Precomputed `loo` object | `loo_pred_measure(loo = loo_sleep, measure = ...)` |
| `ylp` + `psis_object` | `loo_pred_measure(ylp = ylp, psis_object = psis_obj, measure = ...)` |
| `ylp` only (PSIS computed internally) | `loo_pred_measure(ylp = ylp, measure = ...)` |

To add measures incrementally after a base-only LOO call, pass the existing
object as `predperf` to `pred_measure()`. When
accumulating after computing only `elpd`, use `save_psis = TRUE` in
`loo_pred_measure()` so PSIS weights are stored for subsequent measures.

Compare the in-sample and LOO ELPD values directly:

```{r elpd-compare-table}
data.frame(
  method = c("In-sample", "PSIS-LOO"),
  elpd = c(
    sleep_measure$estimates["elpd", "Estimate"],
    loo_measure$estimates["elpd_loo", "Estimate"]
  ),
  se = c(
    sleep_measure$estimates["elpd", "SE"],
    loo_measure$estimates["elpd_loo", "SE"]
  )
)
```

### K-fold cross-validation

K-fold CV holds out groups of observations rather than single points.
The same computation pattern applies to this approach: pass all inputs and 
measures to `kfold_pred_measure()` with a `kfold` object from `brms::kfold()`:

```{r kfold-measures}
kfold_sleep <- brms::kfold(fit_sleep, save_fits = TRUE)
mupred_kfold <- brms::kfold_predict(kfold_sleep, method = "fitted")$yrep
ypred_kfold <- brms::kfold_predict(kfold_sleep, method = "predict")$yrep

kfold_measure <- kfold_pred_measure(
  y = fit_sleep$data$y,
  ypred = ypred_kfold,
  mupred = mupred_kfold,
  kfold = kfold_sleep,
  measure = c("rps", "srps", "r2")
)
print(kfold_measure)
```

Measure names carry a `_kfold` suffix (e.g. `elpd_kfold`, `rps_kfold`).
The returned object inherits the attributes from the `brms::kfold` object:

```{r}
str(kfold_measure)
```

### Holdout evaluation

For genuinely new data, use `test_pred_measure()` in the same one-shot way.
The inputs come from `fit_sleep_train`, which never saw the five held-out
subjects:

```{r test-measures}
test_measure <- test_pred_measure(
  y = test_data$y,
  ypred = brms::posterior_predict(fit_sleep_train, newdata = test_data),
  mupred = brms::posterior_epred(fit_sleep_train, newdata = test_data),
  ylp = brms::log_lik(fit_sleep_train),
  ylp_test = brms::log_lik(fit_sleep_train, newdata = test_data),
  measure = c("rps", "srps", "r2")
)
print(test_measure)
```

Holdout measures use a `_test` suffix and reflect performance on the held-out
subjects only. The base summary `elpd_test` comes from `ylp_test`.

## Other outcome types

The same pattern applies to non-Gaussian models. Each subsection uses
a pre-fitted **brms** model.

### Binary outcomes

```{r binary-example}
binary_measure <- insample_pred_measure(
  ylp = brms::log_lik(fit_binary),
  y = fit_binary$data$y,
  mupred = brms::posterior_epred(fit_binary),
  measure = c("acc", "bacc")
)
print(binary_measure)
```

### Count outcomes

```{r count-example}
ypred_roaches <- brms::posterior_predict(fit_roaches)

roaches_measure <- insample_pred_measure(
  ylp = brms::log_lik(fit_roaches),
  y = fit_roaches$data$y,
  ypred = ypred_roaches,
  measure = c("rps", "srps")
)
print(roaches_measure)
```

### Categorical outcomes

```{r categorical-example}
iris_measure <- insample_pred_measure(
  ylp = brms::log_lik(fit_iris),
  y = as.integer(fit_iris$data$y),
  mupred = brms::posterior_epred(fit_iris),
  measure = c("acc", "bacc")
)
print(iris_measure)
```

See the [overview of scores and metrics](overview-measures.html) article for which
measures apply to each outcome type.

## Further reading

- `?supported_measures_list` — character vector of all supported measure names
- `?pred_measure` — optional helper for adding measures to an existing predperf
  object
- `?insample_pred_measure`, `?loo_pred_measure`, `?kfold_pred_measure`,
  `?test_pred_measure` — entry points for each evaluation mode
- [Overview of scores and metrics](overview-measures.html) — measure
  definitions, formulas, and orientation (higher vs lower is better)