---
title: "Predictive schemes with pred_measure()"
author: "Florence Bockting"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    css:
      - !expr system.file("rmarkdown/templates/html_vignette/resources/vignette.css", package = "rmarkdown")
      - styles.css
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Predictive schemes with pred_measure()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="../children/SEE-ONLINE.txt", eval = if (isTRUE(exists("params"))) !params$EVAL else TRUE}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = params$EVAL,
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

```{r, child="../children/LOAD-BRMS-GITHUB.txt", eval=params$EVAL}
```

## Introduction

How well does a fitted model predict data that were not used to fit it? The
formula reference (Vehtari et al., 2026, Sec. 1) divides this question into
three questions:

1. **What counts as a good prediction?** This is the *predictive measure*, for
   example the log score or the RMSE.
2. **Which predictions are evaluated?** This is the *predictive scheme*. It
   sets the observations that train the model before observation $i$ is
   evaluated.
3. **How are the predictions computed?** From posterior draws, from refits, or
   with importance weights.

The [overview of predictive measures](overview-measures.html) answers the first
question. This article answers the second and the third. It shows how the
`pred_measure()` family computes any measure under four predictive schemes:
in-sample, PSIS-LOO, K-fold, and test data.

::: {.callout .callout-note}
**What this article does not cover**

+ **Measure definitions.** See the
  [overview of predictive measures](overview-measures.html).
+ **Model comparison.** See the [model comparison](model-comparison.html)
  article.
+ **Leave-one-group-out and leave-future-out schemes.** The `pred_measure()`
  family has no entry point for them yet.
:::

## Training data and evaluation data {#data-sets}

A predictive scheme is a rule that splits the observed data. Let
$y = (y_1, \ldots, y_n)$ be the observed data. In the example of this article,
$y_i$ is the reaction time of one subject on one day, and $n = 144$.

A predictive measure compares a prediction for $y_i$ with the observed $y_i$.
Before **loo** can compute it, two questions need an answer:

1. **Which observations are scored?** These observations are the *evaluation
   data*. The formula reference writes their indices as the evaluation set
   $\mathcal{E}$.
2. **Which observations fit the model that predicts $y_i$?** These
   observations are the *training data* of observation $i$. Their indices are
   the training set $\mathcal{T}(i)$.

Both sets hold indices of observations. They do not hold parameters or draws.
Two examples:

+ **Leave-one-out.** Every observation is scored, so
  $\mathcal{E} = \{1, \ldots, n\}$. All observations except $y_i$ fit the
  model that predicts $y_i$, so $\mathcal{T}(i) = \{1, \ldots, n\} \setminus \{i\}$.
  The training data change with $i$.
+ **Test data.** Only the 40 observations of five held-out subjects are scored,
  so $\mathcal{E} = \mathcal{S}$. The other 104 observations fit the model, so
  $\mathcal{T}(i) = \mathcal{R}$ for every $i$.

With the two sets, the estimator of an additive measure is
(Vehtari et al., 2026, Eq. 1.7)

$$
\hat U = \sum_{i \in \mathcal{E}} u_i\!\left(p(y_i \mid y_{\mathcal{T}(i)}), y_i\right).
$$

$\mathcal{T}(i)$ sets the predictive distribution of $y_i$. $\mathcal{E}$ sets
the observations in the sum. Other sets answer other predictive questions
(Vehtari et al., 2026, Sec. 1.2).

## Overview of the schemes

**loo** has one entry point for each scheme
(Vehtari et al., 2026, Table 1, Eqs. 2.1–2.4):

| Scheme | Training data $\mathcal{T}(i)$ | Evaluation data $\mathcal{E}$ | Predictive distribution $p^q(y_i)$ | Weights $w_i^s$ | Entry point |
| :--- | :--- | :--- | :--- | :--- | :--- |
| [in-sample](#in-sample) | $\{1, \ldots, n\}$ | $\{1, \ldots, n\}$ | $p(y_i \mid y)$ | $1/S$ | `insample_pred_measure()` |
| [PSIS-LOO](#psis-loo) | $\{1, \ldots, n\} \setminus \{i\}$ | $\{1, \ldots, n\}$ | $p(y_i \mid y_{-i})$ | PSIS weights | `loo_pred_measure()` |
| [K-fold](#k-fold) | $\{1, \ldots, n\} \setminus \mathcal{I}_{k(i)}$ | $\{1, \ldots, n\}$ | $p(y_i \mid y_{-\mathcal{I}_{k(i)}})$ | $1/S$ | `kfold_pred_measure()` |
| [test data](#test-data) | $\mathcal{R}$ | $\mathcal{S}$ | $p(y_i \mid y_{\mathcal{R}})$ | $1/S$ | `test_pred_measure()` |
A fifth function, `pred_measure()`, [adds measures](#add-measures) to an
existing result of any scheme.

::: {.callout .callout-note}
**One formula, four schemes**

A measure uses the same formula in every scheme. Only the draws and the
weights change (Vehtari et al., 2026, Eq. 2.5):

$$
\hat p^q(y_i) = \sum_{s=1}^S w_i^s\, p(y_i \mid \theta^s) .
$$

Thus `measure = "rps"` gives $\widehat{\text{rps}}^{\text{in}}$ in
`insample_pred_measure()` and $\widehat{\text{rps}}^{\text{LOO}}$ in
`loo_pred_measure()`.
:::

## Notation

The article uses the notation of the formula reference. The table maps each
symbol to an argument of the entry points or to a field of their result.

| Symbol | Meaning | In **loo** |
| :--- | :--- | :--- |
| $q$ | predictive scheme | the entry point; `attr(x, "source")` |
| $\mathcal{E}$ | indices of the evaluation data | the observations in `y` and in the columns of the input matrices |
| $\mathcal{T}(i)$ | indices of the training data for observation $i$ | the data of the fit that gave the draws |
| $\mathcal{R}$, $\mathcal{S}$ | training set and test set of a single split | `train_data`, `test_data` |
| $\mathcal{I}_k$ | observations in fold $k$ | `attr(x, "folds")` |
| $\theta^s$ | posterior draw $s$, $s = 1, \ldots, S$ | rows of the input matrices |
| $w_i^s$ | normalized weight of draw $s$ | `log_weights` of the PSIS object for PSIS-LOO; $1/S$ otherwise |
| $\ell_i^s = \log p(y_i \mid \theta^s)$ | log density of $y_i$ under draw $s$ | `ylp`; `ylp_test` for test data |
| $\tilde y_i^s$ | posterior predictive draw | `ypred` |
| $\mathbb{E}[\tilde y_i \mid \theta^s]$ | predictive mean under draw $s$ | `mupred` |
| $u_i^q$ | pointwise term of a measure under scheme $q$ | `$pointwise` |
| $\hat U^q$, $\hat u^q$ | estimate on the sum scale and the mean scale | `$estimates[, "Estimate"]` |
| $\widehat{\text{SE}}$ | standard error of the estimate | `$estimates[, "SE"]` |

**Inputs by measure.** Each measure needs one kind of draw. Pass only the inputs
that the requested measures need:

| Measure | `ylp` | `ypred` | `mupred` |
| :--- | :---: | :---: | :---: |
| `elpd`, `mlpd`, `ic` | ✓ | | |
| `rps`, `srps`, `brier` | | ✓ | |
| `mae`, `mse`, `rmse`, `r2`, `acc`, `bacc` | | | ✓ |

All measures except the log score also need the observations `y`. The list of
built-in names is in `supported_measures_list`:

```{r load-libraries}
library(loo)
library(ggplot2)
theme_set(theme_minimal())

supported_measures_list
```

**Inputs by scheme.** The schemes differ in the fit that gives the draws:

| Scheme | `ylp` | `ypred`, `mupred` | Extra input |
| :--- | :--- | :--- | :--- |
| in-sample | `log_lik(fit)` | `posterior_predict(fit)`, `posterior_epred(fit)` | |
| PSIS-LOO | as in-sample | as in-sample | `loo`, or `psis_object` |
| K-fold | from `kfold` | `kfold_predict(kfold)$yrep` | `kfold` |
| test data | `ylp_test = log_lik(fit_train, newdata = test_data)` | `posterior_predict(fit_train, newdata = test_data)`, … | |

## Data and model

**Prediction task.** The `sleepstudy` data (**lme4**) come from a study of sleep
restriction (Belenky et al., 2003). Eighteen subjects slept at most three hours
each night. The outcome `y` is the average reaction time (ms) of a subject on
one day. Days 0 and 1 were adaptation and training, so the article keeps days
2 to 9. The predictor `Days` counts the days of sleep restriction, from 0 to 7.
This gives 144 observations.

The model is a normal regression of `y` on `Days`. It has no subject effects.
Fitting is not the topic of this article. The model code is in the
[appendix](#appendix-model-code).

| Object | Data | Used for |
| :--- | :--- | :--- |
| `fit_sleep` | all 18 subjects, 144 observations | in-sample, PSIS-LOO, K-fold |
| `fit_sleep_train` | 13 subjects, 104 observations ($\mathcal{R}$) | test data |
| `test_data` | the other 5 subjects, 40 observations ($\mathcal{S}$) | test data |

Each fit uses 2 chains with 200 draws after warmup, so $S = 400$.

```{r fit-models, echo=FALSE, results="hide"}
SEED <- 42
set.seed(SEED)

data("sleepstudy", package = "lme4")
sleepstudy <- subset(sleepstudy, Days >= 2)
sleepstudy$Days <- sleepstudy$Days - 2
sleepstudy$y <- sleepstudy$Reaction

prior_sleep <- brms::prior(normal(200, 100), class = b, coef = "Intercept") +
  brms::prior(normal(0, 20), class = b, coef = "Days") +
  brms::prior(exponential(0.02), class = sigma)

# all subjects
fit_sleep <- brms::brm(
  y ~ 0 + Intercept + Days,
  data = sleepstudy,
  family = gaussian(),
  prior = prior_sleep,
  chains = 2, iter = 400, seed = SEED, refresh = 0
)

# five subjects held out as test data
test_subjects <- sample(unique(sleepstudy$Subject), size = 5)
train_data <- subset(sleepstudy, !Subject %in% test_subjects)
test_data <- subset(sleepstudy, Subject %in% test_subjects)

fit_sleep_train <- update(fit_sleep, newdata = train_data, seed = SEED, refresh = 0)
```
The draws of `fit_sleep` are the input for the in-sample and the PSIS-LOO
scheme. All schemes request the same measures:

```{r inputs}
y_sleep <- fit_sleep$data$y
ylp_sleep <- brms::log_lik(fit_sleep)
ypred_sleep <- brms::posterior_predict(fit_sleep)
mupred_sleep <- brms::posterior_epred(fit_sleep)

measures <- c("elpd", "mlpd", "rps", "rmse", "r2")
```

## In-sample {#in-sample}

**Predictive question.** How well does the fitted model describe the data used
to fit it?

**Predictive distribution.** The draws come from the posterior of all data. All
draws get the same weight, and $\mathcal{E} = \{1, \ldots, n\}$
(Vehtari et al., 2026, Eqs. 2.1, 2.5):

$$
\hat p^{\text{in}}(y_i) = \frac{1}{S}\sum_{s=1}^S p(y_i \mid \theta^s),
\qquad \theta^s \sim p(\theta \mid y) .
$$

**Inputs.** `ylp`, `ypred`, and `mupred` from `fit_sleep`, and the
observations `y`.

**Implementation.**

```{r insample}
insample_sleep <- insample_pred_measure(
  y = y_sleep,
  ypred = ypred_sleep,
  mupred = mupred_sleep,
  ylp = ylp_sleep,
  measure = measures
)
insample_sleep
```

**Result.** The rows have no suffix. The result holds `$estimates` and
`$pointwise`. `elpd` is on the sum scale, and the other measures are on the
mean scale (Vehtari et al., 2026, Eq. 2.7). Two attributes identify the result:

```{r insample-attributes}
attributes(insample_sleep)[c("source", "dims")]
```

**Interpretation.** The model has already seen each $y_i$ that it predicts.
Thus the in-sample estimates are optimistic for new data. They are not an
estimate of out-of-sample performance. Use them as a reference point: the
difference to a cross-validation estimate shows how optimistic they are
(Vehtari et al., 2026, Sec. 1.2).

## PSIS-LOO {#psis-loo}

**Predictive question.** How well does the model predict a new observation that
is exchangeable with the observed data?

**Predictive distribution.** Exact leave-one-out cross-validation needs one
refit for each observation, here 144. PSIS-LOO uses the draws of the full-data
fit instead. It gives them Pareto-smoothed importance weights, so that each
$y_i$ is scored as if it was left out (Vehtari et al., 2026, Eqs. 2.2, 2.5;
Vehtari et al., 2024):

$$
\hat p^{\text{LOO}}(y_i) = \sum_{s=1}^S w_i^s\, p(y_i \mid \theta^s),
\qquad \theta^s \sim p(\theta \mid y) .
$$

**Inputs.** The same draws as in-sample. Add the PSIS weights in one of three
ways:

| Input | Call |
| :--- | :--- |
| a `loo` object | `loo_pred_measure(loo = loo_sleep, ...)` |
| `ylp` and a `psis` object | `loo_pred_measure(ylp = ylp_sleep, psis_object = psis_sleep, ...)` |
| `ylp` only | `loo_pred_measure(ylp = ylp_sleep, ...)` |

With `ylp` only, `loo_pred_measure()` computes the weights with `r_eff = 1`.

**Implementation.** Set `save_psis = TRUE` in `loo()`. Without it,
`loo_pred_measure()` stops with an error.

```{r loo}
loo_sleep <- brms::loo(fit_sleep, save_psis = TRUE)

loo_sleep_measure <- loo_pred_measure(
  loo = loo_sleep,
  y = y_sleep,
  ypred = ypred_sleep,
  mupred = mupred_sleep,
  measure = measures
)
loo_sleep_measure
```

**Result.** The rows have the suffix `_loo`. The measure `elpd` also adds
`p_loo`, the difference between the in-sample and the LOO log score
(Vehtari et al., 2026, Eq. 2.19). The result has two more fields:

+ `$diagnostics$pareto_k`: one Pareto $\hat k$ for each observation. A value
  above $\min(1 - 1/\log_{10} S,\, 0.7)$ shows an unreliable weight. For
  $S = 400$ this threshold is 0.62. `print()` shows a table of the $\hat k$
  values when at least one value is above the threshold. Otherwise it prints
  one line.
+ `$log_weights`: the log PSIS weights.

**Plot.** Each point is one observation. The x-axis shows its in-sample log
score, and the y-axis its LOO log score. The vertical distance to the dashed
diagonal is the pointwise $p_{\text{loo}}$. The color shows the Pareto
$\hat k$.

```{r loo-plot}
df_loo <- data.frame(
  insample = insample_sleep$pointwise[, "elpd"],
  loo = loo_sleep_measure$pointwise[, "elpd_loo"],
  pareto_k = loo_sleep_measure$diagnostics$pareto_k
)

ggplot(df_loo, aes(insample, loo, color = pareto_k)) +
  geom_abline(linetype = "dashed") +
  geom_point() +
  scale_color_viridis_c() +
  labs(x = "pointwise in-sample log score",
       y = "pointwise LOO log score",
       color = "Pareto k")
```

**Interpretation.** Points below the diagonal are observations that are harder
to predict when they are left out. The farther a point lies from the diagonal,
the more the observation influences the posterior. Such observations often have
a high Pareto $\hat k$. The model has three parameters. A `p_loo` much larger
than three can show that the model is misspecified (see `?loo-glossary`).

## K-fold {#k-fold}

**Predictive question.** How well does the model predict held-out folds?

**Predictive distribution.** The data are split into $K$ folds
$\mathcal{I}_1, \ldots, \mathcal{I}_K$. The model is fit once for each fold,
without the observations in that fold. This needs $K$ fits, not $n$
(Vehtari et al., 2026, Eqs. 2.3, 2.5):

$$
\hat p^{K\text{-fold}}(y_i) = \frac{1}{S}\sum_{s=1}^S p(y_i \mid \theta^s),
\qquad \theta^s \sim p(\theta \mid y_{-\mathcal{I}_{k(i)}}) .
$$

**Inputs.** **loo** does not refit the model. `kfold_pred_measure()` takes
`elpd_kfold` and `p_kfold` from a `kfold` object, for example from
`brms::kfold()`. For the other measures, pass the draws of the held-out fits
from `brms::kfold_predict()`: `method = "predict"` gives `ypred`, and
`method = "fitted"` gives `mupred`.

**Implementation.** Set `save_fits = TRUE` in `brms::kfold()`. Without the
fits, `brms::kfold_predict()` cannot compute the held-out draws.

```{r kfold, results="hide"}
set.seed(SEED)
kfold_sleep <- brms::kfold(fit_sleep, K = 10, save_fits = TRUE)
ypred_kfold <- brms::kfold_predict(kfold_sleep, method = "predict")$yrep
mupred_kfold <- brms::kfold_predict(kfold_sleep, method = "fitted")$yrep
```

```{r kfold-measures}
kfold_sleep_measure <- kfold_pred_measure(
  y = y_sleep,
  ypred = ypred_kfold,
  mupred = mupred_kfold,
  kfold = kfold_sleep,
  measure = measures
)
kfold_sleep_measure
```

**Result.** The rows have the suffix `_kfold`. The result takes the attributes
`K`, `folds`, `fold_type`, `joint`, `yhash`, and `model_name` from the `kfold`
object. It has `$diagnostics` only when the `kfold` object has them; recent
versions of `brms::kfold()` do. Their Pareto $\hat k$ describes the Monte Carlo
estimate of the log predictive density in each fold. K-fold uses no importance
weights.

**Interpretation.** With $K = 10$, each fit uses 90% of the data. Thus K-fold
and PSIS-LOO answer almost the same question, and their estimates are usually
close. Use K-fold when PSIS-LOO shows many high Pareto $\hat k$ values.

`brms::kfold()` also accepts `folds = "grouped"` with `group = "Subject"`. Then
each fold holds out whole subjects, and the predictive question changes to new
subjects.

## Test data {#test-data}

**Predictive question.** How well does the model predict data that were
withheld from the start?

**Predictive distribution.** The data are split once into a training set
$\mathcal{R}$ and a test set $\mathcal{S}$. The model is fit to $\mathcal{R}$,
and $\mathcal{E} = \mathcal{S}$ (Vehtari et al., 2026, Eqs. 2.4, 2.5):

$$
\hat p^{\text{test}}(y_i) = \frac{1}{S}\sum_{s=1}^S p(y_i \mid \theta^s),
\qquad \theta^s \sim p(\theta \mid y_{\mathcal{R}}),
\qquad i \in \mathcal{S} .
$$

**Inputs.** The draws of `fit_sleep_train`, evaluated at `test_data` with
`newdata`. Pass the log densities of the test observations as `ylp_test`, not
as `ylp`. The log-score measures `elpd`, `mlpd`, and `ic` need `ylp_test`.

**Implementation.**

```{r test}
test_sleep_measure <- test_pred_measure(
  y = test_data$y,
  ypred = brms::posterior_predict(fit_sleep_train, newdata = test_data),
  mupred = brms::posterior_epred(fit_sleep_train, newdata = test_data),
  ylp_test = brms::log_lik(fit_sleep_train, newdata = test_data),
  measure = measures
)
test_sleep_measure
```

**Result.** The rows have the suffix `_test`. The attribute `dims` gives the
size of the test set, 40 observations, not the size of the training set.

**Interpretation.** The test set holds five subjects that the model did not
see. Thus the predictive question is about new subjects, not about new
observations of known subjects. The test set is small, so the standard errors
are larger than in the other schemes (Vehtari et al., 2026, Eq. 2.8). One split
also depends on which subjects are in the test set. A different split can give
a different estimate.

## Computing leave-one-out predictions {#computing-loo}

The scheme sets the predictive distribution $p(y_i \mid y_{-i})$. The
computation of that distribution is a separate choice
(Vehtari et al., 2026, Sec. 1.3). Exact LOO fits the model once for each
observation, here 144 times. The approximations in the table need fewer fits.

::: {.callout .callout-warning}
**Moment matching and reloo give correct `elpd` only**

Both methods update `elpd_loo` for the observations with a high Pareto
$\hat k$. The other measures do not get this update:

+ `loo_moment_match()` computes new weights for *transformed* draws. It stores
  the weights in the `psis` object, but not the transformed draws. `ypred` and
  `mupred` still hold the original draws, so they do not match the weights.
+ `reloo` refits the model and replaces `elpd_loo`. It keeps the old PSIS
  weights and sets the Pareto $\hat k$ of these observations to 0. The other
  measures then use unreliable weights, and the diagnostics do not show it.

For `elpd`, `mlpd`, and `ic`, pass the updated `loo` object to
`loo_pred_measure()`. For the other measures, use exact LOO, mixture importance
sampling, or K-fold.
:::

| Method | Fits | Use when | Call | Entry point |
| :--- | :--- | :--- | :--- | :--- |
| exact LOO | $n$ | the model is fast, or $n$ is small | `brms::kfold(fit, folds = "loo", save_fits = TRUE)` | `kfold_pred_measure()`, all measures |
| PSIS-LOO | 1 | all Pareto $\hat k$ are below the threshold | `loo(fit, save_psis = TRUE)` | `loo_pred_measure()`, all measures |
| PSIS-LOO with moment matching | 1 | some $\hat k$ are high; the fit keeps all draws (`save_pars(all = TRUE)` in **brms**) | `loo(fit, moment_match = TRUE, save_psis = TRUE)` | `loo_pred_measure()`, log score only |
| PSIS-LOO with reloo | 1, plus one for each high $\hat k$ | few $\hat k$ are high | `loo(fit, reloo = TRUE, save_psis = TRUE)` | `loo_pred_measure()`, log score only |
| mixture importance sampling | 1, of a mixture posterior | many $\hat k$ are high | a fit with one extra term, weights by hand | `loo_pred_measure()` with `ylp` and `psis_object`, all measures |
| subsampling | 1 | $n$ is large | `loo_subsample(fit, observations = m)` | not supported |
| K-fold | $K$ | many $\hat k$ are high, or LOO is too slow | `brms::kfold(fit, K = 10, save_fits = TRUE)` | `kfold_pred_measure()`, all measures |

The **loo** articles describe the methods in detail:
[moment matching](../loo2-moment-matching.html),
[mixture importance sampling](../loo2-mixis.html), and
[subsampling for large data](../loo2-large-data.html).

The subsections below run each method on `fit_sleep`. The
[comparison](#compare-schemes) at the end of the article plots all results.
PSIS-LOO and K-fold are in the sections above.

### Exact LOO {#exact-loo}

`brms::kfold()` with `folds = "loo"` fits the model 144 times, once without
each observation. The result is a `kfold` object. Thus its rows have the suffix
`_kfold`, although the scheme is leave-one-out. Exact LOO is the reference for
all approximations in this section.

```{r exact-loo, results="hide"}
exact_loo <- brms::kfold(fit_sleep, folds = "loo", save_fits = TRUE)
ypred_exact <- brms::kfold_predict(exact_loo, method = "predict")$yrep
mupred_exact <- brms::kfold_predict(exact_loo, method = "fitted")$yrep
```

```{r exact-loo-measures}
exact_measure <- kfold_pred_measure(
  y = y_sleep,
  ypred = ypred_exact,
  mupred = mupred_exact,
  kfold = exact_loo,
  measure = measures
)
exact_measure
```

### Moment matching {#moment-matching}

Moment matching transforms the draws for each observation with a high Pareto
$\hat k$, until the importance weights are reliable
(Paananen et al., 2021). **brms** needs all draws of the fit, so refit the
model with `save_pars(all = TRUE)`.

```{r moment-matching, results="hide"}
fit_sleep_mm <- update(fit_sleep, save_pars = brms::save_pars(all = TRUE),
                       seed = SEED, refresh = 0)
loo_mm <- brms::loo(fit_sleep_mm, moment_match = TRUE, save_psis = TRUE)
```

Request only the log score, see the warning above:

```{r moment-matching-measures}
mm_measure <- loo_pred_measure(loo = loo_mm, measure = c("elpd", "mlpd"))
mm_measure
```

All Pareto $\hat k$ of this model are below the threshold. Thus moment matching
transforms no draws, and the result equals PSIS-LOO of `fit_sleep_mm`.

### Reloo {#reloo}

`reloo` refits the model once for each observation with a high Pareto $\hat k$.

```{r reloo, results="hide"}
loo_reloo <- brms::loo(fit_sleep, reloo = TRUE, save_psis = TRUE)
```

```{r reloo-measures}
reloo_measure <- loo_pred_measure(loo = loo_reloo, measure = c("elpd", "mlpd"))
reloo_measure
```

No Pareto $\hat k$ is above the threshold, so `reloo` runs no refits. The
result equals PSIS-LOO.

### Mixture importance sampling {#mixture-is}

PSIS-LOO takes its draws from the full-data posterior. When many observations
have a high Pareto $\hat k$, these draws cover the leave-one-out posteriors
badly. Mixture importance sampling takes its draws from a mixture of all
leave-one-out posteriors instead (Silva and Zanella, 2022):

$$
q_{\text{mix}}(\theta) \propto p(\theta \mid y)
  \sum_{i=1}^n p(y_i \mid \theta)^{-1} .
$$

**Fit.** The mixture adds one term to the log density of the model. Add it to
the Stan code of **brms** with a `stanvar`. `Y`, `X`, `b`, and `sigma` are the
names in the code that **brms** writes for this model; check them with
`brms::stancode()` for other models.

```{r mixis-fit, results="hide"}
mixis_term <- brms::stanvar(
  scode = "
    {
      vector[N] log_lik_mix;
      for (n in 1:N) {
        log_lik_mix[n] = normal_lpdf(Y[n] | X[n] * b, sigma);
      }
      target += log_sum_exp(-log_lik_mix);
    }",
  block = "likelihood"
)

fit_sleep_mix <- brms::brm(
  y ~ 0 + Intercept + Days,
  data = sleepstudy,
  family = gaussian(),
  prior = prior_sleep,
  stanvars = mixis_term,
  chains = 2, iter = 400, seed = SEED, refresh = 0
)
```

**Weights.** For observation $i$, the importance weight of the mixture draw
$\theta^s$ is

$$
\log w_i^s = -\ell_i^s - \log \sum_{j=1}^n \exp(-\ell_j^s) + \text{const} .
$$

`loo_pred_measure()` normalizes the weights. Pass them in a list as
`psis_object`, together with the log densities and the predictive draws of the
mixture fit:

```{r mixis-measures}
ylp_mix <- brms::log_lik(fit_sleep_mix)
log_weights_mix <- -ylp_mix - matrixStats::rowLogSumExps(-ylp_mix)

mixis_measure <- loo_pred_measure(
  ylp = ylp_mix,
  psis_object = list(log_weights = log_weights_mix),
  y = y_sleep,
  ypred = brms::posterior_predict(fit_sleep_mix),
  mupred = brms::posterior_epred(fit_sleep_mix),
  measure = c("mlpd", "rps", "rmse", "r2")
)
mixis_measure
```

The draws come from the mixture, not from the posterior. Thus the result has
no Pareto $\hat k$. It has no `elpd` row either: `p_loo` compares the LOO log
score with the in-sample log score of the posterior draws, and those draws are
missing.

### Subsampling {#subsampling}

For large $n$, even PSIS-LOO is slow. `loo_subsample()` computes the LOO log
score for a random subsample of $m$ observations. It estimates `elpd_loo` for
all $n$ observations from this subsample.

```{r subsampling}
set.seed(SEED)
loo_ss <- brms::loo_subsample(fit_sleep, observations = 50)
loo_ss
```

`loo_pred_measure()` does not accept this result. Its PSIS object holds weights
for the 50 subsampled observations only. Thus subsampling gives the log score
only, with an additional subsampling standard error.

## Add measures to an existing result {#add-measures}

`pred_measure()` adds measures to a result of any entry point. It takes the
scheme from the result. It does not recompute the rows that are already
stored. If a requested measure is already stored, `pred_measure()` warns and
skips it.

| Approach | When to use | Call |
| :--- | :--- | :--- |
| **one call** | you know all measures beforehand | `loo_pred_measure(..., measure = c("rps", "r2"))` |
| **step by step** | you explore, or you add measures later | `pred_measure(..., predperf = result, measure = "r2")` |

**Implementation.** For PSIS-LOO, set `save_psis = TRUE` in the first call.
`pred_measure()` then reuses the stored PSIS object and its diagnostics.

```{r add-measures}
loo_elpd <- loo_pred_measure(loo = loo_sleep, save_psis = TRUE)
loo_elpd

pred_measure(
  y = y_sleep,
  mupred = mupred_sleep,
  predperf = loo_elpd,
  measure = c("mae", "rmse")
)
```

The first call uses the default `measure = NULL`, which gives `elpd`. The
second call adds `mae_loo` and `rmse_loo` with the stored PSIS weights.

## Custom measures {#custom-measures}

`measure` also accepts functions. A custom function must follow three rules:

+ **Arguments.** Name the arguments from `y`, `ypred`, `mupred`, `ylp`, and
  `log_weights`. **loo** supplies each of these names that appears in the
  function.
+ **Weights.** `log_weights` holds the PSIS weights for PSIS-LOO. In the other
  schemes it is `NULL`. A function that ignores `log_weights` gives in-sample
  values under PSIS-LOO.
+ **Return value.** Return a list with `estimate`, `se`, and `pointwise`.

Pass the function in a named list. The name becomes the row name, for example
`list("rps", cover90 = coverage)`.

**Example.** The coverage of the central 90% predictive interval is the
proportion of observations inside that interval. For a calibrated model it is
close to 0.9. Observation $i$ is inside the interval when its probability
integral transform lies between 0.05 and 0.95:

$$
\text{pit}_i = \sum_{s=1}^S w_i^s\, \mathbb{1}\{\tilde y_i^s \le y_i\},
\qquad
\text{cover}_i = \mathbb{1}\{0.05 < \text{pit}_i \le 0.95\} .
$$

```{r custom-function}
coverage <- function(y, ypred, log_weights = NULL, prob = 0.9) {
  if (is.null(log_weights)) {
    log_weights <- matrix(-log(nrow(ypred)), nrow(ypred), ncol(ypred))
  }
  w <- exp(sweep(log_weights, 2, matrixStats::colLogSumExps(log_weights)))
  pit <- colSums(w * sweep(ypred, 2, y, "<="))
  alpha <- 1 - prob
  cover_i <- as.numeric(pit > alpha / 2 & pit <= 1 - alpha / 2)
  list(
    estimate = mean(cover_i),
    se = sqrt(var(cover_i) / length(cover_i)),
    pointwise = cover_i
  )
}
```

**Implementation.** The same function works in every scheme:

```{r custom-measures}
insample_pred_measure(
  y = y_sleep,
  ypred = ypred_sleep,
  measure = list("rps", cover90 = coverage)
)

loo_pred_measure(
  loo = loo_sleep,
  y = y_sleep,
  ypred = ypred_sleep,
  measure = list("rps", cover90 = coverage)
)
```

## Compare the schemes and methods {#compare-schemes}

The results use the same model and the same measures. They differ in the
scheme, which sets the predictive question, and in the method, which computes
the predictions.

Compare the results on the mean scale. `elpd` is a sum over $\mathcal{E}$
(Vehtari et al., 2026, Eq. 2.7). The test set has 40 observations and the other
schemes have 144, so their `elpd` values are not comparable. `mlpd` removes
this dependence. The subsampling result has only `elpd_loo`. Divide its
estimate and its standard error by $n$ to get `mlpd`.

```{r compare-table}
results <- list(
  "in-sample" = insample_sleep,
  "exact LOO" = exact_measure,
  "PSIS-LOO" = loo_sleep_measure,
  "PSIS-LOO, moment matching" = mm_measure,
  "PSIS-LOO, reloo" = reloo_measure,
  "mixture IS" = mixis_measure,
  "K-fold" = kfold_sleep_measure,
  "test data" = test_sleep_measure
)

df_methods <- do.call(rbind, lapply(names(results), function(method) {
  est <- results[[method]]$estimates
  measure <- sub("_(loo|kfold|test)$", "", rownames(est))
  keep <- measure %in% c("mlpd", "rps", "rmse", "r2")
  data.frame(
    method = method,
    measure = measure[keep],
    estimate = est[keep, "Estimate"],
    se = est[keep, "SE"],
    row.names = NULL
  )
}))

n_sleep <- length(y_sleep)
df_methods <- rbind(df_methods, data.frame(
  method = "subsampling",
  measure = "mlpd",
  estimate = loo_ss$estimates["elpd_loo", "Estimate"] / n_sleep,
  se = loo_ss$estimates["elpd_loo", "SE"] / n_sleep
))

method_levels <- c("in-sample", "exact LOO", "PSIS-LOO",
                   "PSIS-LOO, moment matching", "PSIS-LOO, reloo",
                   "mixture IS", "subsampling", "K-fold", "test data")
df_methods$method <- factor(df_methods$method, levels = rev(method_levels))
df_methods$scheme <- factor(
  ifelse(df_methods$method %in% c("in-sample", "K-fold", "test data"),
         as.character(df_methods$method), "leave-one-out"),
  levels = c("in-sample", "leave-one-out", "K-fold", "test data")
)
df_methods[order(df_methods$measure, df_methods$method, decreasing = TRUE), ]
```

**Plot.** Each panel shows one measure. Each row is one method, and the color
shows its scheme. The point is the estimate, and the bar is $\pm 1$ standard
error. `mlpd` and `r2` are utilities: higher is better. `rps` and `rmse` are
losses: lower is better. Moment matching, reloo, and subsampling have a point
in the `mlpd` panel only.

```{r compare-plot, fig.height=6}
ggplot(df_methods, aes(estimate, method, color = scheme)) +
  geom_pointrange(aes(xmin = estimate - se, xmax = estimate + se)) +
  facet_wrap(~measure, scales = "free_x") +
  scale_color_manual(values = c("in-sample" = "grey50",
                                "leave-one-out" = "#0072B2",
                                "K-fold" = "#009E73",
                                "test data" = "#D55E00")) +
  labs(x = "estimate ± SE", y = NULL, color = "scheme") +
  theme(legend.position = "bottom")
```

**Interpretation.** In-sample, leave-one-out, and K-fold score the same 144
observations. The in-sample estimates are the most optimistic for every
measure, because the model has seen each $y_i$. K-fold answers almost the same
question as leave-one-out, so its estimates are close.

All leave-one-out methods agree with exact LOO. The largest difference is in
`rps`: PSIS-LOO gives 28.47 and exact LOO 28.76. This difference is much
smaller than the standard error of 1.8. Mixture importance sampling gives 28.76.
All Pareto $\hat k$ are good, so moment matching and reloo change nothing, and
their `mlpd` equals that of PSIS-LOO. For a model with many high $\hat k$, the
methods differ more. Then exact LOO, mixture importance sampling, or K-fold
give the reliable values for all measures.

The test data score better than all other schemes here. One split gives a noisy
estimate: the five test subjects are easier to predict than the average
subject. The standard errors show this. The `rmse` of the test data has a
standard error more than twice as large as in the other schemes, because the
test set has 40 observations, not 144.

The standard errors describe the uncertainty from the finite
evaluation set. They do not include the Monte Carlo error from the finite
number of draws (Vehtari et al., 2026, Sec. 2.2).

## References

Belenky, G., Wesensten, N. J., Thorne, D. R., et al. (2003). Patterns of
performance degradation and restoration during sleep restriction and subsequent
recovery: A sleep dose-response study. *Journal of Sleep Research*,
12(1):1–12.

Paananen, T., Piironen, J., Bürkner, P.-C., and Vehtari, A. (2021). Implicitly
adaptive importance sampling. *Statistics and Computing*, 31(2):16.

Silva, L. and Zanella, G. (2022). Robust leave-one-out cross-validation for
high-dimensional Bayesian models. Preprint at
[arXiv:2209.09190](https://arxiv.org/abs/2209.09190).

Vehtari, A., Axen, S., Weber, F., and Bockting, F. (2026). Predictive measures
and model comparison in the loo package: A formula reference. [TODO: arXiv ID]

Vehtari, A., Gelman, A., and Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics and
Computing*, 27(5):1413–1432. doi:10.1007/s11222-016-9696-4

Vehtari, A., Simpson, D., Gelman, A., Yao, Y., and Gabry, J. (2024). Pareto
smoothed importance sampling. *Journal of Machine Learning Research*,
25(72):1–58.

## See also

- [Overview of predictive measures](overview-measures.html) — definitions,
  formulas, and standard errors of the measures.
- [Model comparison](model-comparison.html) — differences between models and
  their standard errors.
- [Cross-validation FAQ](https://users.aalto.fi/~ave/CV-FAQ.html) — when to use
  which cross-validation scheme.

## Appendix: model code {#appendix-model-code}

```{r fit-models-code, ref.label="fit-models", eval=FALSE}
```
